---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file.          -->
<!-- To generate README.md, run knitr::knit('README.Rmd') in this directory. -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

## PhotoGEA <a href="https://eloch216.github.io/PhotoGEA/"><img src="man/figures/logo.svg" align="left" style="height:138px"></a>

**PhotoGEA** (short for **photo**synthetic **g**as **e**xchange **a**nalysis) is
an R package that provides a suite of tools for loading, processing, and
analyzing photosynthetic gas exchange data.

This package is a work in progress and is not yet fully documented. As
documentation proceeds, functions are subject to modification or removal without
notice.

### Installing the R Package

The easiest way to install `PhotoGEA` is to type the following from within the R
terminal:

```{r, eval = FALSE}
remotes::install_github('eloch216/PhotoGEA')
```

Note that this method requires the `remotes` package, which can be installed
from within R by typing `install.packages('remotes')`.

### Learning to Use PhotoGEA

The best way to learn about using `PhotoGEA` is to visit the
[PhotoGEA website](https://eloch216.github.io/PhotoGEA/index.html) and click the
[Get started](https://eloch216.github.io/PhotoGEA/articles/PhotoGEA.html) link
in the top menu bar. The website includes documentation for all the functions
and data sets included in the package, as well as articles that describe its
general features and several important use cases.

### Example Scripts

Several example scripts are provided in the `example_scripts` directory of the
[source code repository](https://github.com/eloch216/PhotoGEA). To run one of
these scripts, set the working directory to a folder that contains a local copy
of the script and use the `source` command to execute the code in the script. No
guarantees are made that these scripts will run on your machine or be
compatible with your data, but they may be a useful source of ideas.

### License

The `PhotoGEA` R package, its documentation, and its associated example scripts
are licensed under the MIT license.
