---
title: "Frequently Asked Questions"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  fig.align = "center"
)
```

# How Do I Install The Latest Release of PhotoGEA?

`PhotoGEA` is available on CRAN, so the easiest way to install the latest
release is to type the following from within an R terminal:

```{r, eval = FALSE}
install.packages('PhotoGEA')
```

There may be a short period of time where a new version has been released but is
not yet available on CRAN. In this case, the latest release can be installed
directly from the main branch of the GitHub repository by typing the following:

```{r, eval = FALSE}
remotes::install_github('eloch216/PhotoGEA')
```

Note that this method requires the `remotes` package, which can be installed
from within R by typing `install.packages('remotes')`.

# How Do I Install An Old Version of PhotoGEA?

All `PhotoGEA` releases are tagged on GitHub, and the tag names are formatted as
`vX.Y.Z`, where `X.Y.Z` is the version number. Because of this,
`remotes::install_github` can be used to install old versions from within R,
supplying the appropriate tag name as the `ref` input argument. For example,
version `1.0.0` can be installed as follows:

```{r, eval = FALSE}
remotes::install_github('eloch216/PhotoGEA', ref = 'v1.0.0')
```

Note that this command requires the `remotes` package, which can be installed
from within R by typing `install.packages('remotes')`.

# How Do I Install The Development Version of PhotoGEA?

The development version of `PhotoGEA` contains the latest changes, although it
may be a "work in progress" and subject to suddent changes without any warning.
It can always be found on the `unreleased` branch on GitHub, and the package can
be installed from this branch as follows:

```{r, eval = FALSE}
remotes::install_github('eloch216/PhotoGEA', ref = 'unreleased')
```

Note that this command requires the `remotes` package, which can be installed
from within R by typing `install.packages('remotes')`.

# What Should I Do When check_response_curve_data Fails?

When beginning to work with a new data set, it is fairly common to encounter
problems that cause `check_response_curve_data` to fail. The most common issues
are:

- The manual log button was accidentally pressed while measuring, causing a
  response curve to have an extra point.

- One or more "User Constant" was not updated before starting a to measure a new
  curve, causing two curves to have the same identifying metadata.

- A curve was started, but needed to be ended early, resulting in a curve with
  fewer points than expected.

These are exactly the kinds of issues that `check_response_curve_data` is
designed to detect. It may feel irritating to encounter these errors, but it's
better to know about issues with extra log points or user constants early in
your analysis, before they cause additional downstream issues.

In each of these cases, it is often simplest to fix the issues by manually
editing the log files in Excel. This is an easy way to delete extra rows or
alter the values of user constants. When altering log files, it is always a good
idea to keep an "original" version in case any of the changes need to be
reverted.

Sometimes `check_response_curve_data` fails for a different reason -- for
example, perhaps different setpoint sequences were intentionally used while
measuring the curves, or perhaps you have already cleaned up your data in Excel.
In these cases, please see the "What If My Response Curves Have Different
Numbers of Points?" and "What If I Clean My Data In Excel?" sections of this
article.

# What If My Response Curves Have Different Numbers of Points?

The analysis guides use data sets where each response curve has the same number
of points, making it easy to apply `check_response_curve_data` and
`organize_response_curve_data`. However, data sets may sometimes have curves
measured using different sequences of setpoints.

To deal with this situation, one strategy is to split the full set into groups
that are expected to use the same sequences, then separately check and organize
each group, and finally recombine all the groups back together. The code
snippets below show two examples of how this could be accomplished.

## Splitting response curves according to the number of points in each curve

The following code would replace the calls to `check_response_curve_data` and
`organize_response_curve_data` in the user guides. This code was originally
written for a set of A-Ci curves that used different numbers of recovery points.
The user wished to keep the final recovery point to use for subsequent analysis.

The curves used the following sequences of `CO2_r` setpoint values:

- 16 points: 400, 300, 200, 120, 70, 30, 10, 400, 400, 400, 600, 800, 1200,
  1500, 1800, 400

- 18 points: 400, 300, 200, 120, 70, 30, 10, 400, 400, 400, 400, 400, 600, 800,
  1200, 1500, 1800, 400

- 19 points: 400, 300, 200, 120, 70, 30, 10, 400, 400, 400, 400, 400, 400, 600,
  800, 1200, 1500, 1800, 400

```{r, eval = FALSE}
# Add a new column called `curve_npts` that stores the number of points in each
# response curve
licor_data <- do.call(rbind, by(licor_data, licor_data[, 'curve_identifier'], function(x) {
  x[, 'curve_npts'] <- nrow(x)
  x
}))

# Choose points to remove, depending on how many points are in the curve
pts_to_remove <- list(
  '16' = c(1, 8:9,  16),
  '18' = c(1, 8:11, 18),
  '19' = c(1, 8:12, 19)
)

# Check and process each group of curves depending on how many points are in the
# curve
licor_exdf_list_processed <- by(licor_data, licor_data[, 'curve_npts'], function(x) {
  # Get the number of points in these curves
  npts <- x[1, 'curve_npts']

  # Make sure info is specified for this group of curves
  if (!as.character(npts) %in% names(pts_to_remove)) {
    stop('Points to remove were not specified for npts = `', npts, '`')
  }

  # Make sure the data meets basic requirements
  check_response_curve_data(x, 'curve_identifier', npts, 'CO2_r_sp')

  # Remove points with duplicated `CO2_r_sp` values and order by `Ci`
  organize_response_curve_data(
    x,
    'curve_identifier',
    pts_to_remove[[as.character(npts)]],
    'Ci'
  )
})

# Use `rbind` to recombine all the data
licor_data <- do.call(rbind, licor_exdf_list_processed)
```

## Splitting the response curves according to the date they were measured and the number of points in each curve

This code would replace the calls to `check_response_curve_data` and
`organize_response_curve_data` in the user guides. This code was originally
written for a set of A-Ci curves that used different sequences of `CO2_r`
setpoints on different days.

The curves used the following sequences of `CO2_r` setpoint values:

- 2023-03-21: 18 points: 400, 300, 200, 150, 100, 75, 50, 40, 30, 20, 10, 400,
  400, 600, 800, 1000, 1200, 1500

- 2023-03-23: 19 points: 400, 300, 200, 150, 100, 75, 50, 40, 30, 20, 10, 400,
  400, 500, 600, 800, 1000, 1200, 1500

- 2023-03-24: 19 points: 400, 300, 200, 150, 100, 75, 50, 40, 30, 20, 400, 400,
  450, 500, 600, 800, 1000, 1200, 1500

```{r, eval = FALSE}
# Add a new column called `curve_npts` that stores the number of points in each
# response curve
licor_data <- do.call(rbind, by(licor_data, licor_data[, 'curve_identifier'], function(x) {
  x[, 'curve_npts'] <- nrow(x)
  x
}))

# Add a new column called `date_ymd` that stores the date formatted as
# YYYY-MM-DD
licor_data[, 'date_ymd'] <- paste(
  substring(licor_data[, 'date'], 1, 4),
  substring(licor_data[, 'date'], 5, 6),
  substring(licor_data[, 'date'], 7, 8),
  sep = '-'
)

# Add a new column called `date_ymd_npts` that combines the date and the number
# of points
licor_data[, 'date_ymd_npts'] <-
  paste(licor_data[, 'date_ymd'], licor_data[, 'curve_npts'], sep = ' - ')

# Choose points to remove, depending on the date the curve was measured and the
# number of points it contains
pts_to_remove <- list(
  '2023-03-21 - 18' = c(12, 13),
  '2023-03-23 - 19' = c(12, 13),
  '2023-03-24 - 19' = c(11, 12)
)

# Check and process each group of curves depending on the date and the number of
# points
licor_exdf_list_processed <- by(licor_data, licor_data[, 'date_ymd_npts'], function(x) {
  # Get the date and number of points in these curves
  date_ymd_npts <- x[1, 'date_ymd_npts']
  npts <- x[1, 'curve_npts']

  # Make sure info is specified for this group of curves
  if (!date_ymd_npts %in% names(pts_to_remove)) {
    stop('Points to remove were not specified for date_ymd_npts = `', date_ymd_npts, '`')
  }

  # Make sure the data meets basic requirements
  check_response_curve_data(x, 'curve_identifier', npts, 'CO2_r_sp')

  # Remove points with duplicated `CO2_r_sp` values and order by `Ci`
  organize_response_curve_data(
    x,
    'curve_identifier',
    pts_to_remove[[date_ymd_npts]],
    'Ci'
  )
})

# Use `rbind` to recombine all the data
licor_data <- do.call(rbind, licor_exdf_list_processed)
```

# What If I Clean My Data In Excel?

In the user guides, the `organize_response_curve_data` and `remove_points`
functions are used to remove recovery points and other points from sets of
response curves. However, it is also possible to remove points in Excel before
reading log files into R. In this case, it is necessary to make a few small
alterations to the code used in the user guides.

One consideration is that after cleaning the curves in Excel, it is likely that
not all curves have the same number of points or follow the same sequence of
setpoint values. Because of this, the checks in `check_response_curve_data` are
likely to fail. In this case, we recommend setting `expected_npts` to 0 (the
default value) and `error_on_failure` to `FALSE` when calling
`check_response_curve_data`. This will provide potentially useful information
about the number of points in each curve, but won't throw an error that would
cause a script to stop running.

Another consideration is that `organize_response_curve_data` is not needed to
remove any points from the curves. Yet, its other features, such as reordering
the points and calculating average values, are still useful. In this case, we
recommend setting `measurement_numbers_to_remove` to `c()` and leaving the other
arguments as-is.

Putting this all together would produce something like the following code, which
would replace the regular calls to `check_response_curve_data` and
`organize_response_curve_data` from the user guides:

```{r, eval = FALSE}
# Print info about the number of points in each curve to make sure
# `curve_identifier` is able to properly split the set into individual curves
check_response_curve_data(licor_data, 'curve_identifier', error_on_failure = FALSE)

# Reorder by `Ci` and calculate average values of leaf temperature and Qin
licor_data <- organize_response_curve_data(
    licor_data,
    'curve_identifier',
    c(),
    'Ci',
    columns_to_average = c('TleafCnd', 'Qin')
)
```

# What If I Have Log Files From a Licor LI-6400, PP Systems CIRAS, or Other Gas Exchange Measurement System?

At the moment, `PhotoGEA` can read plaintext and Excel files created by Licor
LI-6800 gas exchange measurement systems, but does not have functions for
reading log files from other instruments.

Fortunately, there are other R packages that do, such as
[GasanalyzeR](https://doi.org/10.1093/aobpla/plae035). Using a function from
another package is a little complicated, because generally they do not use the
same data structures or variable names as `PhotoGEA`. So, after reading a file
with another package, there is typically some extra "conversion" that needs to
take place.

The following example shows how to create a wrapper for the `read_6400_xls`
function from the `gasanalyzer` R package. It also provides a simple function
that "detects" the file type, so a mix of Licor LI-6400 and LI-6800 log files
could be used. The code in the example would replace the lines that use
`read_gasex_file` to create `licor_exdf_list` in the user guides.

As of the time this article was written (April 4, 2025), the `read_6400_xls`
function is not available in the CRAN version of the `gasanalyzer` package.
Instead, the package must be installed directly from its GitLab repository to
get the latest version. This can be done by calling the following command from
R:

```{r, eval = FALSE}
remotes::install_gitlab('plantphys/gasanalyzer')
```

This example only processes the columns that are absolutely essential for
fitting A-Ci curves. Depending on your goals, you may need to modify this code
to include other columns.

```{r, eval = FALSE}
# Helper function for reading 6400 Excel files. This is a wrapper for the
# read_6400_xls function from the gasanalyzer package that converts the output
# into a format usable by PhotoGEA.
#
# Debug option: If DEBUG_PRINT is set to TRUE below, the raw column names will
# be printed immediately after the file is loaded. This may be helpful if there
# are other columns (such as user constants) that also need to be renamed.
read_6400_xls_wrapper <- function(fpath) {
    # Read the contents of the file
    rawdata <- gasanalyzer::read_6400_xls(fpath)

    # Optional debug printing
    DEBUG_PRINT <- FALSE
    if (DEBUG_PRINT) {
      print(colnames(rawdata))
    }

    # Rename a few columns so they meet PhotoGEA's expectations; the "new"
    # PhotoGEA column name is on the left, and the "original" gasanalyzer
    # column name is on the right.
    new_column_names <- list(
        A = 'GasEx.A',
        Ca = 'GasEx.Ca',
        Ci = 'GasEx.Ci',
        DeltaPcham = 'Meas.DeltaPcham',
        gsw = 'GasEx.gsw',
        oxygen = 'Const.Oxygen',
        Pa = 'Meas.Pa',
        Qin = 'LeafQ.Qin',
        TleafCnd = 'GasEx.TleafCnd'
    )

    for (i in seq_along(new_column_names)) {
        gasanalyzer_name <- new_column_names[[i]]
        photogea_name    <- names(new_column_names)[i]

        colnames(rawdata)[colnames(rawdata) == gasanalyzer_name] <- photogea_name
    }

    # Drop units and convert to a regular data frame
    rawdata_df <- as.data.frame(units::drop_units(rawdata))

    # Convert to an exdf object
    exdf_obj <- exdf(rawdata_df)

    # Supply units for a few columns so they meet PhotoGEA's expectations, and
    # return
    document_variables(
        exdf_obj,
        c('read_6400_xls_wrapper', 'A',          'micromol m^(-2) s^(-1)'),
        c('read_6400_xls_wrapper', 'Ca',         'micromol mol^(-1)'),
        c('read_6400_xls_wrapper', 'Ci',         'micromol mol^(-1)'),
        c('read_6400_xls_wrapper', 'DeltaPcham', 'kPa'),
        c('read_6400_xls_wrapper', 'gsw',        'mol m^(-2) s^(-1)'),
        c('read_6400_xls_wrapper', 'oxygen',     'percent'),
        c('read_6400_xls_wrapper', 'Pa',         'kPa'),
        c('read_6400_xls_wrapper', 'Qin',        'micromol m^(-2) s^(-1)'),
        c('read_6400_xls_wrapper', 'TleafCnd',   'degrees C')
    )
}

# Helper function that tries to read a file using PhotoGEA::read_gasex_file, but
# then uses `read_6400_xls_wrapper` if there is an error.
read_gasex_file_plus_6400 <- function(fpath) {
  cat(paste0(
    '\nAttempting to read `', fpath, '` using PhotoGEA::read_gasex_file\n'
  ))

  tryCatch(
    read_gasex_file(fpath),
    error = function(e) {
      cat(paste0(
        '\nAn error occurred. Attempting to read `', fpath,
        '` using gasanalyzer::read_6400_xls\n'
      ))

      read_6400_xls_wrapper(fpath)
    }
  )
}

# Load each file, storing the result in a list
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_gasex_file_plus_6400(fpath)
})
```

# What If My Files Don't Have User Constants or Other Metadata?

Please see the
[Guide to Licor LI-6800 User Constants](LI6800_user_constants.html).
