---
title: "Analyzing Mesophyll Conductance Data"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Analyzing Mesophyll Conductance Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

In this vignette, we will give an example showing how to use combined gas
exchange and isotope discrimination data to estimate values of mesophyll
conductance with the `PhotoGEA` package. The commands in this vignette can be
used to initialize your own script, as described in [Customizing Your Script].

# Background

Because the process of photosynthesis tends to preferentially assimilate ^12^C
more often than the rarer and heavier stable isotope ^13^C, carbon isotope
discrimination measurements have proven to be incredibly useful in the fields of
plant biology and climate science [@evans_temperature_2013]. In the context of
plant biology, a common approach is to hold a leaf or plant under controlled
environmental conditions and measure isotope discrimination; models that predict
discrimination can then be used to estimate key parameter values such as the
mesophyll conductance to CO~2~ ($g_{mc}$) in C~3~ plants or the bundle-sheath
leakiness to CO~2~ ($\phi$) in C~4~ plants.

In general, this technique requires simultaneous measurements of the net CO~2~
assimilation rate $A_n$, the intercellular CO~2~ concentration $C_i$, the ^12^C
and ^13^C concentrations in air before and after flowing over the leaf, and
others. To make these measurements, a gas exchange measurement system (such as
a Licor Li-6800) is combined with an isotope concentration measurement system,
making this a challenging experimental technique. Although several methods are
available for measuring the relative concentrations of ^12^C and ^13^C gas
mixtures, tunable diode laser (TDL) absorption spectroscopy has emerged as one
of the most popular due to its high sensitivity and relatively short measurement
times [@salesse-smith_modifying_2022].

## Isotope Discrimination Measurements

To understand isotope discrimination, it is helpful to first discuss
_fractionation_. Fractionation is a process where the relative abundance of one
isotope differs between the product and the reactant of a chemical reaction. The
term _fractionation_ also refers to a measurable quantity $\alpha$ defined by
$$\alpha = \frac{R_r}{R_p},$$ where $R_r$ and $R_p$ are the ratios of isotope
concentrations in the reactant and product, respectively. In the case of
photosynthetic carbon isotope discrimination, $\alpha^{13}C$ is the
fractionation of ^13^C relative to ^12^C, air is the reactant, plant carbon is
the product, and the ratios are given by ^13^C / ^12^C. Because photosynthesis
assimilates more ^12^C than ^13^C, $R_p$ is always smaller than $R_r$, and
$\alpha^{13}C$ is always greater than 1.

The isotope discrimination $\Delta$ is related to the isotope fractionation
$\alpha$ according to $$\Delta = \alpha - 1 = \frac{R_r - R_p}{R_p}.$$ So, in
the case of photosynthetic carbon isotope discrimination, $\Delta^{13}C$ is
always positive (because $\alpha$ is always greater than 1). The discrimination
can also be expressed in terms of the isotopic composition $\delta$, which is
related to the isotope ratio $R$ by $\delta = R - 1$:
$$\Delta^{13}C = \frac{\delta^{13}C_r - \delta^{13}C_p}{1 + \delta^{13}C_p}.$$
The isotopic compositions can be expressed relative to an international
standard, making them the preferred way to express the relative abundance of
different isotopes.

During photosynthetic carbon isotope discrimination measurements, it is not
typically possible to actually measure $\delta^{13}C_p$ for a variety of
reasons. Instead, $\delta^{13}C$ can be measured in the outgoing air from the
leaf. Then, it is possible to relate the isotopic compositions of the incoming
and outgoing air to and from the leaf ($\delta^{13}C_{in}$ and
$\delta^{13}C_{out}$, respectively) to $\delta^{13}C_p$. In this case, the full
equation for $\Delta^{13}C$ becomes $$\Delta^{13}C =
\frac{\xi (\delta^{13}C_{out} - \delta^{13}C_{in})}
{1 + \delta^{13}C_{out} - \xi (\delta^{13}C_{out} - \delta^{13}C_{in})}.$$ Here,
$\xi$ is defined by $$\xi = \frac{C_{in}}{C_{in} - C_{out}},$$ where $C_{in}$
and $C_{out}$ are the ^12^C concentrations in the incoming and outgoing air.

In summary, the photosynthetic carbon isotope discrimination $\Delta^{13}C$ is
an experimentally accessible quantity that reflects the degree of preference for
^12^C during the process of photosynthesis. For more information about this
discrimination and its measurement via TDL, see @ubierna_using_2018.

## Isotope Discrimination Calculations

Values of $\Delta^{13}C$ can be predicted from knowledge of the photosynthetic
chemical pathway, as first done by @farquhar_relationship_1982. This approach
breaks down photosynthetic carbon assimilation into several steps including
diffusion through air and liquid, and the binding of CO~2~ to RuBP as catalyzed
by Rubisco. Each step along the pathway has its own distinct fractionation. It
is also necessary to include the release of carbon by photorespiration and day
respiration, which also have their own associated fractionations. The entire
equation relating $\Delta^{13}C$ to $A_n$, $C_i$, and other variables is
different for C~3~ and C~4~ plants. These equations can also each be written in
several different ways depending on assumptions made while deriving them.
Due to these complexities, we do not reproduce the equations here. However, they
can be found in many places, such as Box A in @ubierna_using_2018.

The most important aspect of these equations is that they can be solved for
key parameters such as the mesophyll conductance to CO~2~ (in C~3~ plants) or
the bundle-sheath leakiness to CO~2~ (in C~4~ plants). Then,
experimentally-measured values of $\Delta^{13}C$ can be used to calculate values
of $g_{mc}$ or $\phi$. There are many subtelties to these calculations, such as
uncertainties in the fractionation values due to photorespiration and day
respiration. We will not attempt to discuss them all here, but more information
can be found in @busch_revisiting_2020.

## A Note About This Example

Estimating mesophyll conductance from combined gas exchange and TDL measurements
is a complicated process, making this example longer than most of the other
`PhotoGEA` vignettes. In general, the analysis follows these steps:

* Separately load the gas exchange and TDL data.
* Pair the data together, making sure that each log point in the gas exchange
  data can be identified with carbon isotope concentrations in the sample and
  reference lines as measured by the TDL.
* Calculate photosynthetic carbon isotope discrimination and mesophyll
  conductance.

For brevity, we will not discuss the types of log files used in this example,
but more information can be found in the
[Analyzing TDL Data](analyzing_tdl_data.html) and
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) vignettes.

# Loading Packages

As always, the first step is to load the packages we will be using. In addition
to `PhotoGEA`, we will also use the `lattice` package for generating plots.

```{r setup}
# Load required packages
library(PhotoGEA)
library(lattice)
```

If the `lattice` package is not installed on your R setup, you can install it by
typing `install.packages('lattice')`.

# Loading Gas Exchange Data

The `PhotoGEA` package includes two files representing gas exchange data that
was measured along with the TDL data discussed in the next section. The data was
recorded by Licor Li-6800 instruments and is stored in Microsoft Excel files.
Each file includes two different `CO2_r` setpoints (715 ppm and 450 ppm); six
logs were recorded at each setpoint. Although these two files are based on real
data, noise was added since they are unpublished, so these files should only be
used as examples.

The files will be stored on your computer somewhere in your R package
installation directory, and full paths to these files can be obtained with
`PhotoGEA_example_file_path`:

```{r licor_file_names}
# Define a vector of paths to the files we wish to load; in this case, we are
# loading example files included with the PhotoGEA package
licor_file_paths <- c(
  PhotoGEA_example_file_path('licor_for_gm_site11.xlsx'),
  PhotoGEA_example_file_path('licor_for_gm_site13.xlsx')
)
```

(**Note:** When loading your own files for analysis, it is not advisable to use
`PhotoGEA_example_file_path` as we have done here. Instead, file paths can be
directly written, or files can be chosen using an interactive window. See
[Input Files] below for more information.)

To actually read the data in the files and store them in R objects, we will use
the `read_gasex_file` function from `PhotoGEA`. Later, we will need to match
the timestamps in the gas exchange and isotope discrimination data; to make sure
the timestamps are interpreted correctly, we will also specify the time zone
where the measurements were made (`US/Central`). Since there are multiple files
to read, we will call this function once for each file using `lapply`:

```{r loading_licor_data}
# Load each Licor file, storing the result in a list
licor_exdf_list <- lapply(licor_file_paths, function(fpath) {
  read_gasex_file(fpath, 'time', posix_options = list(tz = 'US/Central'))
})
```

The result from this command is an R list of "extended data frames" (abbreviated
as `exdf` objects). The `exdf` class is a special data structure defined by the
`PhotoGEA` package. In many ways, an `exdf` object is equivalent to a data
frame, with the major difference being that an `exdf` object includes the units
of each column. For more information, type `?exdf` in the R terminal to access
the built-in help menu entry, or check out the
[Working With Extended Data Frames](working_with_extended_data_frames.html)
vignette.

Later, it will also be necessary to know the "site number" for each gas exchange
system. This refers to the TDL valve numbers corresponding to the sample and
reference lines of each gas exchange instrument. At UIUC, there is a convention
where the sample line valve number is stored in the Licor log file name. It can
be extracted using the `get_sample_valve_from_filename` function from the
`PhotoGEA` package, which also provides a method for specifying the reference
line valve number.

```{r site_number}
# Get TDL sample and reference valve numbers from the Licor filenames
licor_exdf_list <- lapply(licor_exdf_list, function(exdf_obj) {
  get_sample_valve_from_filename(exdf_obj, list(
    '13' = 12, # The reference valve is 12 when the sample valve is 13
    '11' = 10  # The reference valve is 10 when the sample valve is 11
  ))
})
```

Generally, it is more convenient to work with a single `exdf` object rather than
a list of them, so our next step will be to combine the objects in the list.
This action can be accomplished using the `rbind` function, which combines
table-like objects by their rows; in other words, it stacks two or more tables
vertically. This action only makes sense if the tables have the same columns, so
before we combine the `exdf` objects, we should make sure this is the case.

The `PhotoGEA` package includes a function called `identify_common_columns` that
can be used to get the names of all columns that are present in all of the Licor
files. Then, we can extract just those columns, and then combine the `exdf`
objects into a single one.

```{r combining_licor_data}
# Get the names of all columns that are present in all of the Licor files
licor_columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , licor_columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
licor_data <- do.call(rbind, licor_exdf_list)
```

Now we have a single R object called `licor_data` that includes all the data
from several Licor Excel files. For more information about consolidating
information from multiple files, see the _Common Patterns_ section of the
[Working With Extended Data Frames](working_with_extended_data_frames.html#combining-data-from-several-files)
vignette.

# Loading Tunable Diode Laser Data

The `PhotoGEA` package includes one file representing TDL data that was
recorded with a Campbell Scientific CR3000 data logger simultaneously with the
gas exchange data discussed in the previous section. Although this two file is
based on real data, noise was added to it since it is unpublished, so it
should only be used as an example.

The file will be stored on your computer somewhere in your R package
installation directory, and a full path to the file can be obtained with
`PhotoGEA_example_file_path`:

```{r tdl_file_names}
# Define a vector of paths to the files we wish to load; in this case, we are
# loading example files included with the PhotoGEA package
tdl_file_paths <- c(
  PhotoGEA_example_file_path('tdl_for_gm.dat')
)
```

(**Note:** When loading your own files for analysis, it is not advisable to use
`PhotoGEA_example_file_path` as we have done here. Instead, file paths can be
directly written, or files can be chosen using an interactive window. See
[Input Files] below for more information.)

To actually read the data in the files and store them in R objects, we will use
the `read_gasex_file` function from `PhotoGEA`. Later, we will need to match
the timestamps in the gas exchange and isotope discrimination data; to make sure
the timestamps are interpreted correctly, we will also specify the time zone
where the measurements were made (`US/Central`). Since there are (potentially)
multiple files to read, we will call this function once for each file using
`lapply`:

```{r loading_tdl_data}
# Load each TDL file, storing the result in a list
tdl_exdf_list <- lapply(tdl_file_paths, function(fpath) {
  read_gasex_file(fpath, 'TIMESTAMP', posix_options = list(tz = 'US/Central'))
})
```

As before, the result from this command is a list of `exdf` objects. (In this
case, it is a list with one element. However, we have written this example in
a general way so it will be able to accomodate multple TDL log files if it is
necessary to use more than one.) Generally, it is more convenient to work with a
single `exdf` object rather than a list of them, so our next step will be to
combine the objects in the list. This process is the same as was used in the
previous section for the Licor log files.

```{r combining_tdl_data}
# Get the names of all columns that are present in all of the TDL files
tdl_columns_to_keep <- do.call(identify_common_columns, tdl_exdf_list)

# Extract just these columns
tdl_exdf_list <- lapply(tdl_exdf_list, function(x) {
  x[ , tdl_columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
tdl_data <- do.call(rbind, tdl_exdf_list)
```

Now we have a single R object called `tdl_data` that includes all the data
from several TDL data files. For more information about consolidating
information from multiple files, see the _Common Patterns_ section of the
[Working With Extended Data Frames](working_with_extended_data_frames.html#combining-data-from-several-files)
vignette.

# Basic Checks of the Gas Exchange Data

Next, we should make sure there is a column in the data whose value uniquely
identifies each related set of measurements. In this particular data set,
several "user constants" were defined while making the measurements that help to
identify each set: `machine` and `replicate`. However, neither of these
columns alone are sufficient to uniquely identify each set. We can solve this
issue by creating a new column that combines the values from each of these:

```{r adding_id_column}
# Create a new identifier column formatted like `machine - replicate`
licor_data[ , 'curve_identifier'] <-
  paste(licor_data[ , 'machine'], licor_data[ , 'replicate'], sep = ' - ')
```

When measuring response curves (or any other type of gas exchange data), it is
strongly recommended to record key pieces of metadata in the log files using
"user constants," as was done in this example. For more information, please see
the [Guide to Licor LI-6800 User Constants](LI6800_user_constants.html), which
also includes advice for dealing with data sets that do not contain such
metadata.

The next step is to make sure that this column correctly identifies each
response curve. To do this, we can use the `check_response_curve_data` function
from `PhotoGEA`. Here we will supply the name of a column that should uniquely
identify each response curve (`curve_identifier`), the expected number of points
in each curve (`12`), the name of a "driving" column that should follow the same
sequence in each curve (`CO2_r_sp`). If the data passes the checks, this
function will have no output and will not produce any messages. (For more
information, see the built-in help menu entry by typing
`?check_response_curve_data`.)

```{r check_response_curve}
# Make sure the data meets basic requirements
check_response_curve_data(licor_data, 'curve_identifier', 12, 'CO2_r_sp')
```

However, if `check_response_curve_data` detects an issue, it will print a
helpful message to the R terminal. For example, if we had specified the wrong
number of points or the wrong identifier column, we would get error messages:

```{r check_response_curve_error, error = TRUE}
check_response_curve_data(licor_data, 'curve_identifier', 13)

check_response_curve_data(licor_data, 'replicate', 12)
```

Now that we have confirmed that our curve identifier properly identifies sets
of data, we can also perform additional organization using the
`organize_response_curve_data` function from `PhotoGEA`:

```{r organize_data}
# Order by `Ci`.
licor_data <- organize_response_curve_data(
  licor_data,
  'curve_identifier',
  c(), # keep all the measured points
  'Ci'
)
```

Now there is a new column called `seq_num`, where each point in each set is
given a number 1-12 corresponding to the measurement order. The points have also
been re-ordered according to `Ci`.

# Combining the Gas Exchange and Isotope Discrimination Data

## Identifying TDL Cycles

Typically a TDL system periodically cycles between multiple gas lines during
measurements. Some of the gas lines represent gas mixtures with known
composition that can be used for calibration, while others are the "unknown"
mixtures whose composition is being measured. A collection of valves are used
to control which gas line is being measured at any given time, and the
"active" valve for each recorded data point is included in a measurement file.

The first step towards applying calibrations and combining the data with the
associated gas exchange measurements is to identify complete measurement cycles
within the data set. This can be accomplished using the `identify_tdl_cycles`
function from the `PhotoGEA` package. To use this function, we need to know the
following information:

* The name of the column in the TDL data that indicates which valve is being
  measured.
* The valve number that marks the beginning of a new cycle.
* How many valves are measured in each cycle.
* The name of the column in the TDL data that indicates the time at which each
  measurement was made.
* The amount of time it should take to cycle through the valves.

This information is usually known beforehand from the settings that were
specified when operating the TDL; alternatively, if it isn't known already, it
can often be determined by taking a look at the data via `View(tdl_files)`. Once
this information is obtained, `identify_tdl_cycles` can be used to automatically
assign a number to each measurement cycle:

```{r identifying_cycles}
# Assign numbers to all full cycles in the TDL data set
tdl_data <- identify_tdl_cycles(
  tdl_data,
  valve_column_name = 'valve_number',
  cycle_start_valve = 20,
  expected_cycle_length_minutes = 2.7,
  expected_cycle_num_valves = 9,
  timestamp_colname = 'TIMESTAMP'
)
```

Now the `tdl_data` object has two new columns: one is called `cycle_num` and
specifies each TDL cycle, and the other is called `elapsed_time` and indicates
the elapsed time of each TDL log relative to the first one.

## Calibrating TDL Cycles

Now that the individual cycles have been identified, the next step is to use the
reference valves to calibrate the TDL readings. In general, this procedure will
depend strongly on the individual TDL system, since each one may have different
types of reference tanks. The data used in this vignette was measured using the
TDL in Carl Bernacchi's lab in the Edward R. Madigan Laboratory at the
University of Illinois, Urbana-Champaign. It includes five reference tanks that
can be broken into three types:

* One is a certified tank whose total CO~2~ concentration and carbon isotope
  ratio are supplied from NOAA.
* One is a nitrogen tank that should have no carbon in it.
* The final type of reference is another CO~2~ tank whose isotype ratio was
  measured at UIUC using a different method; this tank is mixed with nitrogen in
  three different ratios to provide a range of carbon concentrations.

The full procedure for using these references to calibrate the TDL signal is
somewhat complicated. Fortunately, it is easy to determine and apply the
calibrations for a single TDL cycle using the `process_tdl_cycle_erml` function
from `PhotoGEA`. To apply this function to each TDL cycle in a larger data set
and then consolidate the results, we can use it in conjunction with `by` and
`consolidate`, which are also part of `PhotoGEA`. (For more information about
these functions, see the built-in help menu entries by typing
`?process_tdl_cycle_erml`, `?by.exdf`, or `?consolidate`, or check out the
_Common Patterns_ section of the
[Working With Extended Data Frames](working_with_extended_data_frames.html#processing-multiple-pieces-of-an-extended-data-frame)
vignette.) As with the previous function, many of the inputs here (such as
`noaa_cylinder_isotope_ratio`) must be obtained from whoever is managing the TDL
system being used:

```{r calibrating_tdl}
# Use the data from the calibration valves to determine calibrated carbon
# concentrations from all valves in each TDL cycle
processed_tdl <- consolidate(by(
  tdl_data,
  tdl_data[, 'cycle_num'],
  process_tdl_cycle_erml,
  valve_column_name = 'valve_number',
  noaa_valve = 2,
  calibration_0_valve = 20,
  calibration_1_valve = 21,
  calibration_2_valve = 23,
  calibration_3_valve = 26,
  raw_12c_colname = 'Conc12C_Avg',
  raw_13c_colname = 'Conc13C_Avg',
  noaa_cylinder_co2_concentration = 294.996,
  noaa_cylinder_isotope_ratio = -8.40,
  calibration_isotope_ratio = -11.505
))
```

The output from this function -- `processed_tdl` -- is a list of several `exdf`
objects that include calibrated TDL readings from each valve and information
about the calibration parameters that were determined while processing the data.

## Pairing Gas Exchange and TDL Data

When analyzing combined gas exchange and isotope discrimination data, a key step
is to combine TDL and gas exchange data that were measured at the same times.
The \code{pair_gasex_and_tdl} function performs this operation by locating the
TDL cycle whose timestamp is closest to each Licor file entry. Then, the ^12^C,
^13^C, total CO~2~, and $\delta^{13}$C values measured by the TDL from the
Licor's sample and reference lines during that cycle are added to the gas
exchange data as new columns.

```{r pairing_licor_and_tdl_data}
# Pair the Licor and TDL data by locating the TDL cycle corresponding to each
# Licor measurement
licor_data <- pair_gasex_and_tdl(licor_data, processed_tdl$tdl_data)
```

# Qualitative Validation

The data has now undergone several basic validation checks. Now we can make
qualitative validation checks by plotting various quantities and looking for any
outliers or otherwise strange measurement points. In this section, we will just
look at a few of the many possible variables that could be plotted. Other
possibilities can be found in some of the other PhotoGEA vignettes, such as
[Analyzing TDL Data](analyzing_tdl_data.html#checking-the-calibrated-data)
and
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html#additional-plots-for-qualitative-validation).

## Environment Control

One type of validation is to make sure that the leaf environment was properly
controlled. This could include temperature, humidity, and/or CO~2~
concentration. As an example, we will first look at the temperature. For these
measurements, the leaf temperature was set to 28 degrees C, so it is a good idea
to confirm that the leaf temperature was constant during the measurements.

```{r plot_temperature}
# Plot the leaf temperature, grouped by CO2_r setpoint
xyplot(
  TleafCnd ~ seq_num | curve_identifier,
  group = CO2_r_sp,
  data = licor_data$main_data,
  type = 'p',
  pch = 16,
  auto.key = list(space = 'right'),
  grid = TRUE,
  ylim = c(26, 30),
  xlab = 'Measurement number',
  ylab = paste('Leaf temperature', '[', licor_data$units$TleafCnd, ']')
)
```

The temperature control is not perfect for the last point in each curve, but
only deviates from the setpoint by 0.5 degrees C. This is an acceptable amount
of variation and is not a reason for concern.

We can also look at `CO2_r`, which should be 715 ppm for the first six points
and 450 ppm for the following points.

```{r plot_co2_r}
# Plot CO2_r, grouped by CO2_r setpoint
xyplot(
  CO2_r ~ seq_num | curve_identifier,
  group = CO2_r_sp,
  data = licor_data$main_data,
  type = 'p',
  pch = 16,
  auto.key = list(space = 'right'),
  grid = TRUE,
  ylim = c(400, 800),
  xlab = 'Measurement number',
  ylab = paste('CO2 concentration', '[', licor_data$units$CO2_r, ']')
)
```

Here we can see that the CO~2~ control was good for all the measurement points.

## TDL Readings

A key requisite for reliable mesophyll conductance values is that the TDL
readings should be stable with time at each CO~2~ setpoint. We can check this
by looking at the calibrated TDL readings.

```{r plot_tdl}
xyplot(
  Conc13C_Avg ~ elapsed_time,
  group = valve_number,
  data = processed_tdl$tdl_data$main_data,
  auto.key = list(space='right'),
  grid = TRUE,
  type = 'b',
  xlab = paste0('Elapsed time (', processed_tdl$tdl_data$units$elapsed_time, ')'),
  ylab = paste0(
    'Calibrated TDL 13C concentration (',
    processed_tdl$tdl_data$units$Conc13C_Avg, ')'
  ),
  par.settings = list(
    superpose.line = list(col = multi_curve_colors()),
    superpose.symbol = list(col = multi_curve_colors(), pch = 16)
  )
)
```

Here we can see that some of the valves are variable at the start of the TDL log
file (such as valves 26 and 23). However, they are quite stable by the time the
gas exchange logs begin (at around 25 minutes of elapsed time).

## Cleaning the Data

At this point, if any issues had been identified, we could clean the data by
removing any problematic points. In this example, this procedure is not
required. For examples of how to clean a data set by removing points, please see
the
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html#cleaning-the-licor-data)
vignette, or one of the other vignettes.

# Calculating Isotope Discrimination

Now, the `licor_data` `exdf` object includes the TDL measurements from the Licor
sample and reference gas lines that correspond to each log point. This
information can now be used to calculate $\Delta^{13}C$, the photosynthetic
isotope discrimination:

```{r isotope_discrimination}
# Calculate isotope discrimination
licor_data <- calculate_isotope_discrimination(licor_data)
```

We can also make a quick validity check here. Photosynthetic carbon isotope
discrimination should be not vary too much at each CO~2~ setpoint. We can check
this with a plot.

```{r plot_isotope_discrimination}
# Plot the photosynthetic isotope discrimination, grouped by CO2_r setpoint
xyplot(
  Delta_obs_tdl ~ seq_num | curve_identifier,
  group = CO2_r_sp,
  data = licor_data$main_data,
  type = 'p',
  pch = 16,
  auto.key = list(space = 'right'),
  grid = TRUE,
  ylim = c(16, 22),
  xlab = 'Measurement number',
  ylab = paste('Photosynthetic isotope discrimination', '[', licor_data$units$Delta_obs_tdl, ']')
)
```

Here there is a large amount of noise, but this is to be expected for these
measurements.

# Calculating Mesophyll Conductance

Now we have loaded the raw data, combined the two main data sources (gas
exchange and TDL measurements), performed some validity checks, and calculated
the isotope discrimination values. There are just a few more steps required to
calculate mesophyll conductance values. In particular, the mesophyll conductance
equations require the following inputs:

* The rate of day respiration ($R_d$).
* The CO~2~ compensation point in the absence of day respiration ($\Gamma^*$).
* The ternary correction factor ($t$).
* The observed photosynthetic carbon isotope discrimination under growth
  conditions ($\Delta_{obs}^{growth}$).

The value of day respiration $R_d$ must be known from other measurements;
typically it is determined by placing a leaf into a gas exchange measurement
chamber with the light off, and measuring the steady-state net assimilation
rate. Here we will use 1.2 $\mu$mol m$^{-2}$ s$^{-1}$.

```{r set_rd}
# Set day respiration rate
licor_data <- set_variable(
  licor_data,
  'RL',
  'micromol m^(-2) s^(-1)',
  value = 1.2
)
```

There are several possible approaches to choosing a value for $\Gamma^*$. Here
we will calculate its value from the leaf temperature and the Rubisco
specificity using the `calculate_gamma_star` function from `PhotoGEA`, using
92 M M$^{-1}$ for the specificity.

```{r set_gamma_star}
# Set Rubisco specificity
licor_data <- set_variable(
    licor_data,
    'rubisco_specificity_tl',
    'M / M',
    value = 92
)

# Calculate Gamma_star
licor_data <- calculate_gamma_star(licor_data)
```

The ternary correction factor $t$ can be calculated using the
`calculate_ternary_correction` function from `PhotoGEA`. This function has a few
of its own requirements, which can be met by first calling
`calculate_total_pressure` and `calculate_gas_properties`.

```{r set_t}
# Calculate total pressure
licor_data <- calculate_total_pressure(licor_data)

# Calculate gbc, gsc, Csurface
licor_data <- calculate_gas_properties(licor_data)

# Calculate t
licor_data <- calculate_ternary_correction(licor_data)
```

Like $R_d$, the observed photosynthetic isotope discrimination under growth
conditions ($\Delta_{obs}^{growth}$) must be measured from each leaf or plant
species. Ideally, this would be determined by setting the ambient CO~2~
concentration around the leaf to the ambient value and using the TDL to measure
the photosynthetic isotope discrimination $\Delta_{obs}$. In this data set, the
first six points in each file have `CO2_r` set to 715 ppm, which produces a
value of $C_a$ close to 420 ppm (the global atmospheric value at the time these
measurements were made). Thus, we can use the average value of $\Delta_{obs}$
across the first six points as the value of $\Delta_{obs}^{growth}$ for each
replicate.

```{r set_delta_obs_growth}
# Set Delta_obs_growth
licor_data <- set_variable(
  licor_data,
  'Delta_obs_growth',
  'ppt',
  id_column = 'curve_identifier',
  value_table = as.list(tapply(
    licor_data[licor_data[, 'CO2_r_sp'] == 715, 'Delta_obs_tdl'],
    licor_data[licor_data[, 'CO2_r_sp'] == 715, 'curve_identifier'],
    mean
  ))
)
```

Now we are finally able to calculate values of mesophyll conductance!

```{r calculate_gm}
# Calculate mesophyll conductance
licor_data <- calculate_gm_busch(licor_data, f = 0)
```

In this example, we have found that the mesophyll conductance values are more
reasonable with $f$ (the assumed fractionation due to day respiration) set to 0.
In general, $f$ is typically not zero, and can be varied during analysis.

# Examining the Results

One way to view the results is to simply plot each calculated value of the
mesophyll conductance.

```{r plot_gm_1}
# Plot the mesophyll conductance values, grouped by CO2_r setpoint
xyplot(
  gmc ~ seq_num | curve_identifier,
  group = CO2_r_sp,
  data = licor_data$main_data,
  type = 'p',
  pch = 16,
  auto.key = list(space = 'right'),
  grid = TRUE,
  xlab = 'Measurement number',
  ylab = paste('Mesophyll conductance to CO2', '[', licor_data$units$gmc, ']')
)
```

Another approach is to plot the average mesophyll conductance against the
average $C_i$ for each CO~2~ setpoint.

```{r plot_gm_2}
# Plot the average mesophyll conductance values against average Ci values,
# grouped by CO2_r setpoint
xyplot_avg_rc(
  licor_data[, 'gmc'],
  licor_data[, 'Ci'],
  licor_data[, 'CO2_r_sp'],
  licor_data[, 'curve_identifier'],
  type = 'b',
  auto = TRUE,
  grid = TRUE,
  pch = 16,
  ylab = paste('Mesophyll conductance to CO2', '[', licor_data$units$gmc, ']'),
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']')
)
```

As usual, the appropriate synthesis operations will depend on the details of the
project and how the data will be used.

# Customizing Your Script

Note that most of the commands in this vignette have been written in a general
way so they can be used as the basis for your own analysis script (see
[Commands From This Document]). In order to use them in your own script, some or
all of the following changes may be required. There may also be others not
specifically mentioned here.

## Input Files

The file paths specified in `licor_file_paths` and `tdl_file_paths` will need to
be modified so they point to your own files. One way to do this in your own
script is to simply write out relative or absolute paths to the files you wish
to load. For example, you could replace the previous definitions with these
ones:

```{r write_file_paths, eval = FALSE}
# Define a vector of paths to the files we wish to load
licor_file_paths <- c(
  'myfile1.xlsx',        # `myfile1.xlsx` must be in the current working directory
  'C:/documents/myfile2' # This is an absolute path to `myfile2`
)

tdl_file_paths <- c(
  'tdl_1.dat',     # `tdl_1.dat` must be in the current working directory
  'data/tdl_2.dat' # The current working directory must contain a subdirectory
)                  # called `data` that contains `tdl2.dat`
```

You may also want to consider using the `choose_input_licor_files` and
`choose_input_tdl_files` functions from `PhotoGEA`; these functions will create
pop-up browser windows where you can interactively select a set of files.
Sometimes this is more convenient than writing out file paths or names. For
example, you could replace the previous definitions with these ones:

```{r choose_files, eval = FALSE}
# Interactively define a vector of paths to the files we wish to load
licor_file_paths <- choose_input_licor_files()

tdl_file_paths <- choose_input_tdl_files()
```

Unfortunately, `choose_input_licor_files` and `choose_input_tdl_files` are only
available in interactive R sessions running on Microsoft Windows, but there is
also a platform-independent option: `choose_input_files`. See the _Translation_
section of the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html#translation)
vignette for more details.

## Mesophyll Conductance Calculations

Sometimes it may not be possible to estimate $\Delta^{obs}_{growth}$. In this
case, set `e_star_equation = 19` when calling `calculate_gm_busch` or use
`calculate_gm_ubierna` instead of `calculate_gm_busch`.

## Saving Results

You may want to use `write.csv` to save some or all of the results as `csv`
files. For example, the following command will allow you to interactively
choose an output filename for a `csv` file containing the combined gas exchange
and TDL data, along with the calculated values of isotope discrimination and
mesophyll conductance:

```{r saving_csv_files, eval = FALSE}
write.csv(licor_data, file.choose(), row.names = FALSE)
```

# Commands From This Document

The following code chunk includes all the central commands used throughout this
document. They are compiled here to make them easy to copy/paste into a text
file to initialize your own script. Annotation has also been added to clearly
indicate the four steps involved in data analysis, as described in the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html)
vignette.

```{r, eval = FALSE}
###
### PRELIMINARIES:
### Loading packages, defining constants, creating helping functions, etc.
###

<<setup>>

###
### TRANSLATION:
### Creating convenient R objects from raw data files
###

##
## Gas exchange log files
##

## IMPORTANT: When loading your own files, it is not advised to use
## `PhotoGEA_example_file_path` as in the code below. Instead, write out the
## names or use the `choose_input_licor_files` function.

<<licor_file_names>>

<<loading_licor_data>>

<<site_number>>

<<combining_licor_data>>

##
## Isotope discrimination log files
##

## IMPORTANT: When loading your own files, it is not advised to use
## `PhotoGEA_example_file_path` as in the code below. Instead, write out the
## names or use the `choose_input_tdl_files` function.

<<tdl_file_names>>

<<loading_tdl_data>>

<<combining_tdl_data>>

###
### VALIDATION:
### Organizing the data, checking its consistency and quality, cleaning it
###

<<adding_id_column>>

<<check_response_curve>>

<<organize_data>>

##
## Combining the gas exchange and isotope discrimination data
##

<<identifying_cycles>>

<<calibrating_tdl>>

<<pairing_licor_and_tdl_data>>

##
## Qualitative validation
##

<<plot_temperature>>

<<plot_co2_r>>

<<plot_tdl>>

###
### PROCESSING:
### Extracting new pieces of information from the data
###

<<isotope_discrimination>>

<<plot_isotope_discrimination>>

<<set_rd>>

<<set_gamma_star>>

<<set_t>>

<<set_delta_obs_growth>>

###
### SYNTHESIS:
### Using plots and statistics to help draw conclusions from the data
###

<<plot_gm_1>>

<<plot_gm_2>>
```
