---
title: "Analyzing C4 A-Ci Curves"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Analyzing C4 A-Ci Curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5.5,
  fig.align = "center"
)
```

# Overview

In this vignette, we will give an example showing how to analyze C~4~ _A-C~i~_
data using the `PhotoGEA` package. The commands in this vignette can be used to
initialize your own script, as described in [Customizing Your Script].

# Background

## Understanding C~4~ _A-C~i~_ Curves

An _A-C~i~ curve_ (or _CO~2~ response curve_) is a particular type of gas
exchange measurement where a living leaf is exposed to varying concentrations of
CO~2~. For each CO~2~ concentration in the sequence, the net assimilation rate
($A_n$), stomatal conductance to H~2~O ($g_{sw}$), intercellular CO~2~
concentration ($C_i$), and other important quantities are measured and recorded.
Typically, other environmental variables such as temperature, humidity, and
incident photosynthetically-active photon flux density (PPFD) are held constant
during the measurement sequence so that changes in photosynthesis can be
attributed to CO~2~ alone.

Because of their different cell structures and biochemical pathways, C~3~ and
C~4~ plants have very different responses to CO~2~. Here, we will only be
discussing C~3~ plants.

The full C~4~ photosynthetic pathway is quite complicated, consisting of at
least two hundred individual reactions, each of which may have an impact on a
measured _A-C~i~_ curve. However, simplified models for photosynthesis are
available and are much easier to understand and work with. These models tend to
be based around PEP carboxylase and rubisco kinetics, describing how the net
assimilation rate responds to the partial pressure of CO~2~ in the mesophyll or
bundle sheath cells. The most widely-used model is described in in
_Biochemical Models of Leaf Photosynthesis_ [@caemmerer_biochemical_2000].

This model provides a framework for understanding the changes in $A_n$ that
occur as a C~4~ plant is exposed to successively higher concentrations of CO~2~.
Overall, the photosynthetic response to CO~2~ under high light conditions can be
understood as follows:

- For low levels of CO~2~, CO~2~ assimilation is primarily limited by PEP
  carboxylase activity in the mesophyll cells.

- For higher levels of CO~2~, CO~2~ assimilation can be limited by Rubisco
  activity in the bundle sheath cells, PEP regeneration in the mesophyll cells,
  or light availability.

More specifically, the model provides equations that calculate the net
assimilation rate $A_n$ from the maximumum rate of PEP carboxylation
($V_{p,max}$), the maximum rate of Rubisco activity ($V_{c,max}$), the maximum
rate of PEP carboxylase regeneration ($V_{p,r}$), the electron transport rate
($J$), the total rate of mitochondrial respiration across the mesophyll and
bundle sheath cells ($R_L$), the partial pressure of CO~2~ in the mesophyll
($P_{cm}$), and several other parameters. The full equations are quite
complicated. $A_n$ is given by the smaller of the enzyme-limited rate ($A_c$)
and the light-limited rate ($A_j$). $A_c$ is co-limited by PEP carboxylase
activity, Rubisco activity, and PEP regeneration, while $A_j$ is co-limited by
electron transport in the mesophyll and bundle-sheath cells. Under certain
circumstances (when bundle sheath conductance is zero and all photosystem II
activity occurs in the mesophyll), then the complicated equations reduce to a
simpler situation where $A_n$ is given by the minimum of five separate rates:

- The PEP-carboxylase-limited rate ($A_{pc}$)

- The Rubisco-limited rate ($A_r$)

- The PEP-regeneration-limited rate ($A_{pr}$)

- The electron-transport-limited rate in the mesophyll ($A_{jm}$)

- The electron-transport-limited rate in the bundle sheath ($A_{jbs}$)

The simplified version can be helpful for understanding the rough model
behavior, but it is important to keep in mind that the real model is different,
and that $A_n$ is generally not equal to the minimum of these five rates in the
full model. The plot below shows an example of the full model output:

```{r, echo = FALSE}
library(PhotoGEA)
library(lattice)

npts <- 101

Tleaf = 30

inputs <- exdf(data.frame(
  PCm = seq(0, 500, length.out = npts),
  Qin = 1800,
  Tleaf = Tleaf,
  total_pressure = 1,
  oxygen = 21
))

inputs <- document_variables(
  inputs,
  c('', 'PCm',            'microbar'),
  c('', 'Qin',            'micromol m^(-2) s^(-1)'),
  c('', 'Tleaf',          'degrees C'),
  c('', 'total_pressure', 'bar'),
  c('', 'oxygen',         'percent')
)

inputs <- calculate_temperature_response(inputs, c4_temperature_param_vc, 'Tleaf')

alpha_psii <- 0
gbs <- 0.003
J_at_25 <- 220
RL_at_25 <- 1
Rm_frac <- 0.5
Vcmax_at_25 <- 30
Vpmax_at_25 <- 150
Vpr <- 80

assim <- calculate_c4_assimilation(
  inputs,
  alpha_psii, gbs, J_at_25, RL_at_25, Rm_frac, Vcmax_at_25, Vpmax_at_25, Vpr
)

xyplot(
  An + Ac + Aj + Apr + Apc + Ar + Ajm + Ajbs ~ PCm,
  data = cbind(inputs, assim)$main_data,
  type = 'l',
  lwd = c(6, 4, 4, 2, 2, 2, 2, 2),
  lty = c(1, 5, 5, 3, 3, 3, 3, 3),
  grid = TRUE,
  auto = TRUE,
  ylim = c(-10, 100),
  xlab = paste('CO2 pressure in mesophyll [', inputs$units$PCm, ']'),
  ylab = paste('Assimilation rate [', assim$units$An, ']'),
  par.settings = list(
    superpose.line = list(col = multi_curve_colors()),
    superpose.symbol = list(col = multi_curve_colors())
  )
)

```

(Note: this figure was generated using the `calculate_c4_assimilation` function
from the `PhotoGEA` package, and it represents the photosynethetic response of a
C~4~ leaf according to the model with
`RL_at_25` = `r RL_at_25` `r assim$units$RL_at_25`,
`Vpmax_at_25` = `r Vpmax_at_25` `r assim$units$Vpmax_at_25`,
`Vcmax_at_25` = `r Vcmax_at_25` `r assim$units$Vcmax_at_25`,
`Vpr` = `r Vpr` `r assim$units$Vpr`,
`J_at_25` = `r J_at_25` `r assim$units$J_at_25`, and a leaf
temperature of `r Tleaf` `r inputs$units$Tleaf`. Arrhenius temperature response
parameters were taken from @von_caemmerer_updating_2021.)

There are several important things to notice about this plot:

- For these conditions, $A_n$ is always equal to $A_c$. Thus, assimilation is
  enzyme-limited across the entire range of CO~2~ values.

- The enzyme-limited rate $A_c$ is always smaller than either of the three rates
  related to individual enzymes ($A_{pc}$, $A_{pr}$, and $A_r$).

- The light-limited rate $A_j$ is always smaller than either of the two rates
  related to electron transport in individual cell types ($A_{jm}$ and
  $A_{jb}$).

- The dependence of $A_r$, $A_j$, and $A_{pr}$ on CO~2~ concentration is very
  similar, where all three are nearly independent of CO~2~.

- The PEP-carboxylase-limited rate ($A_{pc}$) is the only rate that approaches
  zero, which happens as the CO~2~ concentration decreases.

- Regardless of which process may be limiting assimilation, $A_n$ _never_
  decreases as CO~2~ increases.

An important conclusion here is that PEP carboxylase always limits assimilation
at low CO~2~ concentrations, and that one or more of Rubisco activity, electron
transport, and PEP regeneration limits assimilation at high CO~2~
concentrations. These three processes each produce a plateau with a nearly flat
response to CO~2~.

One of the most common reasons to measure an _A-C~i~_ curve is to interpret it
in the context of this model. In other words, by fitting the model's equations
to a measured curve, it is possible to estimate values for $R_L$, $V_{pmax}$,
$V_{cmax}$, $J$, and others. See the documentation for
`calculate_c4_assimilation` for more information about these important
quantities.

From the discussion above, it is evident that determining the rate-limiting
process at high $C_i$ in a C~4~ _A-C~i~_ curve is difficult or impossible
because the three potentially limiting processes each exhibit nearly identical
responses to CO~2~. There are two main ways to deal with this issue:

- One way is to assume a particular limiting process based on outside
  information. For example, curves measured under high light conditions (where
  electron transport rates will be high), and at warm temperatures (where PEP
  regeneration rates will be high) are likely to be Rubisco-limited at high
  $C_i$.

- Another option is to use an empirical non-rectangular hyperbola to fit the
  curve. This enables estimates of $V_{max}$, the maximum rate of gross
  assimilation, which is related to the highest value of $A_n$ ($A_{nmax}$) by
  $A_{nmax} = V_{max} - R_L$. This quantity has no meaningful mechanistic
  interpretation but can be used to compare assimilation rates between groups of
  plants. When taking this approach, values of $V_{pmax}$ can still be estimated
  from the low $C_i$ part of the curve using the mechanistic model.

These approaches will be demonstrated below in this vignette.

## Practicalities

There are a few other important practicalities to keep in mind when thinking
about CO~2~ response curves.

One point is that C~4~ photosynthesis models generally predict the response of
assimilation to the partial pressure of CO~2~ in the mesophyll ($P_{cm}$), but
gas exchange measurements can only determine the CO~2~ concentration in the
leaf's intercellular spaces ($C_i$). Thus, an extra step is required when
interpreting _A-C~i~_ curves. If the mesophyll conductance to CO~2~ ($g_{mc}$)
and the total pressure ($P$) are known, then it is possible to calculate values
of $P_{cm}$ from $A_n$, $C_i$, $P$, and $g_{mc}$. Otherwise, it is also possible
to assume an infinite mesophyll conductance; in this case, $C_m = C_i$, and the
estimated values of $V_{cmax}$ and other parameters can be considered to be
"effective values" describing the plant's response to intercellular CO~2~.

Another important point is that plants generally do not appreciate being starved
of CO~2~, so it is not usually possible to start a response curve at low CO~2~
and proceed upwards. A more typical approach is to:

1. Begin at ambient atmospheric CO~2~ levels.

2. Decrease towards a low value.

3. Return to ambient levels and wait for the plant to reacclimate; this waiting
   period is usually accomplished by logging several points at ambient CO~2~
   levels.

4. Increase to higher values.

When taking this approach, it therefore becomes necessary to remove the extra
points measured at ambient CO~2~ levels and to reorder the points according to
their CO~2~ values before plotting or analyzing them.

## The Data

_A-C~i~_ curves are commonly measured using a portable photosynthesis system
such as the Licor Li-6800. These machines record values of $A_n$, $g_{sw}$,
$C_i$, and many other important quantities. They produce two types of output
files: plain-text and Microsoft Excel. It is often more convenient to work with
the Excel files since the entries can be easily modified (for example, to remove
an extraneous row or add a new column). On the other hand, it can be more
difficult to access the file contents using other pieces of software such as R.
However, the `PhotoGEA` package reduces this barrier by including tools for
reading Licor Excel files in R, which will be demonstrated in the following
section.

# Loading Packages

As always, the first step is to load the packages we will be using. In addition
to `PhotoGEA`, we will also use the `lattice` package for generating plots.

```{r setup, eval = FALSE}
# Load required packages
library(PhotoGEA)
library(lattice)
```

If the `lattice` package is not installed on your R setup, you can install it by
typing `install.packages('lattice')`.

# Loading Licor Data

The `PhotoGEA` package includes two files representing _A-C~i~_ curves measured
using two Li-6800 instruments. The data is stored in Microsoft Excel files, and
includes curves measured from two different crop species (sorghum and maize)
and several different plots of each. Each curve is a sixteen-point CO~2~
response curve; in other words, the CO~2~ concentration in the air surrounding
the leaf was varied, and $A_n$ (among other variables) was measured at each
CO~2~ setpoint. Although these two files are based on real data, noise was added
to it since it is unpublished, so these files should only be used as examples.

The files will be stored on your computer somewhere in your R package
installation directory, and full paths to these files can be obtained with
`PhotoGEA_example_file_path`:

```{r licor_file_names}
# Define a vector of paths to the files we wish to load; in this case, we are
# loading example files included with the PhotoGEA package
file_paths <- c(
  PhotoGEA_example_file_path('c4_aci_1.xlsx'),
  PhotoGEA_example_file_path('c4_aci_2.xlsx')
)
```

(**Note:** When loading your own files for analysis, it is not advisable to use
`PhotoGEA_example_file_path` as we have done here. Instead, file paths can be
directly written, or files can be chosen using an interactive window. See
[Input Files] below for more information.)

To actually read the data in the files and store them in R objects, we will use
the `read_gasex_file` function from `PhotoGEA`. Since there are multiple files
to read, we will call this function once for each file using `lapply`:

```{r loading_licor_data}
# Load each file, storing the result in a list
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_gasex_file(fpath, 'time')
})
```

The result from this command is an R list of "extended data frames" (abbreviated
as `exdf` objects). The `exdf` class is a special data structure defined by the
`PhotoGEA` package. In many ways, an `exdf` object is equivalent to a data
frame, with the major difference being that an `exdf` object includes the units
of each column. For more information, type `?exdf` in the R terminal to access
the built-in help menu entry, or check out the
[Working With Extended Data Frames](working_with_extended_data_frames.html)
vignette.

Generally, it is more convenient to work with a single `exdf` object rather than
a list of them, so our next step will be to combine the objects in the list.
This action can be accomplished using the `rbind` function, which combines
table-like objects by their rows; in other words, it stacks two or more tables
vertically. This action only makes sense if the tables have the same columns, so
before we combine the `exdf` objects, we should make sure this is the case.

The `PhotoGEA` package includes a function called `identify_common_columns` that
can be used to get the names of all columns that are present in all of the Licor
files. Then, we can extract just those columns, and then combine the `exdf`
objects into a single one.

```{r combining_licor_data}
# Get the names of all columns that are present in all of the Licor files
columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
licor_data <- do.call(rbind, licor_exdf_list)
```

Now we have a single R object called `licor_data` that includes all the data
from several Licor Excel files. For more information about consolidating
information from multiple files, see the _Common Patterns_ section of the
[Working With Extended Data Frames](working_with_extended_data_frames.html#combining-data-from-several-files)
vignette.

# Validating the Data

Before attempting to fit the curves, it is a good idea to do some basic checks
of the data to ensure it is organized properly and that it was measured
properly.

## Basic Checks

First, we should make sure there is a column in the data whose value uniquely
identifies each curve. In this particular data set, several "user constants"
were defined while making the measurements that help to identify each curve:
`instrument`, `species`, and `plot`. However, neither of these columns alone are
sufficient to uniquely identify each curve. We can solve this issue by creating
a new column that combines the values from each of these:

```{r adding_id_column}
# Create a new identifier column formatted like `instrument - species - plot`
licor_data[ , 'curve_identifier'] <-
  paste(licor_data[ , 'instrument'], '-', licor_data[ , 'species'], '-', licor_data[ , 'plot'])
```

The next step is to make sure that this column correctly identifies each
response curve. To do this, we can use the `check_response_curve_data` function
from `PhotoGEA`. Here we will supply the name of a column that should uniquely
identify each response curve (`curve_identifier`), the expected number of points
in each curve (`16`), the name of a "driving" column that should follow the same
sequence in each curve (`CO2_r_sp`). If the data passes the checks, this
function will have no output and will not produce any messages. (For more
information, see the built-in help menu entry by typing
`?check_response_curve_data`.)

```{r check_response_curve}
# Make sure the data meets basic requirements
check_response_curve_data(licor_data, 'curve_identifier', 16, 'CO2_r_sp')
```

However, if `check_response_curve_data` detects an issue, it will print a
helpful message to the R terminal. For example, if we had specified the wrong
number of points or the wrong identifier column, we would get error messages:

```{r check_response_curve_error, error = TRUE}
check_response_curve_data(licor_data, 'curve_identifier', 15)

check_response_curve_data(licor_data, 'species', 16)

check_response_curve_data(licor_data, 'curve_identifier', 16, 'Ci')
```

## Plotting the _A-C~i~_ Curves

One qualitative way to check the data is to simply create a plot of the _A-C~i~_
curves. In this situation, the `lattice` library makes it simple to include each
curve as its own separate subplot of a figure. For example:

```{r plot_aci}
# Plot all A-Ci curves in the data set
xyplot(
  A ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']'),
  ylab = paste('Net CO2 assimilation rate [', licor_data$units$A, ']')
)
```

Whoops! Why do these curves look so strange? Well, some of the issues are
related to the sequence of CO~2~ values that was used when measuring the curves.
As discussed in [Practicalities], there are several repeated points logged at
the same CO~2~ concentration, and the points are not logged in order of
ascending or descending concentration. In fact, the sequence of CO~2~ setpoints
is as follows:

```{r}
licor_data[licor_data[, 'curve_identifier'] == 'ripe2 - maize - 2', 'CO2_r_sp']
```

Ideally, we would like to remove the ninth and tenth points (where the setpoint
has been reset to 400 to allow the leaf to reacclimate to ambient CO~2~ levels),
and reorder the data so it is arranged from low to high values of `Ci`. This can
be done using the `organize_response_curve` function from `PhotoGEA`:

```{r reorganizing_curves}
# Remove points with duplicated `CO2_r_sp` values and order by `Ci`
licor_data <- organize_response_curve_data(
    licor_data,
    'curve_identifier',
    c(9, 10),
    'Ci'
)
```

Now we can plot them again:

```{r}
<<plot_aci>>
```

They still look a bit strange. Some of this is related to the noise that was
intentionally added to the data. Nevertheless, there are a few points that we
should probably exclude before attempting to fit the curves. One issue is that
the model _never_ predicts a decrease in $A$ when $Ci$ increases. So, it is
usually a good idea to exclude any points at high $Ci$ where $A$ is observed to
decrease by a large amount. This dropoff in assimilation is due to one or more
processes that are not captured by the model, and the fits will be unreliable.
We will remove these points later, after we have made several other data quality
checks.

## Additional Plots for Qualitative Validation

Sometimes a Licor will override the temperature or humidity controls while
making measurements; in this case, conditions inside the measurement chamber may
not be stable, and we may wish to exclude some of these points. We can check for
these types of issues by making more plots. In the following sections, we will
generate several different plots to check each curve for quality.

### Humidity Control

```{r plot_humidity}
# Make a plot to check humidity control
xyplot(
  RHcham + `Humidifier_%` + `Desiccant_%` ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(0, 100),
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']')
)
```

Here, `Humidifier_%` and `Desiccant_%` represent the flow from the humidifier
and desiccant columns, where a value of 0 indicates that the valve to the column
is fully closed and a value of 100 indicates that the valve to the column is
fully opened. `RHcham` represents the relative humidity inside the chamber as a
percentage (in other words, as a value between 0 and 100).

When these curves were measured, a chamber humidity setpoint was specified. So,
when looking at this plot, we should check that the relative humidity is fairly
constant during each curve. Typically, this should be accompanied by relatively
smooth changes in the valve percentages as they accomodate changes in ambient
humidity and leaf photosynthesis. In this plot, all the data looks good.

### Temperature Control

```{r plot_temperature}
# Make a plot to check temperature control
xyplot(
  TleafCnd + Txchg ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(25, 40),
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']'),
  ylab = paste0('Temperature (', licor_data$units$TleafCnd, ')')
)
```

Here, `TleafCnd` is the leaf temperature measured using a thermocouple, and
`Txchg` is the temperature of the heat exhanger that is used to control the air
temperature in the measurement instrument. When these curves were measured, an
exchanger setpoint was specified. So, when looking at this plot, we should check
that `Txchg` is constant during each curve and that the leaf temperature does
not vary in an erratic way. In this plot, all the data looks good.

### CO~2~ Control

```{r plot_co2}
# Make a plot to check CO2 control
xyplot(
  CO2_s + CO2_r + CO2_r_sp ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']'),
  ylab = paste0('CO2 concentration (', licor_data$units$CO2_r, ')')
)
```

Here, `CO2_s` is the CO~2~ concentration in the sample cell, `CO2_r` is the
CO~2~ concentration in the reference cell, and `CO2_r_sp` is the setpoint for
`CO2_r`. When these curves were measured, a sequence of `CO2_r` values was
specified, so, when looking at this plot, we should check that `CO2_r` is
close to `CO2_r_sp`. We also expect that `CO2_s` should be a bit lower than
`CO2_r` because the leaf in the sample chamber is assimilating CO~2~, which
should reduce its concentration in the surrounding air. (An exception to this
rule occurs at very low values of `CO2_r_sp`, since in this case there is not
enough carbon available to assimilate, and the leaf actually releases CO~2~ due
to respiration.) In this plot, all the data looks good.

### Stability

```{r plot_stability}
# Make a plot to check stability criteria
xyplot(
  `A:OK` + `gsw:OK` + Stable ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']')
)
```

When measuring response curves with a Licor, it is possible to specify stability
criteria for each point in addition to minimum and maximum wait times. In other
words, once the set point for the driving variable is changed, the machine waits
until the stability criteria are met; there is a minimum waiting period, and
also a maximum to prevent the machine from waiting for too long.

When these curves were measured, stability criteria were supplied for the net
assimilation rate `A` and the stomatal conductance `gsw`. The stability status
for each was stored in the log file because the appropriate logging option for
stability was set. Now, for each point, it is possible to check whether
stability was achieved or whether the point was logged because the maximum
waiting period had been met. If the maximum waiting period is reached and the
plant has still not stabilized, the data point may be unreliable, so it can be
helpful to check this information.

In the plot, `A:OK` indicates whether `A` was stable (0 for no, 1 for yes),
`gsw:OK` indicates whether `gsw` was stable (0 for no, 1 for yes), and
`Stable` indicates the total number of stability conditions that were met. So,
we are looking for points where `Stable` is 2. Otherwise, we can check the other
traces to see whether `A` or `gsw` was unstable.

Comparing these plots with the ones in [Plotting the _A-C~i~_ Curves], it seems
that none of the unstable points look particularly strange, so we will not
remove them.

## Cleaning the Licor Data

While checking over the plots in the previous sections, we noticed that some of
the curves exhibit a sharp decrease in $A$ at high values of $C_i$.

We can use the `remove_points` function from `PhotoGEA` to exclude these points.
It just so happens that all of these points where $A_n$ suddenly decreases at
high $C_i$ were measured by the `ripe1` instrument and occur at the
highest CO~2~ setpoint value, so it is easy to specify them all at once.

```{r remove_points}
# Remove points where `instrument` is `ripe1` and `CO2_r_sp` is 1800
licor_data <- remove_points(
  licor_data,
  list(instrument = 'ripe1', CO2_r_sp = 1800),
  method = 'exclude'
)
```

# Fitting Licor Data

Now that we have checked the data quality, we are ready to perform the fitting.
In order to fit the curves, there are several required pieces of information
that are not included in the Licor data files as produced by the instrument:
temperature-dependent values of important photosynthetic parameters such as
$\gamma^*$, values of the total pressure, and values of the partial pressure of
CO~2~ in the mesophyll $P_{cm}$. However, the `PhotoGEA` package includes three
functions to help with these calculations: `calculate_temperature_response`,
`calculate_total_pressure`, and `apply_gm`. Each of these requires an `exdf`
object containing Licor data. The units for each required column will be checked
in an attempt to avoid unit-related errors. More information about these
functions can be obtained from the built-in help system by typing
`?calculate_temperature_response`, `?calculate_total_pressure`, or `?apply_gm`.

First, we can use `calculate_temperature_response`:

```{r calculate_t_response}
# Calculate temperature-dependent values of C4 photosynthetic parameters
licor_data <- calculate_temperature_response(licor_data, c4_temperature_param_vc)
```

With these simple commands, we have used the leaf temperature from the Licor
file to calculate values of several key C~4~ photosynthetic parameters
(`r paste(names(c4_temperature_param_vc), collapse = ', ')` according to the
temperature response parameters specified in @von_caemmerer_updating_2021.
Notice that this command calculated values of the mesophyll conductance to CO2
(`gmc`) in units of mol m$^{-2}$ s$^{-1}$ bar$^{-1}$. These values were estmated
for _Setaria viridis_ and may not be appropriate for all C~4~ plants.
Nevertheless, we will use them here.

Now we can use `calculate_total_pressure` and `apply_gm` to calculate `PCm`:

```{r apply_gm}
# Calculate the total pressure in the Licor chamber
licor_data <- calculate_total_pressure(licor_data)

# Calculate PCm
licor_data <- apply_gm(
  licor_data,
  gmc_at_25 = 1, # mol / m^2 / s / bar (von Caemmerer 2021)
  photosynthesis_type = 'C4' # Indicate C4 photosynthesis
)
```

Together, these functions have added several new columns to `licor_data`,
including `gmc`, `PCm`, `gamma_star`, and others. With this information, we are
now ready to fit the curves.

Here we will demonstrate three different approaches to curve fitting. For these
fits, we will be using the `fit_c4_aci` and `fit_c4_aci_hyperbola` functions
from the `PhotoGEA` package, which each fit single response curves to estimate
the values of key photosynthetic parameters.

To apply these functions to each curve in a larger data set and then consolidate
the results, we can use them in conjunction with `by` and `consolidate`, which
are also part of `PhotoGEA`. (For more information about these functions, see
the built-in help menu entries by typing `?fit_c4_aci`, `?fit_c4_aci_hyperbola`,
`?by.exdf`, or `?consolidate`, or check out the _Common Patterns_ section of the
[Working With Extended Data Frames](working_with_extended_data_frames.html#processing-multiple-pieces-of-an-extended-data-frame)
vignette.) Together, these functions will split apart the main data using
the curve identifier column we defined before ([Basic Checks]), fit each
_A-C~i~_ curve using the models discussed in
[Understanding C~4~ _A-C~i~_ Curves], and return the resulting parameters and
fits.

For viewing the resulting fits, we will be using the `plot_c4_aci_fit` and
`plot_c4_aci_hyperbola_fit` functions from `PhotoGEA`.

## Mechanistic Fit, Assuming Rubisco Limitations

To ensure that the fitted assimilation rates are limited by Rubisco activity at
high $C_i$, we can set $V_{pr}$ and $J$ to extremely high values, which
will ensure that PEP regeneration and light availability are never limiting.
At the same time, we will allow $V_{cmax}$ to vary during the fit. This is the
default behavior of `fit_c4_aci`, but here we will use the `fit_options` input
argument to make it clear that we are requiring Rubisco limitations.

```{r fit_aci_rubisco}
# The default optimizer uses randomness, so we will set a seed to ensure the
# results from this fit are always identical
set.seed(1234)

# Fit the C4 A-Ci curves with the mechanistic model, assuming Rubisco limitations
c4_aci_results_rubisco <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c4_aci,                       # The function to apply to each chunk of `licor_data`
  Ca_atmospheric = 420,             # Used to estimate the operating point
  fit_options = list(Vcmax_at_25 = 'fit', Vpr = 1000, J_at_25 = 1000)
))
```

Having made the fits, we can view the results:

```{r plot_aci_fit_rubisco}
# Plot the C4 A-Ci fits that were made assuming Rubisco limitations
plot_c4_aci_fit(
  c4_aci_results_rubisco,
  'curve_identifier',
  'Ci',
  ylim = c(-10, 100),
  main = 'Mechanistic fits assuming Rubisco limitations'
)
```

Overall, these fits look good. Note that because Rubisco activity and PEP
carboxylase activity co-limit the net assimilation rate, $A_n$ is always smaller
than either $A_r$ or $A_{pc}$ in these results.

## Mechanistic Fit, Assuming Light Limitations

To ensure that the fitted assimilation rates are limited by light at high $C_i$,
we can set $V_{pr}$ and $V_{cmax}$ to extremely high values, which will ensure
that PEP regeneration and Rubisco activity are never limiting. At the same time,
we will allow $J$ to vary during the fit. This can be done using the
`fit_options` input argument to `fit_c4_aci`.

```{r fit_aci_light}
# The default optimizer uses randomness, so we will set a seed to ensure the
# results from this fit are always identical
set.seed(1234)

# Fit the C4 A-Ci curves with the mechanistic model, assuming light limitations
c4_aci_results_light <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c4_aci,                       # The function to apply to each chunk of `licor_data`
  Ca_atmospheric = 420,             # Used to estimate the operating point
  fit_options = list(Vcmax_at_25 = 1000, Vpr = 1000, J_at_25 = 'fit')
))
```

Having made the fits, we can view the results:

```{r plot_aci_fit_light}
# Plot the C4 A-Ci fits that were made assuming Rubisco limitations
plot_c4_aci_fit(
  c4_aci_results_light,
  'curve_identifier',
  'Ci',
  ylim = c(-10, 100),
  main = 'Mechanistic fits assuming light limitations'
)
```

Overall, these fits look good. Note that because electron transport and PEP
carboxylase are independent in the model, there is a sharp transition from
$A_n = A_{pc}$ to $A_n = A_j$ in these fits, in contrast to the Rubisco fits
above, where there is a smooth transition between limiting states.

## Semi-Empirical Fit

We can also take an empirical approach to curve fitting, where we use a
non-rectangular hyperbola instead of the mechanistic model.

```{r fit_aci_hyperbola}
# The default optimizer uses randomness, so we will set a seed to ensure the
# results from this fit are always identical
set.seed(1234)

# Fit the C4 A-Ci curves with the empirical model
c4_aci_results_hyperbola <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c4_aci_hyperbola              # The function to apply to each chunk of `licor_data`
))
```

Having made the fits, we can view the results:

```{r plot_aci_fit_hyperbola}
# Plot the C4 A-Ci fits that were made using the empirical model
plot_c4_aci_hyperbola_fit(
  c4_aci_results_hyperbola,
  'curve_identifier',
  ylim = c(-10, 100),
  main = 'Empirical fits using a hyperbola'
)
```

The hyperbola is used to estimate $V_{max}$, which characterizes the maximum
gross assimilation rate that occurs at high $C_i$, such that $A_{max} =
V_{max} - R_L$ is the highest net assimilation rate. It is common to pair this
with an estimate $V_{pmax}$ made by fitting a mechanistic equation to
the low-$C_i$ part of the curve:

$A_n = \frac{V_{pmax} \cdot C_i}{C_i + K_p} - R_L$

The full mechanistic model reduces to this particular equation when the
following conditions are met:

- The bundle sheath conductance ($g_{bs}$) is zero.

- No photosystem II activity occurs in the bundle sheath ($\alpha_{PSII} = 0$).

- The CO~2~ concentration in the mesophyll is identical to the intercellular
  CO~2~ concentration. This is the case when mesophyll conductance is infinitely
  large.

Here we will fit this equation to just the measured points where $C_i$ is below
60 ppm. To do this, we will also need to recalculate $PCm$ using an infinite
value of mesophyll conductance.

```{r fit_aci_vpmax}
# Get a subset of the data where Ci is below the threshold value
licor_data_low_ci <- licor_data[licor_data[, 'Ci'] <= 60, , TRUE]

# Recalculate PCm using infinite mesophyll conductance
licor_data_low_ci <- apply_gm(
  licor_data_low_ci,
  gmc_at_25 = Inf,
  photosynthesis_type = 'C4' # Indicate C4 photosynthesis
)

# The default optimizer uses randomness, so we will set a seed to ensure the
# results from this fit are always identical
set.seed(1234)

# Fit the C4 A-Ci curves with the mechanistic Vpmax equation
c4_aci_results_vpmax <- consolidate(by(
  licor_data_low_ci,                       # The `exdf` object containing the curves
  licor_data_low_ci[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c4_aci,                              # The function to apply to each chunk of `licor_data`
  fit_options = list(
    Vcmax_at_25 = 1000,
    Vpr = 1000,
    J_at_25 = 1000,
    alpha_psii = 0,
    gbs = 0
  )
))
```

Having made the fits, we can view the results:

```{r plot_aci_fit_vpmax}
# Plot the C4 A-Ci fits that were made to the low Ci part of the curves
plot_c4_aci_fit(
  c4_aci_results_vpmax,
  'curve_identifier',
  'Ci',
  ylim = c(-10, 60),
  main = 'Mechanistic fits to the low Ci part of each curve'
)
```

# Examining the Results

## Accessing Best-Fit Estimates and Confidence Intervals

Each fit returns best-fit values and confidence limits for each parameter, which
are stored in `parameters` element of the output list, another `exdf` object.
These values can be viewed directly from R. For example, we can look at the
estimated $V_{pmax}$ values from each mechanistic fit as follows:

```{r view_j_lim}
# Specify columns to view
vpmax_columns <- c(
  'curve_identifier',
  'Vpmax_at_25_lower',
  'Vpmax_at_25',
  'Vpmax_at_25_upper'
)

# View Vpmax values from the mechanistic fit that assumed Rubisco limitations
c4_aci_results_rubisco$parameters[, vpmax_columns]

# View Vpmax values from the mechanistic fit that assumed light limitations
c4_aci_results_light$parameters[, vpmax_columns]

# View Vpmax values from the mechanistic fit to the low-Ci part of the curves
c4_aci_results_vpmax$parameters[, vpmax_columns]
```

Here we can see that the estimates made from the mechanistic fit with Rubisco
limitations are generally the largest, while those made from just the low-$C_i$
points are the lowest.

Estimates of other parameters, such as $V_{cmax}$, $J$, $V_{max}$, and
$R_L$ can be accessed with similar commands.

## Visualizing Average Best-Fit Parameters

We can also visualize the best-fit parameters using plots. One way to do this is
by using the `barchart_with_errorbars` function from `PhotoGEA` to create
barcharts of the average values for each species. This function will ignore any
`NA` values from curves where a parameter could not be reliably estimated, and
the error bars will show the standard error of the mean for each species. Here
we will take a look at the $V_{cmax}$ estimates.

```{r vcmax_barchart}
# Make a barchart showing average Vcmax values
barchart_with_errorbars(
  c4_aci_results_rubisco$parameters[, 'Vcmax_at_25'],
  c4_aci_results_rubisco$parameters[, 'species'],
  ylim = c(0, 50),
  xlab = 'Species',
  ylab = paste0('Vcmax at 25 degrees C (', c4_aci_results_rubisco$parameters$units$Vcmax_at_25, ')')
)
```

Another option is to create box-whisper plots using the `bwplot` function from
the `lattice` package:

```{r vcmax_bwplot}
# Make a boxplot showing the distribution of Vcmax values
bwplot(
  Vcmax_at_25 ~ species,
  data = c4_aci_results_rubisco$parameters$main_data,
  ylim = c(0, 50),
  xlab = 'Species',
  ylab = paste0('Vcmax at 25 degrees C (', c4_aci_results_rubisco$parameters$units$Vcmax_at_25, ')')
)
```

## Calculating Average Best-Fit Parameters

We can also calculate average values and standard errors of the best-fit
parameters for each species using the `basic_stats` function from `PhotoGEA`:

```{r calculate_stats}
# Compute the average and standard error of each parameter for each species
# from one of the mechanistic fits
c4_aci_averages_rubisco <- basic_stats(c4_aci_results_rubisco$parameters, 'species')

# View the averages and errors
columns_to_view <- c(
  'species',
  'Vcmax_at_25_avg', 'Vcmax_at_25_stderr',
  'Vpmax_at_25_avg', 'Vpmax_at_25_stderr',
  'RL_at_25_avg', 'RL_at_25_stderr'
)

c4_aci_averages_rubisco[ , columns_to_view]
```

## Other Ideas for Synthesizing Results

Statistical tests to check for differences between groups can be performed
within R using other packages such as `onewaytests` or `DescTools`.
Alternatively, the parameter values can be exported to a comma-separated-value
(CSV) file and analyzed in another software environment like `jmp`.

# Customizing Your Script

Note that most of the commands in this vignette have been written in a general
way so they can be used as the basis for your own analysis script (see
[Commands From This Document]). In order to use them in your own script, some or
all of the following changes may be required. There may also be others not
specifically mentioned here.

## Input Files

The file paths specified in `file_paths` will need to be modified so they point
to your Licor files. One way to do this in your own script is to simply write
out relative or absolute paths to the files you wish to load. For example, you
could replace the previous definition of `file_paths` with this one:

```{r write_file_paths, eval = FALSE}
# Define a vector of paths to the files we wish to load
file_paths <- c(
  'myfile1.xlsx',        # `myfile1.xlsx` must be in the current working directory
  'C:/documents/myfile2' # This is an absolute path to `myfile2`
)
```

You may also want to consider using the `choose_input_licor_files` function from
`PhotoGEA`; this function will create a pop-up browser window where you can
interactively select a set of files. Sometimes this is more convenient than
writing out file paths or names. For example, you could replace the previous
definition of `file_paths` with this one:

```{r choose_licor_files, eval = FALSE}
# Interactively define a vector of paths to the files we wish to load
file_paths <- choose_input_licor_files()
```

Unfortunately, `choose_input_licor_files` is only available in interactive R
sessions running on Microsoft Windows, but there is also a platform-independent
option: `choose_input_files`. See the _Translation_ section of the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html#translation)
vignette for more details.

## Curve Identifier

Depending on which user constants are defined in your Licor Excel files, you may
need to modify the definition of the `curve_identifier` column.

## Data Cleaning

Depending on the qualitative data checks, you may need to change the input
arguments to `remove_points`. It might also not be necessary to remove the
unstable points before performing the fits. Often, it is helpful to not perform
any data cleaning at first, and then remove problematic points if they seem to
cause problems with the fits.

## Averages and Standard Errors

Depending on how your data is organized, you may want to change the column used
to divide the data when calculating averages and standard errors.

## Plots

You may need to change the axis limits in some or all of the plots.
Alternatively, you can remove them, allowing `xyplot` to automatically choose
them for you.

You may also want to consider using the `pdf_print` function from `PhotoGEA` to
save some or all of your plots as PDFs. See the help page for more info:
`?pdf_print`.

## Saving Results

You may want to use `write.csv.exdf` to save some or all of the fitting results
as CSV files. When writing the contents of an `exdf` object to a CSV file,
using `write.csv.exdf` rather than `write.csv` will ensure that units are
included with each column.

For example, the following commands will save the results from one of the
mechanistic fits, allowing you to interactively choose output filenames for the
resulting CSV files:

```{r saving_csv_files, eval = FALSE}
write.csv.exdf(c4_aci_results_rubisco$fits, file.choose())
write.csv.exdf(c4_aci_results_rubisco$parameters, file.choose())
write.csv.exdf(c4_aci_averages_rubisco, file.choose())
```

# Commands From This Document

The following code chunk includes all the central commands used throughout this
document. They are compiled here to make them easy to copy/paste into a text
file to initialize your own script. Annotation has also been added to clearly
indicate the four steps involved in data analysis, as described in the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html)
vignette.

```{r, eval = FALSE}
###
### PRELIMINARIES:
### Loading packages, defining constants, creating helping functions, etc.
###

<<setup>>

###
### TRANSLATION:
### Creating convenient R objects from raw data files
###

## IMPORTANT: When loading your own files, it is not advised to use
## `PhotoGEA_example_file_path` as in the code below. Instead, write out the
## names or use the `choose_input_licor_files` function.

<<licor_file_names>>

<<loading_licor_data>>

<<combining_licor_data>>

###
### VALIDATION:
### Organizing the data, checking its consistency and quality, cleaning it
###

<<adding_id_column>>

<<check_response_curve>>

<<reorganizing_curves>>

<<plot_aci>>

<<plot_humidity>>

<<plot_temperature>>

<<plot_co2>>

<<plot_stability>>

## IMPORTANT: When analyzing your own files, it is not advised to remove any
## points for the initial fits. Only remove unstable or unusual points if it is
## necessary in order to get good fits.

<<remove_unstable_points>>

<<remove_points>>

###
### PROCESSING:
### Extracting new pieces of information from the data
###

<<calculate_t_response>>

<<apply_gm>>

##
## Mechanistic fits, assuming Rubisco limitations
##

<<fit_aci_rubisco>>

<<plot_aci_fit_rubisco>>

##
## Mechanistic fits, assuming light limitations
##

<<fit_aci_light>>

<<plot_aci_fit_light>>

##
## Semi-empirical fits
##

<<fit_aci_hyperbola>>

<<plot_aci_fit_hyperbola>>

<<fit_aci_vpmax>>

<<plot_aci_fit_vpmax>>

###
### SYNTHESIS:
### Using plots and statistics to help draw conclusions from the data
###

<<vcmax_barchart>>

<<vcmax_bwplot>>

<<calculate_stats>>
```

# References
