---
title: "Analyzing TDL Data"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Analyzing TDL Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

In this vignette, we will give an example showing how to analyze data from a
tunable diode laser (TDL) using the `PhotoGEA` package. The commands in this
vignette can be used to initialize your own script, as described in
[Customizing Your Script].

# Background

Because the process of photosynthesis tends to preferentially assimilate ^12^C
more often than the rarer and heavier stable isotope ^13^C, carbon isotope
discrimination measurements have proven to be incredibly useful in the fields of
plant biology and climate science [@evans_temperature_2013]. Although several
methods are available for measuring the relative concentrations of ^12^C and
^13^C gas mixtures, tunable diode laser (TDL) absorption spectroscopy has
emerged as one of the most popular due to its high sensitivity and relatively
short measurement times. In this technique, the emission wavelength of the laser
is tuned over the characteristic absorption lines of species present in a gas,
and the resulting decrease in light transmission can be used to determine the
gas concentration and even other properties such as temperature. For more
information about the general technique, please see its
[Wikipedia page](https://en.wikipedia.org/wiki/Tunable_diode_laser_absorption_spectroscopy).

The gas concentrations reported directly by a TDL measurement system tend to be
influenced by electrical noise and subtle changes in air properties within the
measurement room. For this reason, it is usually necessary to make frequent
measurements of gas mixtures with known concentrations so the output signals can
be calibrated. In fact, most TDL systems used in plant biology include a gas
handling system with multiple valves that can be attached to various gas
sources, including calibration tanks as well as the lines with gas whose
isotopic makeup is to be determined. During measurement, the gas handling system
periodically cycles through each valve in succession, and a data logger is used
to record the TDL output signal from each valve. These "TDL cycles" are the
basic unit of a TDL measurement. A typical output file from a TDL being used for
carbon isotope discrimination measurements therefore includes the following
columns (at minimum): time stamp, valve number, raw ^12^C concentration, and raw
^13^C concentration.

To analyze such a data set, it is necessary to identify the individual TDL
cycles within the entire set and use the calibration valves to determine
accurate gas concentrations in the other valves. Additional operations may also
be required, such as removing cycles that seem to be unreliable or smoothing the
data before calibration to decrease noise. In this vignette, we will
demonstrate several tools from the `PhotoGEA` package that can help to
streamline these operations in the R environment.

# Loading Packages

As always, the first step is to load the packages we will be using. In addition
to `PhotoGEA`, we will also use the `lattice` package for generating plots.

```{r setup}
# Load required packages
library(PhotoGEA)
library(lattice)
```

If the `lattice` package is not installed on your R setup, you can install it by
typing `install.packages('lattice')`.

# Loading TDL Data

The `PhotoGEA` package includes two files representing TDL data that was
recorded with a Campbell Scientific CR3000 data logger on two consecutive days.
Although these two files are based on real data, noise was added to it since it
is unpublished, so these files should only be used as examples.

The files will be stored on your computer somewhere in your R package
installation directory, and full paths to these files can be obtained with
`system.file`:

```{r tdl_file_names}
# Define a vector of paths to the files we wish to load
file_paths <- c(
  system.file('extdata', 'tdl_sampling_1.dat', package = 'PhotoGEA'),
  system.file('extdata', 'tdl_sampling_2.dat', package = 'PhotoGEA')
)
```

To actually read the data in the files and store them in R objects, we will use
the `read_tdl_file` function from `PhotoGEA`. Since there are multiple files
to read, we will call this function once for each file using `lapply`:

```{r loading_tdl_data}
# Load each file, storing the result in a list
tdl_exdf_list <- lapply(file_paths, function(fpath) {
  read_tdl_file(fpath, 1, 2, 3, 5, 'TIMESTAMP')
})
```

The result from this command is an R list of "extended data frames" (abbreviated
as `exdf` objects). The `exdf` class is a special data structure defined by the
`PhotoGEA` package. In many ways, an `exdf` object is equivalent to a data
frame, with the major difference being that an `exdf` object includes the units
of each column. For more information, type `?exdf` in the R terminal to access
the built-in help menu entry.

Generally, it is more convenient to work with a single `exdf` object rather than
a list of them, so our next step will be to combine the objects in the list.
This action can be accomplished using the `rbind` function, which combines
table-like objects by their rows; in other words, it stacks two or more tables
vertically. This action only makes sense if the tables have the same columns, so
before we combine the `exdf` objects, we should make sure this is the case.

The `PhotoGEA` package includes a function called `identify_common_columns` that
can be used to get the names of all columns that are present in all of the TDL
files. Then, we can extract just those columns, and then combine the `exdf`
objects into a single one.

```{r combining_tdl_data}
# Get the names of all columns that are present in all of the TDL files
columns_to_keep <- do.call(identify_common_columns, tdl_exdf_list)

# Extract just these columns
tdl_exdf_list <- lapply(tdl_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
tdl_files <- do.call(rbind, tdl_exdf_list)
```

Now we have a single R object called `tdl_files` that includes all the data
from several TDL data files.

# Identifying TDL Cycles

The first step when analyzing TDL data is to identify the "TDL cycles" that are
present within it. To do that, we need to know the following information:

* The name of the column in the TDL data that indicates which valve is being
  measured.
* The valve number that marks the beginning of a new cycle.
* How many valves are measured in each cycle.
* The name of the column in the TDL data that indicates the time at which each
  measurement was made.
* The amount of time it should take to cycle through the valves.

This information is usually known beforehand from the settings that were
specified  when operating the TDL; alternatively, if it isn't known already, it
can often be determined by taking a look at the data via `View(tdl_files)`. Once
this information is obtained, the `identify_tdl_cycles` function from the
`PhotoGEA` package can be used to automatically assign a number to each
measurement cycle:

```{r identifying_cycles}
# Assign numbers to all full cycles in the data set
tdl_files <- identify_tdl_cycles(
  tdl_files,
  valve_column_name = 'valve_number',
  cycle_start_valve = 20,
  expected_cycle_length_minutes = 2.7,
  expected_cycle_num_pts = 9,
  timestamp_colname = 'TIMESTAMP'
)
```

Now the `tdl_files` object has two new columns: one is called `cycle_num` and
specifies each TDL cycle, and the other is called `elapsed_time` and indicates
the elapsed time at the start of each cycle.

# Processing TDL Cycles

Now that the individual cycles have been identified, the next step is to use the
reference valves to calibrate the TDL readings. In general, this procedure will
depend strongly on the individual TDL system, since each one may have different
types of reference tanks. The data used in this vignette was measured using the
TDL in Carl Bernacchi's lab in the Edward R. Madigan Laboratory at the
University of Illinois, Urbana-Champaign. It includes five reference tanks that
can be broken into three types:

* One is a certified tank whose total CO~2~ concentration and carbon isotope
  ratio are supplied from NOAA.
* One is a nitrogen tank that should have no carbon in it.
* The final type of reference is another CO~2~ tank whose isotype ratio was
  measured at UIUC using a different method; this tank is mixed with nitrogen in
  three different ratios to provide a range of carbon concentrations.

The full procedure for using these references to calibrate the TDL signal is
somewhat complicated. Fortunately, it is easy to determine and apply the
calibrations for a single TDL cycle using the `process_tdl_cycle_erml` function
from `PhotoGEA`. To apply this function to each TDL cycle in a larger data set
and then consolidate the results, we can use it in conjunction with `by` and
`consolidate`, which are also part of `PhotoGEA`. (For more information about
these functions, see the built-in help menu entries by typing
`?process_tdl_cycle_erml`, `?by.exdf`, or `?consolidate`.) As with the previous
function, many of the inputs here (such as `noaa_cylinder_isotope_ratio`) must
be obtained from whoever is managing the TDL system being used:

```{r}
# Use the data from the calibration valves to determine calibrated carbon
# concentrations from all valves in each TDL cycle
processed_tdl <- consolidate(by(
  tdl_files,
  tdl_files[, 'cycle_num'],
  process_tdl_cycle_erml,
  valve_column_name = 'valve_number',
  noaa_valve = 2,
  calibration_0_valve = 20,
  calibration_1_valve = 21,
  calibration_2_valve = 23,
  calibration_3_valve = 26,
  raw_12c_colname = 'Conc12C_Avg',
  raw_13c_colname = 'Conc13C_Avg',
  noaa_cylinder_co2_concentration = 294.996,
  noaa_cylinder_isotope_ratio = -8.40,
  calibration_isotope_ratio = -11.505,
  f_other = 0.00474,
  R_VPDB = 0.0111797
))
```

The output from this function -- `processed_tdl` -- is a list of several `exdf`
objects that include calibrated TDL readings from each valve and information
about the calibration parameters that were determined while processing the data.

## Checking the Calibrated Data

Typically it is a good idea to check over the raw data, calibrated data, and
calibration constants for any issues that might affect the reliability of the
results. Since there are many valves involved and the calibration involves
several different steps (see `?process_tdl_cycle_erml`), there are also many
possible things to check. A full check of the data might require creating and
viewing many different plots.

Here we will first take a look at the raw and calibrated values of ^12^C from
each valve. This information is stored in the `tdl_data` element of
`processed_tdl`. The `lattice` library makes it simple to include each valve as
its own separate subplot of a figure, with the raw and calibrated values as
separate traces. Our goal is simply to keep an eye out for anythign that looks
unusual in the data.

```{r}
# Plot the 12C concentration data from all the valves
xyplot(
  Conc12C_Avg + calibrated_12c ~ cycle_num | factor(valve_number),
  data = processed_tdl$tdl_data$main_data,
  type = 'p',
  pch = 20,
  auto = TRUE,
  grid = TRUE,
  xlab = paste0('Elapsed time (', processed_tdl$tdl_data$units$cycle_num, ')'),
  ylab = paste0('12C concentration (', processed_tdl$tdl_data$units$Conc12C_Avg, ')'),
  main = '12C concentrations'
)
```

In this figure, it is apparent that valves 23 and 26 have large spikes, which is
surprising since these are calibration valves; the gas flowing to the TDL from
these valves is expected to have a fairly constant carbon concentration. It is
possible to view just one valve at a time for more detailed analysis. Here we
will take a look at the ^12^C concentration from valve 26, which exhibits a
spike.

```{r plot_single_valve, echo = FALSE, eval = FALSE}
# Get the processed data from the single valve
single_valve <-
  processed_tdl$tdl_data[processed_tdl$tdl_data[, 'valve_number'] == valve_to_plot, , TRUE]

# Plot the 12C concentration data from the single valve
xyplot(
  Conc12C_Avg + calibrated_12c ~ cycle_num,
  data = single_valve$main_data,
  type = 'b',
  pch = 20,
  auto = TRUE,
  grid = TRUE,
  xlab = 'TDL cycle',
  ylab = paste0('12C concentration (', single_valve$units$Conc12C_Avg, ')'),
  main = paste('12C concentrations from valve', valve_to_plot)
)
```

```{r}
# Specify a single valve to plot
valve_to_plot <- 26

<<plot_single_valve>>
```

In this graph, we can see that the calibrated ^12^C concentration has a large
deviation from the raw values near the large spike, although the calibrated and
raw values are generally very similar in the rest of the data. In fact, we can
see that in valve 12 (one of the valves whose composition we are trying to
determine), there is also a large deviation between the raw and calibrated
^12^C signals near the spike.

```{r, echo = FALSE}
valve_to_plot <- 12

<<plot_single_valve>>
```

Finally, we can also see traces of this anomaly in the calibration constants
determined during the TDL cycle processing. In this plot, the ^12^C gain factor
is nearly constant at all times with the exception of the cycles where there is
also a spike in the ^12^C signal in valve 26.

```{r}
# Plot the 12C gain factor determined during the calibration process
xyplot(
  gain_12CO2 ~ cycle_num,
  data = processed_tdl$calibration_12CO2$main_data,
  type = 'b',
  pch = 16,
  grid = TRUE,
  xlab = 'TDL cycle',
  ylab = paste0('12C gain factor (', processed_tdl$calibration_12CO2$units$gain_12CO2, ')')
)
```

# Processing TDL Cycles (With Removal and Smoothing)

In [Processing TDL Cycles], we processed the full contents of the raw TDL file.
However, it is typical to remove unreliable cycles and to apply smoothing
functions to some valves before processing the data. In this section, we
demonstrate how to do this.

## Removing Problematic Cycles

In [Checking the Calibrated Data], we located a spike in the measured ^12^C
concentration from valve 26, which is one of the calibration values. Since the
presence of this spike indicates that something unexpected happened to the TDL
system, and there is evidence that the calibration may be compromised during the
spike, it is probably a good idea to remove the TDL cycles where the ^12^C
signal from valve 26 suddenly becomes large.

Fortunately, this is simple to do in R. The following illustrates one way to
"clean" the data by removing the unreliable TDL cycles; here, we actually remove
two regions of the data that appear to be suspicious:

```{r removing_cycles}
# Define a vector of cycle numbers that should be removed
tdl_cycles_to_remove <- c(
  145:150, # The big spike in valve 26
  103:107  # A smaller spike in valve 26
)

# Remove them
tdl_files_clean <-
  tdl_files[!tdl_files[, 'cycle_num'] %in% tdl_cycles_to_remove, , TRUE]
```

## Smoothing the TDL Data

Since TDL signals are susceptible to several types of noise and the composition
of gas from the calibration tanks is expected to be relatively constant, it is
safe to assume that any high-frequency variations in the calibration valve
signals are due to noise. For this reason, it may be helpful to smooth the data
from the calibration valves before processing the TDL cycles. This may reduce
noise in the calibrated values from the "unknown" valves.

There are many possible strategies for smoothing the time series of values
measured from a valve, and the `PhotoGEA` package does not attempt to provide
its own smoothing algorithms. Instead, `PhotoGEA` provides a general function
called `smooth_tdl_data` that can apply a smoothing function to the data from
one valve. To use this function, a user must specify their own smoothing method.
(For more information, see the help menu entry by typing `?smooth_tdl_data`.)
Here we will show how the spline smoothing function from the base R package can
be used to smooth the ^12^C and ^13^C signals from the calibration valves:

```{r smoothing_valves}
# Define a spline smoothing function
spline_smoothing_function <- function(Y, X) {
    ss <- smooth.spline(X, Y)
    return(ss$y)
}

# Specify the valves that should be smoothed
valves_to_smooth <- c(2, 20, 21, 23, 26)

# Initialize a smoothed copy of the data
tdl_files_smooth <- tdl_files_clean

# Apply the smoothing function to the 12C and 13C signal from each valve
for (valve in valves_to_smooth) {
  for (column in c('Conc12C_Avg', 'Conc13C_Avg')) {
    tdl_files_smooth <-
      smooth_tdl_data(tdl_files_smooth, column, 'valve_number', valve, spline_smoothing_function)
  }
}
```

## Checking the Data After Cleaning and Smoothing

Now that we have removed some problematic cycles and applied a spline smoothing
function to the calibration valves, we can process the TDL cycles again using
the same command from [Processing TDL Cycles]; the only difference here is that
we are operating on the smoothed data:

```{r processing_smooth_tdl_cycles}
# Use the data from the calibration valves to determine calibrated carbon
# concentrations from all valves in each TDL cycle
processed_smooth_tdl <- consolidate(by(
  tdl_files_smooth,
  tdl_files_smooth[, 'cycle_num'],
  process_tdl_cycle_erml,
  valve_column_name = 'valve_number',
  noaa_valve = 2,
  calibration_0_valve = 20,
  calibration_1_valve = 21,
  calibration_2_valve = 23,
  calibration_3_valve = 26,
  raw_12c_colname = 'Conc12C_Avg',
  raw_13c_colname = 'Conc13C_Avg',
  noaa_cylinder_co2_concentration = 294.996,
  noaa_cylinder_isotope_ratio = -8.40,
  calibration_isotope_ratio = -11.505,
  f_other = 0.00474,
  R_VPDB = 0.0111797
))
```

It is important to confirm that the smoothing procedure has produced reasonable
outputs. Here we create a plot comparing the raw and smoothed ^12^C signal from
all valves:

```{r compare_smoothing}
# Add columns indicating which which operations have been applied to each exdf
tdl_files[ , 'data_type'] <- '01 raw data'
tdl_files_clean[ , 'data_type'] <- '02 cleaned data'
tdl_files_smooth[ , 'data_type'] <- '03 cleaned smoothed data'

# Make a big exdf object with the raw and smoothed data
tdl_compare <- rbind(tdl_files, tdl_files_clean, tdl_files_smooth)

# Keep only the valves that were smoothed
tdl_compare <-
  tdl_compare[tdl_compare[ , 'valve_number'] %in% valves_to_smooth, , TRUE]

# Plot the uncalibrated 12C concentration data from all the valves
xyplot(
  Conc12C_Avg ~ cycle_num | factor(valve_number),
  group = data_type,
  data = tdl_compare$main_data,
  type = 'b',
  pch = 20,
  auto = TRUE,
  grid = TRUE,
  xlab = 'TDL cycle',
  ylab = paste0('Uncalibrated 12C concentration (', tdl_compare$tdl_data$units$Conc12C_Avg, ')'),
  main = 'Uncalibrated 12C concentrations'
)
```

Although it is clear that we have successfully removed the spikes from valve 26,
it may be helpful to display just that valve to take a closer look:

```{r compare_single_valve}
# Specify a single valve to plot
valve_to_compare <- 26

# Get the comparison data from the single valve
single_valve_compare <-
  tdl_compare[tdl_compare[, 'valve_number'] == valve_to_compare, , TRUE]

# Plot the 12C concentration data from the single valve
xyplot(
  Conc12C_Avg ~ cycle_num,
  group = data_type,
  data = single_valve_compare$main_data,
  type = 'b',
  pch = 20,
  auto = TRUE,
  grid = TRUE,
  xlab = 'TDL cycle',
  ylab = paste0('Uncalibrated 12C concentration (', single_valve_compare$units$Conc12C_Avg, ')'),
  main = paste('Uncalibrated 12C concentrations from valve', valve_to_compare)
)
```

In this graph we can see that the spike was indeed removed and that the smoothed
data does a good job of following the raw data for this valve. As another check,
we can also see how the smoothed data is altered after the calibration process.
Here we will look at valve 12, as in [Checking the Calibrated Data]:

```{r plot_single_valve_smooth}
# Specify a single valve to plot
valve_to_plot <- 12

# Get the processed data from the single valve
single_valve_smooth <- processed_smooth_tdl$tdl_data[
  processed_smooth_tdl$tdl_data[, 'valve_number'] == valve_to_plot, , TRUE]

# Plot the 12C concentration data from the single valve
xyplot(
  Conc12C_Avg + calibrated_12c ~ cycle_num,
  data = single_valve_smooth$main_data,
  type = 'b',
  pch = 20,
  auto = TRUE,
  grid = TRUE,
  xlab = 'TDL cycle',
  ylab = paste0('12C concentration (', single_valve_smooth$units$Conc12C_Avg, ')'),
  main = paste('12C concentrations from valve', valve_to_plot)
)
```

Here we see that after cleaning and smoothing the data, the calibrated data from
this "unknown" valve is only slightly different than its raw data across all the
cycles. Finally, we can also plot the ^12^C gain factor as in
[Checking the Calibrated Data]:

```{r plot_12c_gain}
# Plot the 12C gain factor determined during the calibration process
xyplot(
  gain_12CO2 ~ cycle_num,
  data = processed_smooth_tdl$calibration_12CO2$main_data,
  type = 'b',
  pch = 16,
  grid = TRUE,
  xlab = 'TDL cycle',
  ylab = paste0('12C gain factor (', processed_smooth_tdl$calibration_12CO2$units$gain_12CO2, ')')
)
```

Again, there are no sudden spikes remaining in the gain factor, indicating that
we did a good job of cleaning and smoothing the data.

# Next steps

Having cleaned, smoothed, and visually checked the data, we can now be confident
that the calibrated ^12^C and ^13^C values are as accurate as they can be. The
next step in the analysis would be to interpret the data from the valves whose
composition is unknown and to be determined. This will depend on your particular
application, so we cannot give general advice here!

# Customizing Your Script

Note that most of the commands in this vignette have been written in a general
way so they can be used as the basis for your own TDL analysis script (see
[Commands From This Document]). In order to use them in your own script, some or
all of the following changes may be required. There may also be others not
specifically mentioned here.

## Input Files

The file paths specified in `file_paths` will need to be modified so they point
to your TDL data files. In your own script, you may want to consider using the
`choose_input_tdl_files` function from `PhotoGEA`; this function will create a
pop-up browser window where you can interactively select a set of `.dat` files.
Sometimes this is more convenient than writing out full file paths. For example,
you could replace the previous definition of `file_paths` with this one:

```{r choose_tdl_files, eval = FALSE}
# Interactively define a vector of paths to the files we wish to load
file_paths <- choose_input_tdl_files()
```

Unfortunately, `choose_input_tdl_files` is only available in interactive R
sessions running on Microsoft Windows.

## Plots for Checking

To keep this vignette shorter, we skipped out on many possible plots that you
may want to view. In addition to the ones generated here, it may also be helpful
to make the following plots:

* Plots of ^13C^ signals in addition to plots of ^12^C.
* Individual plots of the raw, cleaned, and smoothed data for each valve.
* Plots of the ^13C^ calibration constants.
* Plots of the ^13^C calibration fits.

These plots can all be created by making small modifications to the commands in
this document.

## Cycles to Remove

When beginning to analyze a new data set, it is advisable to not remove any TDL
cycles; this can be accomplished by defining `tdl_cycles_to_remove` with the
following command in place of the one used above:

```{r, eval = FALSE}
tdl_cycles_to_remove <- c()
```

After viewing all of the data, unreliable cycles that should be removed can be
identified.

## Smoothing Function

When beginning to analyze a new data set, it is advisable to disable smoothing.
This can be done by definining a "null" smoothing function; in other words, a
function that does not actually perform any smoothing:

```{r null_smoothing, eval = FALSE}
# Define a null smoothing function
null_smoothing_function <- function(Y, X) {return(Y)}
```

Then, this function should be passed to `smooth_tdl_data` in place of
`spline_smoothing_function`.

# Commands From This Document

The following code chunk includes all the central commands used throughout this
document. They are compiled here to make them easy to copy/paste into a text
file to initialize your own script.

```{r, eval = FALSE}
<<setup>>

<<tdl_file_names>>

<<loading_tdl_data>>

<<combining_tdl_data>>

<<identifying_cycles>>

<<removing_cycles>>

<<smoothing_valves>>

<<processing_smooth_tdl_cycles>>

<<compare_smoothing>>

<<compare_single_valve>>

<<plot_single_valve_smooth>>

<<plot_12c_gain>>
```

# References
