---
title: "Analyzing Ball-Berry Data"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Analyzing Ball-Berry Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

In this vignette, we will give an example showing how to analyze Ball-Berry data
using the `PhotoGEA` package. The commands in this vignette can be used to
initialize your own script, as described in [Customizing Your Script].

# Background

## The Model

The Ball-Berry model is a simple way to describe the response of a leaf's
stomata to its assimilation rate and local environmental conditions.
Specifically, it predicts stomatal conductance to water vapor using the
following equation:

$$ g_{sw} = b_0 + b_1 * A_n * h_s / C_s $$

where $g_{sw}$ is the stomatal conductance, $A_n$ is the net assimilation rate,
$h_s$ is the relative humidity at the leaf surface, and $C_s$ is the CO~2~
concentration at the leaf surface. The term $A_n * h_s / C_s$ is commonly
referred to as the Ball-Berry index, and the model can be described as
predicting a linear relationship between stomatal conductance and the Ball-Berry
index. The intercept ($b_0$) and slope ($b_1$) of the linear relationship are
referred to as the Ball-Berry parameters, and they describe the leaf's stomatal
response.

Although this model is certainly an oversimplification, it does encode some
important stomatal responses. For example, when humidity is low, the stomata
of a real plant tend to close, reducing stomatal conductance. Likewise, if the
CO~2~ concentration around the leaf is depleted, the stomata of a real plant
tend to open, increasing somatal conductance. For more information about this
model and some possible alternatives, see the following papers:

* @ball_model_1987
* @tardieu_integration_1993
* @leuning_critical_1995
* @dewar_ballberryleuning_2002

## The Data

Ball-Berry parameters are typically determined by measuring a Ball-Berry curve,
where one or more of the environmental factors that influence the Ball-Berry
index is systematically varied across a range of values. For each value, care is
taken that $A_n$ and $g_{sw}$ have reached their steady-state values, and then
$A_n$, $g_{sw}$, and other key pieces of information are recorded. Finally, a
linear fit of the experimentally observed $g_{sw}$ as a function of the
Ball-Berry index is performed to extract values for the Ball-Berry intercept and
slope.

Such photosynthetic gas exchange measurements are commonly performed using a
Licor portable photosynthesis system such as the Li-6400 or Li-6800. These
machines record values of $A_n$, $g_{sw}$, and many other important quantities.
They produce two types of output files: plain-text and Microsoft Excel. It is
often more convenient to work with the Excel files since the entries can be
easily modified (for example, to remove an extraneous row or add a new column).
On the other hand, it can be more difficult to access the file contents using
other pieces of software such as R. However, the `PhotoGEA` package reduces this
barrier by including tools for reading Licor Excel files in R, which will be
demonstrated in the following section.

# Loading Packages

As always, the first step is to load the packages we will be using. In addition
to `PhotoGEA`, we will also use the `lattice` package for generating plots.

```{r setup}
# Load required packages
library(PhotoGEA)
library(lattice)
```

If the `lattice` package is not installed on your R setup, you can install it by
typing `install.packages('lattice')`.

# Loading Licor Data

The `PhotoGEA` package includes two files representing Ball-Berry data measured
using two Li-6800 instruments. The data is stored in Microsoft Excel files, and
includes curves measured from two different crop species (tobacco and soybean)
and several different plots of each. Each curve is a seven-point light response
curve; in other words, the incident light intensity was varied to change $A_n$
and hence the Ball-Berry index. Although these two files are based on real data,
noise was added to it since it is unpublished, so these files should only be
used as examples.

The files will be stored on your computer somewhere in your R package
installation directory, and full paths to these files can be obtained with
`system.file`:

```{r licor_file_names}
# Define a vector of paths to the files we wish to load
file_paths <- c(
  system.file('extdata', 'ball_berry_1.xlsx', package = 'PhotoGEA'),
  system.file('extdata', 'ball_berry_2.xlsx', package = 'PhotoGEA')
)
```

To actually read the data in the files and store them in R objects, we will use
the `read_licor_file` function from `PhotoGEA`. Since there are multiple files
to read, we will call this function once for each file using `lapply`:

```{r loading_licor_data}
# Load each file, storing the result in a list
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_licor_file(fpath, c(3, 5, 7, 9, 11, 13), 14, 15, 16, 17, 'time')
})
```

The result from this command is an R list of "extended data frames" (abbreviated
as `exdf` objects). The `exdf` class is a special data structure defined by the
`PhotoGEA` package. In many ways, an `exdf` object is equivalent to a data
frame, with the major difference being that an `exdf` object includes the units
of each column. For more information, type `?exdf` in the R terminal to access
the built-in help menu entry.

Generally, it is more convenient to work with a single `exdf` object rather than
a list of them, so our next step will be to combine the objects in the list.
This action can be accomplished using the `rbind` function, which combines
table-like objects by their rows; in other words, it stacks two or more tables
vertically. This action only makes sense if the tables have the same columns, so
before we combine the `exdf` objects, we should make sure this is the case.

The `PhotoGEA` package includes a function called `identify_common_columns` that
can be used to get the names of all columns that are present in all of the Licor
files. Then, we can extract just those columns, and then combine the `exdf`
objects into a single one.

```{r combining_licor_data}
# Get the names of all columns that are present in all of the Licor files
columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
licor_data <- do.call(rbind, licor_exdf_list)
```

Now we have a single R object called `licor_data` that includes all the data
from several Licor Excel files.

# Checking Licor Data

Before extracting Ball-Berry parameters, it is a good idea to do some basic
checks of the data to ensure it is organized properly and that it was measured
properly.

## Basic Checks

First, we should make sure there is a column in the data whose value uniquely
identifies each curve. In this particular data set, several "user constants"
were defined while making the measurements that help to identify each curve:
`instrument`, `species`, and `plot`. However, neither of these columns alone are
sufficient to uniquely identify each curve. We can solve this issue by creating
a new column that combines the values from each of these:

```{r adding_id_column}
# Create a new identifier column formatted like `instrument - species - plot`
licor_data[ , 'curve_identifier'] <-
  paste(licor_data[ , 'instrument'], '-', licor_data[ , 'species'], '-', licor_data[ , 'plot'])
```

The next step is to make sure that this column correctly identifies each
response curve. To do this, we can use the `check_licor_data` function from
`PhotoGEA`. Here we will supply the name of a column that should uniquely
identify each response curve (`curve_identifier`), the expected number of points
in each curve (`7`), the name of a "driving" column that should follow the same
sequence in each curve (`Qin`), and tolerance for the values of the driving
column (`1.0 mol / m^2 / s`). If the data passes the checks, this function will
have no output and will not produce any messages. (For more information, see the
built-in help menu entry by typing `?check_licor_data`.)

```{r check_response_curve}
# Make sure the data meets basic requirements
check_licor_data(licor_data, 'curve_identifier', 7, 'Qin', 1.0)
```

However, if `check_licor_data` detects an issue, it will print a
helpful message to the R terminal. For example, if we had specified the wrong
number of points, the wrong identifier column, or an unrealistic tolerance, we
would get error messages:

```{r check_response_curve_error, error = TRUE}
check_licor_data(licor_data, 'curve_identifier', 8)

check_licor_data(licor_data, 'species', 7)

check_licor_data(licor_data, 'curve_identifier', 7, 'Qin', 0.1)
```

## Qualitative Checks

Since the data has passed the basic checks, we can now take a look at it to make
sure it seems reasonable. Sometimes a Licor will override the temperature or
humidity controls while making measurements; in this case, conditions inside the
measurement chamber may not be stable, and these curves may not accurately
reflect the plant's true steady-state stomatal response. We can check for these
types of issues by making plots. In this situation, the `lattice` library makes
it simple to include each curve as its own separate subplot of a figure. In the
following sections, we will generate several different plots to check each curve
for quality.

### Humidity Control

```{r plot_humidity}
# Make a plot to check humidity control
xyplot(
  RHcham + `Humidifier_%` + `Desiccant_%` ~ Qin | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(0, 100),
  xlim = c(0, 2200),
  xlab = paste0('Incident PPFD (', licor_data$units$Qin, ')')
)
```

Here, `Humidifier_%` and `Desiccant_%` represent the flow from the humidifier
and desiccant columns, where a value of 0 indicates that the valve to the column
is fully closed and a value of 100 indicates that the valve to the column is
fully opened. `RHcham` represents the relative humidity inside the chamber as a
percentage (in other words, as a value between 0 and 100). `Qin` is the incident
photosynthetically active flux density (PPFD). When these curves were measured,
a chamber humidity setpoint was specified. So, when looking at this plot, we
should check that the relative humidity is fairly constant during each curve.
Typically, this should be accompanied by relatively smooth changes in the valve
percentages as they accomodate changes in ambient humidity and leaf
photosynthesis. In this plot, all the data looks good.

### Temperature Control

```{r plot_temperature}
# Make a plot to check temperature control
xyplot(
  TleafCnd + Txchg ~ Qin | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(25, 40),
  xlim = c(0, 2200),
  xlab = paste0('Incident PPFD (', licor_data$units$Qin, ')'),
  ylab = paste0('Temperature (', licor_data$units$TleafCnd, ')')
)
```

Here, `TleafCnd` is the leaf temperature measured using a thermocouple, and
`Txchg` is the temperature of the heat exhanger that is used to control the air
temperature in the measurement instrument. When these curves were measured, an
exchanger setpoint was specified. So, when looking at this plot, we should check
that `Txchg` is constant during each curve and that the leaf temperature does
not vary in an erratic way. In this plot, all the data looks good.

### CO~2~ Control

```{r plot_co2}
# Make a plot to check CO2 control
xyplot(
  CO2_s + CO2_r ~ Qin | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(360, 480),
  xlim = c(0, 2200),
  xlab = paste0('Incident PPFD (', licor_data$units$Qin, ')'),
  ylab = paste0('CO2 concentration (', licor_data$units$CO2_r, ')')
)
```

Here, `CO2_s` is the CO~2~ concentration in the sample cell and `CO2_r` is the
CO~2~ concentration in the reference cell. When these curves were measured, a
sample cell CO~2~ concentration setpoint was supplied. So, when looking at this
plot, we should check that `CO2_s` is constant during each curve. Here, it looks
like the `ripe5` instrument was not controlling CO~2~ as expected during several
of its curves; `CO2_s` is not constant for `ripe5 - soybean - 1`,
`ripe5 - soybean - 5c`, and `ripe5 - soybean - 5`. However, `CO2_r` is
relatively constant during these curves, and the changes in `CO2_s` are smooth,
so it is reasonable to expect that the measurements represent true steady-state
values. Considering this, all of these curves are acceptable based on the CO~2~
plots.

### Stability

```{r plot_stability}
xyplot(
  `A:OK` + `gsw:OK` + Stable ~ Qin | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlim = c(0, 2200),
  xlab = paste0('Incident PPFD (', licor_data$units$Qin, ')')
)
```

When measuring response curves with a Licor, it is possible to specify stability
criteria for each point in addition to minimum and maximum wait times. In other
words, once the set point for the driving variable is changed, the machine waits
until the stability criteria are met; there is a minimum waiting period, and
also a maximum to prevent the machine from waiting for too long. These stability
criteria are especially important for Ball-Berry curves, since the stomata may
take a long time to reach steady state.

When these curves were measured, stability criteria were supplied for the net
assimilation rate `A` and the stomatal conductance `gsw`. The stability status
for each was stored in the log file because the appropriate logging option for
stability was set. Now, for each point, it is possible to check whether
stability was achieved or whether the point was logged because the maximum
waiting period had been met. If the maximum waiting period is reached and the
plant has still not stabilized, the data point may be unreliable, so it is very
important to check this information.

In the plot, `A:OK` indicates whether `A` was stable (0 for no, 1 for yes),
`gsw:OK` indicates whether `gsw` was stable (0 for no, 1 for yes), and
`Stable` indicates the total number of stability conditions that were met. So,
we are looking for points where `Stable` is 2. Otherwise, we can check the other
traces to see whether `A` or `gsw` was unstable. Here it looks like many of the
high light points were not stable, and it may be a good idea to remove them
before proceeding with the Ball-Berry fitting.

### Light-Response Curves

```{r plot_assimilation}
# Make a plot to check light-response curves
xyplot(
  A ~ Qin | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(-5, 50),
  xlim = c(0, 2200),
  xlab = paste0('Incident PPFD (', licor_data$units$Qin, ')'),
  ylab = paste0('Net assimilation (', licor_data$units$A, ')')
)
```

Here we are simply looking for a reasonable light-response curve shape, where
assimilation is low when light is low, assimilation has a roughly linear
response to initial increases in light intensity, and then finally reaches a
plateau. Any strong deviations from the expected shape may indicate that the
plant was stressed or otherwise behaving abnormally, and we might not want to
use such a curve for Ball-Berry analysis. In this plot, three of the curves do
not look like normal C~3~ light response curves: `ripe4 - soybean - 1a`,
`ripe4 - soybean - 1b`, and `ripe4 - soybean - 5`. These strange curves are
likely a byproduct of the noise that was intentionally added to the true
measured data (see [The Data]). Nevertheless, it may be a good idea to remove
them before proceeding with the Ball-Berry fitting.

## Cleaning the Licor Data

While checking over the plots in [Qualitative Checks], two issues were noticed:
(1) some points were logged before stability was achieved and (2) some of the
curves look abnormal. In this section, we will demonstrate how to remove the
unstable points and the weird curves.

The following command will keep only the points where `Stable` is exactly 2;
this condition means that all of the stability criteria were satisfied.
Sometimes, following this procedure, a curve will have very few stable points
remaining; it is usually a good idea to automatically exclude any curve with
fewer than three stable points.

```{r remove_unstable_points}
# Only keep points where stability was achieved
licor_data <- licor_data[licor_data[, 'Stable'] == 2, , TRUE]

# Remove any curves that have fewer than three remaining points
npts <- by(licor_data, licor_data[, 'curve_identifier'], nrow)
ids_to_keep <- names(npts[npts > 2])
licor_data <- licor_data[licor_data[, 'curve_identifier'] %in% ids_to_keep, , TRUE]
```

Since we have identified a few curves that may not be acceptable for Ball-Berry
fitting, we can remove them via the `remove_points` function from `PhotoGEA`:

```{r remove_curves}
# Define a list of curves to remove from the data set
curves_to_remove <- c(
  'ripe4 - soybean - 1a',
  'ripe4 - soybean - 1b',
  'ripe4 - soybean - 5'
)

# Remove them
licor_data <-
  remove_points(licor_data, list(curve_identifier = curves_to_remove))
```

Note that `remove_points` can also be used to exclude individual points instead
of entire curves; see its help entry by typing `?remove_points` for more
information.

# Fitting Licor Data

Now that we have checked the data quality, we are ready to perform the fitting.
As they are produced by the instruments, Licor data files do not include values
of the Ball-Berry index; in fact, they do not even include values of $h_s$ and
$C_s$ required to calculate the Ball-Berry index. However, the `PhotoGEA`
package includes two functions to help with these calculations:
`calculate_gas_properties` and `calculate_ball_berry_index`. Each of these
requires an `exdf` object containing Licor data, as well as the names of
several important columns in the data. The units for each column will be checked
in an attempt to avoid unit-related errors. More information about these
functions can be obtained from the built-in help system by typing
`?calculate_gas_properties` or `?calculate_ball_berry_index`. Here we will use
them sequentially to calculate values of the Ball-Berry index:

```{r calculate_bb_index}
# Calculate additional gas properties, including `RHleaf` and `Csurface`
licor_data <- calculate_gas_properties(
  licor_data,
  'A', 'Ca', 'DeltaPcham', 'E', 'gbw', 'gsw', 'H2O_s', 'Pa', 'TleafCnd'
)

# Calculate the Ball-Berry index
licor_data <- calculate_ball_berry_index(licor_data, 'A', 'RHleaf', 'Csurface')
```

Together, these functions add several new columns to `licor_data`, including one
called `bb_index`, which includes values of the Ball-Berry index. With this
information, we are now ready to perform the fitting procedure. For this
operation, we can use the `fit_ball_berry` function from the `PhotoGEA` package,
which fits a single Ball-Berry curve to extract the Ball-Berry parameters. To
apply this function to each curve in a larger data set and then consolidate the
results, we can use it in conjunction with `by` and `consolidate`, which are
also part of `PhotoGEA`. (For more information about these functions, see the
built-in help menu entries by typing `?fit_ball_berry`, `?by.exdf`, or
`?consolidate`.) Together, these functions will split apart the main data using
the curve identifier column we defined before ([Basic Checks]), make a linear
fit of $g_{sw}$ against the Ball-Berry index, and return the resulting
parameters and fits:

```{r fit_ball_berry}
# Fit a linear model to the Ball-Berry data
ball_berry_results <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_ball_berry,                   # The function to apply to each chunk of `licor_data`
  'gsw',                            # Additional argument required by `fit_ball_berry`
  'bb_index'                        # Additional argument required by `fit_ball_berry`
))
```

## Viewing the Fitted Curves

Having made the fits, it is now a good idea to visually check them, making sure
they look reasonable. As before, we can use `xyplot` from the `lattice` library
to quickly make plots of all the fits, which are stored in
`ball_berry_results$fits`, an `exdf` object:

```{r plot_fits}
# Plot the Ball-Berry fits
xyplot(
  gsw + gsw_fit ~ bb_index | curve_identifier,
  data = ball_berry_results$fits$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste0('Ball-Berry index (', ball_berry_results$fits$units$bb_index, ')'),
  ylab = paste0('Stomatal conductance to H2O (', ball_berry_results$fits$units$gsw, ')'),
  ylim = c(0, 0.8),
  xlim = c(0, 0.1)
)
```

All these fits look good!

## Viewing the Fitted Parameter Values

We can also take a look at the fitted Ball-Berry parameter values, which are
stored in `ball_berry_results$parameters`, another `exdf` object. This object
includes many columns but we only care about a few of them. We can view them as
follows:

```{r ball_berry_parameters}
# View the Ball-Berry parameters
columns_for_viewing <-
  c('instrument', 'species', 'plot', 'bb_intercept', 'bb_slope', 'r_squared')

ball_berry_parameters <-
  ball_berry_results$parameters[ , columns_for_viewing, TRUE]

print(ball_berry_parameters)
```

## Extracting Average Values

Finally, we can extract average values of the Ball-Berry parameters for each
species using the `basic_stats` function from `PhotoGEA`:

```{r calculate_stats}
# Compute the average and standard error for the Ball-Berry slope and intercept
# for each species
ball_berry_averages <- basic_stats(
  ball_berry_results$parameters[ , c('species', 'K', 'bb_intercept', 'bb_slope'), TRUE],
  'species'
)

# View the averages and errors
print(ball_berry_averages)
```

# Customizing Your Script

Note that most of the commands in this vignette have been written in a general
way so they can be used as the basis for your own Ball-Berry analysis script
(see [Commands From This Document]). In order to use them in your own script,
some or all of the following changes may be required. There may also be others
not specifically mentioned here.

## Input Files

The file paths specified in `file_paths` will need to be modified so they point
to your Licor files. In your own script, you may want to consider using the
`choose_input_licor_files` function from `PhotoGEA`; this function will create a
pop-up browser window where you can interactively select a set of Excel files.
Sometimes this is more convenient than writing out full file paths. For example,
you could replace the previous definition of `file_paths` with this one:

```{r choose_licor_files, eval = FALSE}
# Interactively define a vector of paths to the files we wish to load
file_paths <- choose_input_licor_files()
```

Unfortunately, `choose_input_licor_files` is only available in interactive R
sessions running on Microsoft Windows.

## Curve Identifier

Depending on which user constants are defined in your Licor Excel files, you may
need to modify the definition of the `curve_identifier` column.

## Excluded Curves

Depending on the qualitative data checks, you may need to change the definition
of the `curves_to_remove` vector.

## Averages and Standard Errors

Depending on how your data is organized, you may want to change the column used
to divide the data when calculating averages and standard errors.

## Plots

You may need to change the axis limits in some or all of the plots.
Alternatively, you can remove them, allowing `xyplot` to automatically choose
them for you.

## Saving Results

You may want to use `write.csv` to save some or all of the fitting results as
`csv` files. For example, the following commands will allow you to interactively
choose output filenames for the resulting `csv` files:

```{r saving_csv_files, eval = FALSE}
write.csv(ball_berry_results$fits, file.choose(), row.names = FALSE)
write.csv(ball_berry_results$parameters, file.choose(), row.names = FALSE)
write.csv(ball_berry_averages, file.choose(), row.names = FALSE)
```

# Commands From This Document

The following code chunk includes all the central commands used throughout this
document. They are compiled here to make them easy to copy/paste into a text
file to initialize your own script. Annotation has also been added to clearly
indicate the four steps involved in data analysis, as described in the
_Developing a Data Analysis Pipeline_ vignette.

```{r, eval = FALSE}
###
### PRELIMINARIES:
### Loading packages, defining constants, creating helping functions, etc.
###

<<setup>>

###
### TRANSLATION:
### Creating convenient R objects from raw data files
###

<<licor_file_names>>

<<loading_licor_data>>

<<combining_licor_data>>

###
### VALIDATION:
### Organizing the data, checking its consistency and quality, cleaning it
###

<<adding_id_column>>

<<check_response_curve>>

<<plot_humidity>>

<<plot_temperature>>

<<plot_co2>>

<<plot_assimilation>>

<<remove_unstable_points>>

<<remove_curves>>

###
### PROCESSING:
### Extracting new pieces of information from the data
###

<<calculate_bb_index>>

<<fit_ball_berry>>

<<plot_fits>>

<<ball_berry_parameters>>

###
### SYNTHESIS:
### Using plots and statistics to help draw conclusions from the data
###

<<calculate_stats>>
```

# References
