---
title: "Analyzing C4 A-Ci Curves"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Analyzing C4 A-Ci Curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

In this vignette, we will give an example showing how to analyze C~4~ A-Ci data
using the `PhotoGEA` package. The commands in this vignette can be used to
initialize your own script, as described in [Customizing Your Script].

# Background

## Understanding C4 A-Ci Curves

An _A-Ci curve_ (or _CO~2~ response curve_) is a particular type of gas exchange
measurement where a living leaf is exposed to varying concentrations of CO~2~.
For each CO~2~ concentration in the sequence, the net assimilation rate $A_n$,
stomatal conductance to H~2~O ($g_{sw}$), and other important quantities are
measured and recorded. Typically, other environmental variables such as
temperature, humidity, and incident photosynthetically-active photon flux
density are held constant during the measurement sequence so that changes in
photosynthesis can be attributed to CO~2~ alone.

Because of their different cell structures and biochemical pathways, C~3~ and
C~4~ plants have very different responses to CO~2~. Here, we will only be
discussing C~4~ plants.

The full C~4~ photosynthetic pathway is quite complicated, consisting of at
least two hundred individual reactions, each of which may have an impact on a
measured A-Ci curve. However, simplified models for photosynthesis are available
and are much easier to understand and work with. These models tend to be based
around PEP carboxylase and rubisco kinetics, describing how the net assimilation
rate responds to the partial pressure of CO~2~ in the mesophyll or bundle sheath
cells. The most widely-used model is described in in
_Biochemical Models of Leaf Photosynthesis_ [@caemmerer_biochemical_2000].

This model provides a framework for understanding the changes in $A_n$ that
occur as a C~4~ plant is exposed to successively higher concentrations of CO~2~.
Overall, the photosynthetic response to CO~2~ under high light conditions can be
divided into two separate ranges:

- For low levels of CO~2~, CO~2~ assimilation is primarily limited by PEP
  carboxylation in the mesophyll cells.
- For high levels of CO~2~, CO~2~ assimilation is primarily limited by rubisco
  activity in the bundle sheath cells (or possibly by the rate of PEP
  carboxylase regeneration).

More specifically, the model provides equations that calculate the net
assimilation rate $A_n$ from the maximumum rate of PEP carboxylation
($V_{p,max}$), the maximum rate of PEP carboxylase regeneration ($V_{p,r}$), the
maximum rate of rubisco activity ($V_{c,max}$), the total rate of mitochondrial
respiration in the mesophyll and bundle sheath cells ($R_d$), the partial
pressure of CO~2~ in the mesophyll ($P_{cm}$), and several other parameters. The
full equations are quite complicated, but can be approximated by a simpler
version where $A_n$ is given by the minimum of three separate enzyme-limited
assimilation rates:

- The PEP-carboxylation-limited rate ($A_{pc}$).
- The rubisco-limited rate ($A_r$).
- The PEP-regeneration-limited rate ($A_{pr}$).

Using the equations that compose this model, it is possible to calculate plots
like the one below, which illustrates the three different limiting rates and the
overall net assimilation rate calculated using the full equations. The two
distinct ranges can be identified as the initial part of the curve (where there
is a roughly linear increase in $A_n$ with $P_{cm}$) and the later part (where
$A_n$ is nearly flat with very little dependence on $P_{cm}$).

```{r, echo = FALSE}
library(PhotoGEA)
library(lattice)

npts <- 101

Tleaf = 30

inputs <- exdf(data.frame(
  PCm = seq(0, 500, length.out = npts),
  Tleaf = Tleaf
))

inputs <- document_variables(
  inputs,
  c('', 'PCm',   'microbar'),
  c('', 'Tleaf', 'degrees C')
)

inputs <- calculate_arrhenius(inputs, c4_arrhenius_von_caemmerer, 'Tleaf')

alpha_psii <- 0
gbs <- 0.003
Rd_at_25 <- 1
Rm_frac <- 0.5
Vcmax_at_25 <- 30
Vpmax_at_25 <- 150
Vpr <- 80

assim <- calculate_c4_assimilation(
  inputs,
  alpha_psii, gbs, Rd_at_25, Rm_frac, Vcmax_at_25, Vpmax_at_25, Vpr
)

c4_col <- rev(multi_curve_colors()[1:4])

xyplot(
  Apr + Ar + Apc + An ~ PCm,
  data = cbind(inputs, assim)$main_data,
  type = 'l',
  grid = TRUE,
  auto = TRUE,
  xlab = paste('CO2 pressure in mesophyll [', inputs$units$PCm, ']'),
  ylab = paste('Assimilation rate [', assim$units$An, ']'),
  par.settings = list(
    superpose.line = list(col = c4_col),
    superpose.symbol = list(col = c4_col, pch = 16)
  )
)

```

(Note: this figure was generated using the `calculate_c4_assimilation` function
from the `PhotoGEA` package, and it represents the photosynethetic response of a
C~4~ leaf according to the model with
`Rd_at_25` = `r Rd_at_25` `r assim$units$Rd_at_25`,
`Vpmax_at_25` = `r Vpmax_at_25` `r assim$units$Vpmax_at_25`,
`Vcmax_at_25` = `r Vcmax_at_25` `r assim$units$Vcmax_at_25`,
`Vpr` = `r Vpr` `r assim$units$Vpr`, and a leaf temperature of
`r Tleaf` `r inputs$units$Tleaf`. Arrhenius temperature response parameters were
taken from @von_caemmerer_updating_2021.)

Thus, one of the most common reasons to measure an A-Ci curve is to interpret it
in the context of this model. In other words, by fitting the model's equations
to a measured curve, it is possible to estimate values for `Rd`, `Vpmax`,
`Vcmax`, and `Vpr`. See the documentation for `calculate_c4_assimilation` for
more information about these important quantities.

As is evident from the figure above, rubisco activity and PEP regeneration
limitations actually both produce a very similar end result, where $A_n$ becomes
essentially flat as $P_{cm}$ increases. For this reason, it is practically
impossible to distinguish between the two types of limiting rates when analyzing
a C~4~ A-Ci curve; if both limitations are considered simultaneously, it is not
usually possible for an optimization algorithm to determine a unique solution.
Because of this, it is common to set either `Vpr` or `Vcmax` to a high value
when fitting (so it never becomes a limiting factor), and only allow the other
to vary. In the rest of this document, we will be setting `Vpr = 800` and only
allowing `Vcmax` to vary when performing fits; this is the default behavior of
the `fit_c4_aci` function from the `PhotoGEA` package, which will be discussed
later.

## Practicalities

There are a few important practicalities to keep in mind when thinking about
CO~2~ response curves.

One point is that C4 photosynthesis models generally predict the response of
assimilation to the partial pressure of CO~2~ in the mesophyll ($P_{cm}$), but
gas exchange measurements can only determine the CO~2~ concentration in the
leaf's intercellular spaces ($C_i$). Thus, an extra step is required when
interpreting A-Ci curves. If the mesophyll conductance to CO~2~ ($g_{mc}$) and
the total pressure ($P$) are known, then it is possible to calculate values of
$P_{cm}$ from $A_n$, $C_i$, $P$, and $g_{mc}$. Otherwise, it is also possible to
assume an infinite mesophyll conductance; in this case, $C_m = C_i$, and the
estimated values of `Vcmax` and other parameters can be considered to be
"effective values" describing the plant's response to intercellular CO~2~.

Another important point is that plants generally do not appreciate being starved
of CO~2~, so it is not usually possible to start a response curve at low CO~2~
and proceed upwards. A more typical approach is to:

1. Begin at ambient atmospheric CO~2~ levels.
2. Decrease towards a low value.
3. Return to ambient levels and wait for the plant to reacclimate; this waiting
   period is usually accomplished by logging several points at ambient CO~2~
   levels.
4. Increase to higher values.

When taking this approach, it therefore becomes necessary to remove the extra
points measured at ambient CO~2~ levels and to reorder the points according to
their CO~2~ values before plotting or analyzing them.

## The Data

A-Ci curves are commonly measured using a Licor portable photosynthesis system
such as the Li-6400 or Li-6800. These machines record values of $A_n$, $g_{sw}$,
and many other important quantities. They produce two types of output files:
plain-text and Microsoft Excel. It is often more convenient to work with the
Excel files since the entries can be easily modified (for example, to remove an
extraneous row or add a new column). On the other hand, it can be more difficult
to access the file contents using other pieces of software such as R. However,
the `PhotoGEA` package reduces this barrier by including tools for reading Licor
Excel files in R, which will be demonstrated in the following section.

# Loading Packages

As always, the first step is to load the packages we will be using. In addition
to `PhotoGEA`, we will also use the `lattice` package for generating plots.

```{r setup, eval = FALSE}
# Load required packages
library(PhotoGEA)
library(lattice)
```

If the `lattice` package is not installed on your R setup, you can install it by
typing `install.packages('lattice')`.

# Loading Licor Data

The `PhotoGEA` package includes two files representing A-Ci curves measured
using two Li-6800 instruments. The data is stored in Microsoft Excel files, and
includes curves measured from two different crop species (tobacco and soybean)
and several different plots of each. Each curve is a sixteen-point CO~2~
response curve; in other words, the CO~2~ concentration in the air surrounding
the leaf was varied, and $A_n$ (among other variables) was measured at each
CO~2~ setpoint. Although these two files are based on real data, noise was added
to it since it is unpublished, so these files should only be used as examples.

The files will be stored on your computer somewhere in your R package
installation directory, and full paths to these files can be obtained with
`system.file`:

```{r licor_file_names}
# Define a vector of paths to the files we wish to load; in this case, we are
# loading example files included with the PhotoGEA package
file_paths <- c(
  system.file('extdata', 'c4_aci_1.xlsx', package = 'PhotoGEA', mustWork = TRUE),
  system.file('extdata', 'c4_aci_2.xlsx', package = 'PhotoGEA', mustWork = TRUE)
)
```

(**Note:** When loading your own files for analysis, it is not advisable to use
`system.file` as we have done here. Instead, file paths can be directly written,
or files can be chosen using an interactive window. See [Input Files] below for
more information.)

To actually read the data in the files and store them in R objects, we will use
the `read_gasex_file` function from `PhotoGEA`. Since there are multiple files
to read, we will call this function once for each file using `lapply`:

```{r loading_licor_data}
# Load each file, storing the result in a list
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_gasex_file(fpath, 'time')
})
```

The result from this command is an R list of "extended data frames" (abbreviated
as `exdf` objects). The `exdf` class is a special data structure defined by the
`PhotoGEA` package. In many ways, an `exdf` object is equivalent to a data
frame, with the major difference being that an `exdf` object includes the units
of each column. For more information, type `?exdf` in the R terminal to access
the built-in help menu entry, or check out the
[Working With Extended Data Frames](working_with_extended_data_frames.html)
vignette.

Generally, it is more convenient to work with a single `exdf` object rather than
a list of them, so our next step will be to combine the objects in the list.
This action can be accomplished using the `rbind` function, which combines
table-like objects by their rows; in other words, it stacks two or more tables
vertically. This action only makes sense if the tables have the same columns, so
before we combine the `exdf` objects, we should make sure this is the case.

The `PhotoGEA` package includes a function called `identify_common_columns` that
can be used to get the names of all columns that are present in all of the Licor
files. Then, we can extract just those columns, and then combine the `exdf`
objects into a single one.

```{r combining_licor_data}
# Get the names of all columns that are present in all of the Licor files
columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
licor_data <- do.call(rbind, licor_exdf_list)
```

Now we have a single R object called `licor_data` that includes all the data
from several Licor Excel files. For more information about consolidating
information from multiple files, see the _Common Patterns_ section of the
[Working With Extended Data Frames](working_with_extended_data_frames.html#combining-data-from-several-files)
vignette.

# Validating the Data

Before attempting to fit the curves, it is a good idea to do some basic checks
of the data to ensure it is organized properly and that it was measured
properly.

## Basic Checks

First, we should make sure there is a column in the data whose value uniquely
identifies each curve. In this particular data set, several "user constants"
were defined while making the measurements that help to identify each curve:
`instrument`, `species`, and `plot`. However, neither of these columns alone are
sufficient to uniquely identify each curve. We can solve this issue by creating
a new column that combines the values from each of these:

```{r adding_id_column}
# Create a new identifier column formatted like `instrument - species - plot`
licor_data[ , 'curve_identifier'] <-
  paste(licor_data[ , 'instrument'], '-', licor_data[ , 'species'], '-', licor_data[ , 'plot'])
```

The next step is to make sure that this column correctly identifies each
response curve. To do this, we can use the `check_licor_data` function from
`PhotoGEA`. Here we will supply the name of a column that should uniquely
identify each response curve (`curve_identifier`), the expected number of points
in each curve (`16`), the name of a "driving" column that should follow the same
sequence in each curve (`CO2_r_sp`). If the data passes the checks, this
function will have no output and will not produce any messages. (For more
information, see the built-in help menu entry by typing `?check_licor_data`.)

```{r check_response_curve}
# Make sure the data meets basic requirements
check_licor_data(licor_data, 'curve_identifier', 16, 'CO2_r_sp')
```

However, if `check_licor_data` detects an issue, it will print a
helpful message to the R terminal. For example, if we had specified the wrong
number of points or the wrong identifier column, we would get error messages:

```{r check_response_curve_error, error = TRUE}
check_licor_data(licor_data, 'curve_identifier', 15)

check_licor_data(licor_data, 'species', 16)

check_licor_data(licor_data, 'curve_identifier', 16, 'Ci')
```

## Plotting the A-Ci Curves

One qualitative way to check the data is to simply create a plot of the A-Ci
curves. In this situation, the `lattice` library makes it simple to include each
curve as its own separate subplot of a figure. For example:

```{r plot_aci}
# Plot all A-Ci curves in the data set
xyplot(
  A ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']'),
  ylab = paste('Net CO2 assimilation rate [', licor_data$units$A, ']')
)
```

Whoops! Why do these curves look so strange? Well, some of the issues are
related to the sequence of CO~2~ values that was used when measuring the curves.
As discussed in [Practicalities], there are several repeated points logged at
the same CO~2~ concentration, and the points are not logged in order of
ascending or descending concentration. In fact, the sequence of CO~2~ setpoints
is as follows:

```{r}
licor_data[licor_data[, 'curve_identifier'] == 'ripe2 - maize - 2', 'CO2_r_sp']
```

Ideally, we would like to remove the ninth and tenth points (where the setpoint
has been reset to 400 to allow the leaf to reacclimate to ambient CO~2~ levels),
and reorder the data so it is arranged from low to high values of `Ci`. This can
be done using the `organize_response_curve` function from `PhotoGEA`:

```{r reorganizing_curves}
# Remove points with duplicated `CO2_r_sp` values and order by `Ci`
licor_data <- organize_response_curve_data(
    licor_data,
    'curve_identifier',
    c(9, 10),
    'Ci'
)
```

Now we can plot them again:

```{r}
<<plot_aci>>
```

They still look a bit strange. Some of this is related to the noise that was
intentionally added to the data. Nevertheless, there are a few points that we
should probably exclude before attempting to fit the curves. One issue is that
the model _never_ predicts a decrease in $A$ when $Ci$ increases. So, it is
usually a good idea to exclude any points at high $Ci$ where $A$ is observed to
decrease. This dropoff in assimilation is due to one or more processes that are
not captured by the model, and the fits will be unreliable. We will remove these
points later, after we have made several other data quality checks.

## Additional Plots for Qualitative Validation

Sometimes a Licor will override the temperature or humidity controls while
making measurements; in this case, conditions inside the measurement chamber may
not be stable, and we may wish to exclude some of these points. We can check for
these types of issues by making more plots. In the following sections, we will
generate several different plots to check each curve for quality.

### Humidity Control

```{r plot_humidity}
# Make a plot to check humidity control
xyplot(
  RHcham + `Humidifier_%` + `Desiccant_%` ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(0, 100),
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']')
)
```

Here, `Humidifier_%` and `Desiccant_%` represent the flow from the humidifier
and desiccant columns, where a value of 0 indicates that the valve to the column
is fully closed and a value of 100 indicates that the valve to the column is
fully opened. `RHcham` represents the relative humidity inside the chamber as a
percentage (in other words, as a value between 0 and 100).

When these curves were measured, a chamber humidity setpoint was specified. So,
when looking at this plot, we should check that the relative humidity is fairly
constant during each curve. Typically, this should be accompanied by relatively
smooth changes in the valve percentages as they accomodate changes in ambient
humidity and leaf photosynthesis. In this plot, all the data looks good.

### Temperature Control

```{r plot_temperature}
# Make a plot to check temperature control
xyplot(
  TleafCnd + Txchg ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  ylim = c(25, 40),
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']'),
  ylab = paste0('Temperature (', licor_data$units$TleafCnd, ')')
)
```

Here, `TleafCnd` is the leaf temperature measured using a thermocouple, and
`Txchg` is the temperature of the heat exhanger that is used to control the air
temperature in the measurement instrument. When these curves were measured, an
exchanger setpoint was specified. So, when looking at this plot, we should check
that `Txchg` is constant during each curve and that the leaf temperature does
not vary in an erratic way. In this plot, all the data looks good.

### CO~2~ Control

```{r plot_co2}
# Make a plot to check CO2 control
xyplot(
  CO2_s + CO2_r + CO2_r_sp ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']'),
  ylab = paste0('CO2 concentration (', licor_data$units$CO2_r, ')')
)
```

Here, `CO2_s` is the CO~2~ concentration in the sample cell, `CO2_r` is the
CO~2~ concentration in the reference cell, and `CO2_r_sp` is the setpoint for
`CO2_r`. When these curves were measured, a sequence of `CO2_r` values was
specified, so, when looking at this plot, we should check that `CO2_r` is
close to `CO2_r_sp`. We also expect that `CO2_s` should be a bit lower than
`CO2_r` because the leaf in the sample chamber is assimilating CO~2~, which
should reduce its concentration in the surrounding air. (An exception to this
rule occurs at very low values of `CO2_r_sp`, since in this case there is not
enough carbon available to assimilate, and the leaf actually releases CO~2~ due
to respiration.) In this plot, all the data looks good.

### Stability

```{r plot_stability}
# Make a plot to check stability criteria
xyplot(
  `A:OK` + `gsw:OK` + Stable ~ Ci | curve_identifier,
  data = licor_data$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', licor_data$units$Ci, ']')
)
```

When measuring response curves with a Licor, it is possible to specify stability
criteria for each point in addition to minimum and maximum wait times. In other
words, once the set point for the driving variable is changed, the machine waits
until the stability criteria are met; there is a minimum waiting period, and
also a maximum to prevent the machine from waiting for too long.

When these curves were measured, stability criteria were supplied for the net
assimilation rate `A` and the stomatal conductance `gsw`. The stability status
for each was stored in the log file because the appropriate logging option for
stability was set. Now, for each point, it is possible to check whether
stability was achieved or whether the point was logged because the maximum
waiting period had been met. If the maximum waiting period is reached and the
plant has still not stabilized, the data point may be unreliable, so it can be
helpful to check this information.

In the plot, `A:OK` indicates whether `A` was stable (0 for no, 1 for yes),
`gsw:OK` indicates whether `gsw` was stable (0 for no, 1 for yes), and
`Stable` indicates the total number of stability conditions that were met. So,
we are looking for points where `Stable` is 2. Otherwise, we can check the other
traces to see whether `A` or `gsw` was unstable.

Comparing these plots with the ones in [Plotting the A-Ci Curves], it seems that
the unstable points correspond with some of the "odd-looking" points in the A-Ci
curves, so it is probably a good idea to remove them before fitting the data.

## Cleaning the Licor Data

While checking over the plots in the previous sections, two issues were noticed:
(1) some points were logged before stability was achieved and (2) some of the
curves exhibit a decrease in $A$ at high values of $C_i$. In this section, we
will demonstrate how to remove the unstable and unusual points.

The following command will keep only the points where `Stable` is exactly 2;
this condition means that all of the stability criteria were satisfied.
Sometimes, following this procedure, a curve will have very few stable points
remaining; it is usually a good idea to automatically exclude any curve with
fewer than three stable points.

```{r remove_unstable_points}
# Only keep points where stability was achieved
licor_data <- licor_data[licor_data[, 'Stable'] == 2, , TRUE]

# Remove any curves that have fewer than three remaining points
npts <- by(licor_data, licor_data[, 'curve_identifier'], nrow)
ids_to_keep <- names(npts[npts > 2])
licor_data <- licor_data[licor_data[, 'curve_identifier'] %in% ids_to_keep, , TRUE]
```

Next, we can use the `remove_points` function from `PhotoGEA` to exclude the
points where there is a decrease in $A$ at high $C_i$. It just so happens that
all of these points were measured by the `ripe1` instrument and occur at the
highest CO~2~ setpoint value, so it is easy to specify them all at once:

```{r remove_points}
# Remove points where `instrument` is `ripe1` and `CO2_r_sp` is 1800
licor_data <- remove_points(
  licor_data,
  list(instrument = 'ripe1', CO2_r_sp = 1800)
)
```

# Fitting Licor Data

Now that we have checked the data quality, we are ready to perform the fitting.
In order to fit the curves, there are several required pieces of information
that are not included in the Licor data files as produced by the instrument:
temperature-dependent values of important photosynthetic parameters such as
$\gamma^*$, values of the total pressure, and values of the partial pressure of
CO~2~ in the mesophyll $P_{cm}$. However, the `PhotoGEA` package includes three
functions to help with these calculations: `calculate_arrhenius`,
`calculate_total_pressure`, and `apply_gm`. Each of these requires an `exdf`
object containing Licor data. The units for each required column will be checked
in an attempt to avoid unit-related errors. More information about these
functions can be obtained from the built-in help system by typing
`?calculate_arrhenius`, `?calculate_total_pressure`, or `?apply_gm`.

First, we can use `calculate_arrhenius`:

```{r calculate_arrhenius}
# Calculate temperature-dependent values of C4 photosynthetic parameters
licor_data <- calculate_arrhenius(licor_data, c4_arrhenius_von_caemmerer)
```

With this simple command, we have used the leaf temperature from the Licor file
to calculate values of several key C4 photosynthetic parameters
(`r paste(names(c4_arrhenius_von_caemmerer), collapse = ', ')`) according to the
Arrhenius activation energies specified in @von_caemmerer_updating_2021. Notice
that this command calculated values of the mesophyll conductance to CO2 (`gmc`)
in units of mol m$^{-2}$ s$^{-1}$ bar$^{-1}$. These values were estmated for
_Setaria viridis_ and may not be appropriate for all C4 plants. They can easily
be overwritten to another value; here we will set
`gmc = 3.0 mol / m^2 / s / bar` for maize and `gmc = Inf` for sorghum; in this
case, the sorghum parameters determined by the fitting process will be
"effective" parameters.

```{r set_gm}
# Override default mesophyll conductance and specify different values for each
# species
licor_data <- set_variable(
  licor_data, 'gmc',
  id_column = 'species',
  value_table = list(maize = 3.0, sorghum = Inf)
)
```

Now we can use `calculate_total_pressure` and `apply_gm` to calculate `PCm`:

```{r apply_gm}
# Calculate the total pressure in the Licor chamber
licor_data <- calculate_total_pressure(licor_data)

# Calculate PCm
licor_data <- apply_gm(
  licor_data,
  'C4' # Indicate C4 photosynthesis
)
```

Together, these functions have added several new columns to `licor_data`,
including `gmc`, 'PCm', 'gamma_star', and others. With this information, we are
now ready to perform the fitting procedure. For this operation, we can use the
`fit_c4_aci` function from the `PhotoGEA` package, which fits a single response
curve to extract the values of key photosynthetic parameters. To apply this
function to each curve in a larger data set and then consolidate the results, we
can use it in conjunction with `by` and `consolidate`, which are also part of
`PhotoGEA`. (For more information about these functions, see the built-in help
menu entries by typing `?fit_c4_aci`, `?by.exdf`, or `?consolidate`, or check
out the _Common Patterns_ section of the
[Working With Extended Data Frames](working_with_extended_data_frames.html#processing-multiple-pieces-of-an-extended-data-frame)
vignette.) Together, these functions will split apart the main data using
the curve identifier column we defined before ([Basic Checks]), fit each A-Ci
curve using the model discussed in [Understanding C4 A-Ci Curves], and
return the resulting parameters and fits:

```{r fit_aci}
# Fit the C4 A-Ci curves
c4_aci_results <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c4_aci,                       # The function to apply to each chunk of `licor_data`
  Ca_atmospheric = 420              # Additional argument passed to `fit_c4_aci`
))
```

## Viewing the Fitted Curves

Having made the fits, it is now a good idea to visually check them, making sure
they look reasonable. As before, we can use `xyplot` from the `lattice` library
to quickly make plots of all the fits, which are stored in
`c4_aci_results$fits`, an `exdf` object:

```{r plot_fits}
# Plot the C4 A-Ci fits
xyplot(
  A + A_fit ~ Ci | curve_identifier,
  data = c4_aci_results$fits$main_data,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', c4_aci_results$fits$units$Ci, ']'),
  ylab = paste('Net CO2 assimilation rate [', c4_aci_results$fits$units$A, ']')
)
```

Note that in addition to the fitted values of `A`, the output also includes
values of the individual limiting rates `Apc` and `Ar`, as well as the estimated
operating point. We can also take a look at these, although they make the plots
a bit busy. We can also choose to plot assimilation against `PCm` rather than
`Ci`, and include the operating point as an open circle:

```{r plot_fit_aci_limiting}
# Plot the C4 A-PCm fits (including limiting rates)
xyplot(
  A + Apc + Ar + A_fit ~ PCm | curve_identifier,
  data = c4_aci_results$fits$main_data,
  type = 'b',
  auto.key = list(space = 'right'),
  grid = TRUE,
  xlab = paste('Mesophyll CO2 pressure [', c4_aci_results$fits$units$PCm, ']'),
  ylab = paste('Net CO2 assimilation rate [', c4_aci_results$fits$units$A, ']'),
  par.settings = list(
    superpose.line = list(col = multi_curve_line_colors()),
    superpose.symbol = list(col = multi_curve_point_colors(), pch = 16)
  ),
  curve_ids = c4_aci_results$fits[, 'curve_identifier'],
  panel = function(...) {
    panel.xyplot(...)
    args <- list(...)
    curve_id <- args$curve_ids[args$subscripts][1]
    fit_param <-
      c4_aci_results$parameters[c4_aci_results$parameters[, 'curve_identifier'] == curve_id, ]
    panel.points(
        fit_param$operating_An_model ~ fit_param$operating_PCm,
        type = 'p',
        col = 'black',
        pch = 1
    )
  }
)
```

Checking the residuals is also a powerful way to gauge the quality of a fit. The
output from `fit_c4_aci` includes the residuals (calculated as `A - A_fit`) in
its output, so it is easy to plot them:

```{r plot_residuals}
# Plot the residuals
xyplot(
  A_residuals ~ Ci | curve_identifier,
  data = c4_aci_results$fits$main_data,
  type = 'b',
  pch = 16,
  grid = TRUE,
  xlab = paste('Intercellular CO2 concentration [', c4_aci_results$fits$units$Ci, ']'),
  ylab = paste('Assimilation rate residual (measured - fitted)\n[', c4_aci_results$fits$units$A, ']')
)
```

A good fit should produce small and randomly-distributed residuals; here there
is no clear pattern to the residuals, so this model is able to reproduce the
measured values fairly well.

# Examining the Results

## Visualizing Average Fit Parameters

Having performed the fits, we can now take a look at the fitted parameter
values, which are stored in `c4_aci_results$parameters`, another `exdf` object.
One way to do this is by using the `barchart_with_errorbars` function from
`PhotoGEA` to create barcharts of the average values for each species, such as
the following example showing the fitted values of `Vpmax`:

```{r vpmax_barchart}
# Make a barchart showing average Vpmax values
barchart_with_errorbars(
  c4_aci_results$parameters[, 'Vpmax_at_25'],
  c4_aci_results$parameters[, 'species'],
  ylim = c(0, 150),
  xlab = 'Species',
  ylab = paste('Vpmax at 25 degrees C [', c4_aci_results$parameters$units$Vpmax_at_25, ']')
)
```

Another option is to create box-whisper plots using the `bwplot` function from
the `lattice` package. Here we use this function to show values of `Vcmax`:

```{r vcmax_bwplot}
# Make a boxplot showing the distribution of Vcmax values
bwplot(
  Vcmax_at_25 ~ species,
  data = c4_aci_results$parameters$main_data,
  ylim = c(0, 60),
  xlab = 'Species',
  ylab = paste('Vcmax at 25 degrees C [', c4_aci_results$parameters$units$Vcmax_at_25, ']')
)
```

## Accessing Raw and Average Values of Fit Parameters

We can also take a look at the raw numbers of the fitted parameters and their
average values. The `c4_aci_results$parameters` object contains many columns but
we only care about a few of them. We can view them as follows:

```{r c4_aci_parameters}
# View the C4 parameters
columns_for_viewing <-
  c('instrument', 'species', 'plot', 'Rd_at_25', 'Vcmax_at_25', 'Vpmax_at_25')

c4_aci_parameters <-
  c4_aci_results$parameters[ , columns_for_viewing, TRUE]

print(c4_aci_parameters)
```

Alternatively, we can extract average values of the parameters for each species
using the `basic_stats` function from `PhotoGEA`:

```{r calculate_stats}
# Compute the average and standard error of each parameter for each species
c4_aci_averages <- basic_stats(c4_aci_results$parameters, 'species')

# View the averages and errors
columns_to_view <- c(
  'species',
  'Rd_at_25_avg', 'Rd_at_25_stderr',
  'Vcmax_at_25_avg', 'Vcmax_at_25_stderr',
  'Vpmax_at_25_avg', 'Vpmax_at_25_stderr'
)
print(c4_aci_averages[ , columns_to_view, TRUE])
```

Of course, more advanced statistical tests to check for differences between
groups can also be performed within R using other packages such as `onewaytests`
or `DescTools`. Alternatively, the parameter values can be exported to a
comma-separated-value (CSV) file and analyzed in another software environment
like `jmp`.


# Customizing Your Script

Note that most of the commands in this vignette have been written in a general
way so they can be used as the basis for your own analysis script (see
[Commands From This Document]). In order to use them in your own script, some or
all of the following changes may be required. There may also be others not
specifically mentioned here.

## Input Files

The file paths specified in `file_paths` will need to be modified so they point
to your Licor files. One way to do this in your own script is to simply write
out relative or absolute paths to the files you wish to load. For example, you
could replace the previous definition of `file_paths` with this one:

```{r write_file_paths, eval = FALSE}
# Define a vector of paths to the files we wish to load
file_paths <- c(
  'myfile1.xlsx',        # `myfile1.xlsx` must be in the current working directory
  'C:/documents/myfile2' # This is an absolute path to `myfile2`
)
```

You may also want to consider using the `choose_input_licor_files` function from
`PhotoGEA`; this function will create a pop-up browser window where you can
interactively select a set of files. Sometimes this is more convenient than
writing out file paths or names. For example, you could replace the previous
definition of `file_paths` with this one:

```{r choose_licor_files, eval = FALSE}
# Interactively define a vector of paths to the files we wish to load
file_paths <- choose_input_licor_files()
```

Unfortunately, `choose_input_licor_files` is only available in interactive R
sessions running on Microsoft Windows, but there is also a platform-independent
option: `choose_input_files`. See the _Translation_ section of the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html#translation)
vignette for more details.

## Curve Identifier

Depending on which user constants are defined in your Licor Excel files, you may
need to modify the definition of the `curve_identifier` column.

## Data Cleaning

Depending on the qualitative data checks, you may need to change the input
arguments to `remove_points`. It might also not be necessary to remove the
unstable points before performing the fits. Often, it is helpful to not perform
any data cleaning at first, and then remove problematic points if they seem to
cause problems with the fits.

## Averages and Standard Errors

Depending on how your data is organized, you may want to change the column used
to divide the data when calculating averages and standard errors.

## Plots

You may need to change the axis limits in some or all of the plots.
Alternatively, you can remove them, allowing `xyplot` to automatically choose
them for you.

## Saving Results

You may want to use `write.csv` to save some or all of the fitting results as
`csv` files. For example, the following commands will allow you to interactively
choose output filenames for the resulting `csv` files:

```{r saving_csv_files, eval = FALSE}
write.csv(c4_aci_results$fits, file.choose(), row.names = FALSE)
write.csv(c4_aci_results$parameters, file.choose(), row.names = FALSE)
write.csv(c4_aci_averages, file.choose(), row.names = FALSE)
```

# Commands From This Document

The following code chunk includes all the central commands used throughout this
document. They are compiled here to make them easy to copy/paste into a text
file to initialize your own script. Annotation has also been added to clearly
indicate the four steps involved in data analysis, as described in the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html)
vignette.

```{r, eval = FALSE}
###
### PRELIMINARIES:
### Loading packages, defining constants, creating helping functions, etc.
###

<<setup>>

###
### TRANSLATION:
### Creating convenient R objects from raw data files
###

<<licor_file_names>>

## IMPORTANT: When loading your own files, it is not advised to use
## `system.file` as in the above code. Instead, write out the names or use the
## `choose_input_licor_files` function.

<<loading_licor_data>>

<<combining_licor_data>>

###
### VALIDATION:
### Organizing the data, checking its consistency and quality, cleaning it
###

<<adding_id_column>>

<<check_response_curve>>

<<reorganizing_curves>>

<<plot_aci>>

<<plot_humidity>>

<<plot_temperature>>

<<plot_co2>>

<<plot_stability>>

<<remove_unstable_points>>

<<remove_points>>

###
### PROCESSING:
### Extracting new pieces of information from the data
###

<<calculate_arrhenius>>

<<set_gm>>

<<apply_gm>>

<<fit_aci>>

<<plot_fits>>

<<plot_fit_aci_limiting>>

<<plot_residuals>>

###
### SYNTHESIS:
### Using plots and statistics to help draw conclusions from the data
###

<<vpmax_barchart>>

<<vcmax_bwplot>>

<<c4_aci_parameters>>

<<calculate_stats>>
```

# References
