---
title: "Combining PhotoGEA With Other Packages"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Combining PhotoGEA With Other Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

There are many approaches to fitting or otherwise processing pieces of gas
exchange data, and several R packages (such as
[`plantecophys`](https://cran.r-project.org/package=plantecophys),
[`plantecowrap`](https://cran.r-project.org/package=plantecowrap),
and [`photosynthesis`](https://cran.r-project.org/package=photosynthesis))
have been written to implement some of these methods.
These R packages are excellent resources that provide access to well-tested and
peer-reviewed fitting methods. On the other hand, they generally do not provide
tools for a full data analysis pipeline. Instead, these packages assume that the
user has already solved the problems of data translation and validation; in
other words, they assume that the user already has an appropriately validated
data set available as one or R data structures.

To address this issue, it is possible to use functions from these other packages
within the context of `PhotoGEA`. In this case, `PhotoGEA` can be used to load
data, validate it, apply the functions to subsets of the data, and collect the
results, as described in the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html)
vignette.

The key to combining another package with `PhotoGEA` is to create "wrappers" for
the other package's functions. A wrapper is a type of function whose main
purpose is to call another function while making minimal calculations of its
own. Wrappers can be created for many different purposes, as discussed on
[Wikipedia](https://en.wikipedia.org/wiki/Wrapper_function). In this context,
wrappers will be used to reformat the inputs and outputs of a function from
another package so they become compatible with functions from `PhotoGEA`.

As an example, this vignette will demonstrate how to create two wrappers of
varying complexity for the `fitaci` function from the `plantecophys` package.
The general ideas introduced here can also be used to create wrappers for other
functions, although the details of creating a wrapper are heavily dependent on
the details of the function to be wrapped. Note that creating a wrapper for a
fitting function from another package is essentially a special case of
designing a customized processing function, a topic that is also covered in the
[Creating Your Own Processing Tools](creating_your_own_processing_tools.html)
vignette.

# Loading Packages

As always, the first step is to load the packages we will be using. In addition
to `PhotoGEA`, we will also use the `plantecophys` and `lattice` packages.

```{r setup}
# Load required packages
library(PhotoGEA)
library(plantecophys)
library(lattice)
```

If the `plantecophys` or `lattice` packages are not installed on your R setup,
you can install them by typing `install.packages('plantecophys')` and/or
`install.packages('lattice')`.

# Loading and Validating Data

For this example, we will load the same set of C~3~ A-Ci curves that are used in
the [Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) vignette, and we
will perform the same steps for organizing and cleaning the data. See that
vignette for more details about these steps. For brevity, the commands are not
included here, but they can be found at the end of this vignette in
[Commands From This Document].


```{r loading_and_validating_data, echo = FALSE}
# Define a vector of paths to the files we wish to load
file_paths <- c(
  system.file('extdata', 'c3_aci_1.xlsx', package = 'PhotoGEA'),
  system.file('extdata', 'c3_aci_2.xlsx', package = 'PhotoGEA')
)

# Load each file, storing the result in a list
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_gasex_file(fpath, 'time')
})

# Get the names of all columns that are present in all of the Licor files
columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
licor_data <- do.call(rbind, licor_exdf_list)

# Create a new identifier column formatted like `instrument - species - plot`
licor_data[ , 'curve_identifier'] <-
  paste(licor_data[ , 'instrument'], '-', licor_data[ , 'species'], '-', licor_data[ , 'plot'])

# Make sure the data meets basic requirements
check_licor_data(licor_data, 'curve_identifier', 16, 'CO2_r_sp')

# Remove points with duplicated `CO2_r_sp` values and order by `Ci`
licor_data <- organize_response_curve_data(
    licor_data,
    'curve_identifier',
    c(9, 10),
    'Ci'
)

# Only keep points where stability was achieved
licor_data <- licor_data[licor_data[, 'Stable'] == 2, , TRUE]

# Remove any curves that have fewer than three remaining points
npts <- by(licor_data, licor_data[, 'curve_identifier'], nrow)
ids_to_keep <- names(npts[npts > 2])
licor_data <- licor_data[licor_data[, 'curve_identifier'] %in% ids_to_keep, , TRUE]

# Remove points where `instrument` is `ripe1` and `CO2_r_sp` is 1800
licor_data <- remove_points(
  licor_data,
  list(instrument = 'ripe1', CO2_r_sp = 1800)
)
```

# Creating a Minimal Wrapper

The `fitaci` function from the `plantecophys` package is a tool for fitting a
single C~3~ CO~2~ response curve, so it can be thought of as an alternative to
the `fit_c3_aci` function from `PhotoGEA`. In the
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) vignette, the following
command is used to apply `fit_c3_aci` to all the response curves in the data set
and then consolidate the results from each curve into a convenient list of
tables:

```{r, eval = FALSE}
c3_aci_results <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c3_aci,                       # The function to apply to each chunk of `licor_data`
  fixed = c(NA, NA, NA, NA),        # Additional argument passed to `fit_c3_aci`
  cj_crossover_min = 100,           # Wj must be > Wc when Cc < 100 ppm
  cj_crossover_max = 550            # Wj must be < Wc when Cc > 550 ppm
))
```

Ideally, we would like to be able to use a similar command to apply
`plantecophys::fitaci` to all the curves. However, `plantecophys::fitaci` has a
different type of input data table (it uses a data frame while `fit_c3_aci` uses
an `exdf`) and produces a different type of output (it returns an `acifit`
object while `fit_c3_aci` returns a list of `exdf` objects). (There are also a
few other differences that will be discussed later.) In this situation, as
discussed previously, we can use a wrapper function to make
`plantecophys::fitaci` behave more like `PhotoGEA::fit_c3_aci` so it can be used
in the same way.

Here is one way to create such a wrapper:

```{r}
# Make a minimal wrapper for `plantecophys::fitaci`
fit_c3_aci_plantecophys <- function(
  replicate_exdf, # an `exdf` object representing a single A-Ci curve
  ...             # additional named arguments to be passed to `fitaci`
)
{
  # Call `plantecophys::fitaci` by passing the `replicate_exdf$main_data`
  # data frame as the `data` input argument
  fit_res <- plantecophys::fitaci(replicate_exdf$main_data, ...)

  # Get the identifier columns from the original exdf object as a data frame
  replicate_identifiers <- identifier_columns(replicate_exdf$main_data)

  # Get a data frame with the fitted values of assimilation and append the
  # identifier columns
  fits <- fit_res$df
  fits <- cbind(replicate_identifiers, fits)

  # `plantecophys::fitaci` returns absurdly high values of `Ap` when it cannot
  # get a good estimate for `TPU`. These can cause problems when plotting the
  # results, so we replace any `Ap` values above 500 micromol / m^2 / s with
  # NA.
  fits[fits$Ap > 500, 'Ap'] <- NA

  # Create a data frame with the parameter values
  parameters <- data.frame(
    Vcmax          = fit_res$par[1, 1],
    Jmax           = fit_res$par[2, 1],
    Rd             = fit_res$par[3, 1]
  )

  parameters$TPU <- if (fit_res$fitTPU) {
    fit_res$par[4, 1]
  } else {
    NA
  }

  parameters <- cbind(replicate_identifiers, parameters)

  # Return a list of two data frames: `fits` and `parameters`
  return(list(
    fits = fits,
    parameters = parameters
  ))
}
```

Much of the code in `fit_c3_aci_plantecophys` is devoted to converting the
`replicate_exdf` input from an `exdf` object to a `data.frame` and returning a
list of `data.frame` objects that were created from the return value of
`fitaci`. One subtelty in the code is the use of the `identifier_columns`
function.

By default, the output from `fitaci` includes important information like the
modeled assimilation rates and the values of the key C~3~ parameters. However,
it does not include any of the "identifying" information that is essential for
keeping the results organized. (In this case, this would be the values of the
`species` and `plot` columns.) The `identifier_columns` function from `PhotoGEA`
provides a simple way to retrieve this information from the `replicate_exdf`
input. To accomplish this, it determines the columns that only have a single
value and returns the singular values of those columns.

Using `identifier_columns` rather than directly accessing the values of
`species` or `plot` makes the `fit_c3_aci_plantecophys` function more flexible.
In fact, all fitting functions in the `PhotoGEA` package make use of
`identifier_columns` to keep track of these important pieces of "metadata."

Another important point is that we have "regularized" the fitting results by
always including `TPU`, even when it is not being fitted. In general, it is
easier to work with functions that always produce the same output quantities.

Now we can apply this function to all the curves in the data set. Note that we
will need to specify a value for the `varnames` input argument of
`plantecophys::fitaci` because our columns have different names than the default
ones. This can be accomplished with the following command, which makes use of
`by.exdf` and `consolidate` as in the
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) vignette:

```{r}
# Fit each curve
c3_aci_results <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c3_aci_plantecophys,          # The function to apply to each chunk of `licor_data`
  varnames = list(                  # Additional argument to `fit_c3_aci_plantecophys`
    ALEAF = 'A',
    Tleaf = 'TleafCnd',
    Ci = 'Ci',
    PPFD = 'Qin',
    Rd = 'Rd'
  )
))
```

The results can now be examined using commands similar to the ones used in the
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) vignette:

```{r}
# Plot each fit
xyplot(
  Ameas + (Ac - Rd) + (Aj - Rd) + (Ap - Rd) + Amodel ~ Ci_original | curve_identifier,
  data = c3_aci_results$fits,
  type = 'l',
  auto.key = list(space = 'right'),
  grid = TRUE,
  xlab = 'Intercellular CO2 concentration (ppm)',
  ylab = 'Assimilation rate (micromol / m^2 / s)',
  par.settings = list(
    superpose.line = list(col = multi_curve_colors()),
    superpose.symbol = list(col = multi_curve_colors())
  )
)

# View the parameter values
columns_for_viewing <-
  c('instrument', 'species', 'plot', 'Vcmax', 'Jmax', 'Rd', 'TPU')

c3_aci_parameters <-
  c3_aci_results$parameters[ , columns_for_viewing]

print(c3_aci_parameters)
```

There are a few things to notice about these commands and the resulting outputs:

- In the plotting command above, we subtract `Rd` from `Ac`, `Aj`, and
  `Ap` because `plantecophys::fitaci` uses different definitions of these rates
  that are equivalent to `Ac + Rd`, `Aj + Rd`, and `Ap + Rd` in our notation.
- In the results, `TPU` and `Ap` are both `NA` because the `fitTPU` argument of
  `plantecophys::fitaci` was left at its default value of `FALSE`. When we
  attempt to include `Ap` in the plot, `lattice::xyplot` simply does not plot
  anything because the values are `NA`.

# Creating a Fancier Wrapper

The wrapper above is quite useful because it allows us to easily apply `fitaci`
to many response curves in a full data set and automatically combine the results
using the `by` and `consolidate` functions. However, it could be improved by
making a few changes that:

- Provide better default values for the column names.
- Check the units of important columns.
- Include more outputs that are calculated by `fitaci`.
- Return a list of `exdf` objects rather than a list of data frames.
- Return only net assimilation rates.

Here is an example of a more complex wrapper that is a bit more user-friendly
than the previous one. In this wrapper, we explicitly break the code into
several sections devoted to checking inputs, calling `fitaci` with appropriate
units, and collecting outputs:

```{r better_wrapper}
# Make a better wrapper for `plantecophys::fitaci`
fit_c3_aci_plantecophys <- function(
  replicate_exdf, # an `exdf` object representing a single A-Ci curve
  a_column_name = 'A',
  tleaf_column_name = 'TleafCnd',
  ci_column_name = 'Ci',
  qin_column_name = 'Qin',
  rd_column_name = 'Rd',
  useRd = FALSE,
  ... # additional named arguments to be passed to `fitaci`
)
{
  ### Check inputs

  if (!is.exdf(replicate_exdf)) {
      stop('fit_c3_aci_plantecophys requires an exdf object')
  }

  # Make sure the required variables are defined and have the correct units
  required_variables <- list()
  required_variables[[a_column_name]] <- 'micromol m^(-2) s^(-1)'
  required_variables[[tleaf_column_name]] <- "degrees C"
  required_variables[[ci_column_name]] <- 'micromol mol^(-1)'
  required_variables[[qin_column_name]] <- 'micromol m^(-2) s^(-1)'

  if (useRd) {
    # Only check the existence and units of the `Rd` column if it will be used
    required_variables[[rd_column_name]] <- 'micromol m^(-2) s^(-1)'
  }

  check_required_variables(replicate_exdf, required_variables)

  # Make sure the user didn't supply their own `varnames` because we will
  # automatically define it from the other column name inputs
  if ('varnames' %in% names(list(...))) {
    stop('do not supply `varnames` when calling fit_c3_aci_plantecophys')
  }

  ### Call function from external packge with appropriate units

  fit_res <- plantecophys::fitaci(
    replicate_exdf$main_data,
    varnames = list(
        ALEAF = a_column_name,
        Tleaf = tleaf_column_name,
        Ci = ci_column_name,
        PPFD = qin_column_name,
        Rd = rd_column_name
    ),
    useRd = useRd,
    ...
  )

  ### Collect outputs

  # Get the identifier columns from the original exdf object
  replicate_identifiers <- identifier_columns(replicate_exdf)

  # Get a data frame with the fitted values of assimilation and convert it to an
  # `exdf` object, setting the category to `fit_c3_aci_plantecophys` and
  # specifying the units for each column
  fits <- exdf(fit_res$df)

  fits <- document_variables(
    fits,
    c('fit_c3_aci_plantecophys', 'Ci',          'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'Ameas',       'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Amodel',      'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Ac',          'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Aj',          'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Ap',          'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Rd',          'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'VPD',         'kPa'),
    c('fit_c3_aci_plantecophys', 'Tleaf',       'degrees C'),
    c('fit_c3_aci_plantecophys', 'Cc',          'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'PPFD',        'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Patm',        'kPa'),
    c('fit_c3_aci_plantecophys', 'Ci_original', 'micromol mol^(-1)')
  )

  # Append the identifier columns to the fits
  fits <- cbind(replicate_identifiers, fits)

  # `plantecophys::fitaci` returns absurdly high values of `Ap` when it cannot
  # get a good estimate for `TPU`. These can cause problems when plotting the
  # results, so we replace any `Ap` values above 500 micromol / m^2 / s with
  # NA.
  fits[fits[, 'Ap'] > 500, 'Ap'] <- NA

  # Convert `plantecophys::fitaci` outputs to net CO2 assimilation rates for
  # consistency with `fit_c3_aci`.
  fits[, 'Ac'] <- fits[, 'Ac'] - fits[, 'Rd']
  fits[, 'Aj'] <- fits[, 'Aj'] - fits[, 'Rd']
  fits[, 'Ap'] <- fits[, 'Ap'] - fits[, 'Rd']

  # Add a column for the residuals
  fits <- set_variable(
    fits,
    'A_residuals',
    'micromol m^(-2) s^(-1)',
    'fit_c3_aci_plantecophys',
    fits[, 'Ameas'] - fits[, 'Amodel']
  )

  # Create an exdf object with the parameter values that are included in the
  # fitting result. Note that we do not include RMSE because it is not
  # calculated correctly.
  parameters <- exdf(data.frame(
    Ci_transition  = as.numeric(fit_res$Ci_transition),
    Ci_transition2 = as.numeric(fit_res$Ci_transition2),
    Tcorrect       = fit_res$Tcorrect,
    Rd_measured    = fit_res$Rd_measured,
    GammaStar      = fit_res$GammaStar,
    Km             = fit_res$Km,
    kminput        = fit_res$kminput,
    gstarinput     = fit_res$gstarinput,
    fitmethod      = fit_res$fitmethod,
    citransition   = fit_res$citransition,
    gmeso          = fit_res$gmeso,
    fitTPU         = fit_res$fitTPU,
    alphag         = fit_res$alphag,
    Vcmax          = fit_res$par[1, 1],
    Vcmax_err      = fit_res$par[1, 2],
    Jmax           = fit_res$par[2, 1],
    Jmax_err       = fit_res$par[2, 2],
    Rd             = fit_res$par[3, 1],
    Rd_err         = fit_res$par[3, 2],
    ci_star        = fit_res$Ci(0),
    A_transition   = fit_res$Photosyn(Ci=fit_res$Ci_transition)$ALEAF
  ))

  # The value of `TPU` and its error will depend on `fitTPU`
  parameters[, 'TPU'] <- if (fit_res$fitTPU) {
    fit_res$par[4, 1]
  } else {
    NA
  }

  parameters[, 'TPU_err'] <- if (fit_res$fitTPU) {
    fit_res$par[4, 2]
  } else {
    NA
  }

  # Document the parameter units
  parameters <- document_variables(
    parameters,
    c('fit_c3_aci_plantecophys', 'Ci_transition',  'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'Ci_transition2', 'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'Tcorrect',       ''),
    c('fit_c3_aci_plantecophys', 'Rd_measured',    ''),
    c('fit_c3_aci_plantecophys', 'GammaStar',      'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'Km',             'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'kminput',        ''),
    c('fit_c3_aci_plantecophys', 'gstarinput',     ''),
    c('fit_c3_aci_plantecophys', 'fitmethod',      ''),
    c('fit_c3_aci_plantecophys', 'citransition',   'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'gmeso',          'mol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'fitTPU',         ''),
    c('fit_c3_aci_plantecophys', 'alphag',         'dimensionless'),
    c('fit_c3_aci_plantecophys', 'Vcmax',          'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Vcmax_err',      'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Jmax',           'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Jmax_err',       'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Rd',             'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'Rd_err',         'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'TPU',            'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'TPU_err',        'micromol m^(-2) s^(-1)'),
    c('fit_c3_aci_plantecophys', 'ci_star',        'micromol mol^(-1)'),
    c('fit_c3_aci_plantecophys', 'A_transition',   'micromol m^(-2) s^(-1)')
  )

  # Append the identifier columns to the parameters
  parameters <- cbind(replicate_identifiers, parameters)

  # Return a list of two data frames: `fits` and `parameters`
  return(list(
      fits = fits,
      parameters = parameters
  ))
}
```

The code for the wrapper has now gotten significantly longer, but there are
several new benefits to the user, such as the following:

- The units of the input columns are checked.
- The wrapper returns `exdf` objects that include units.
- The assimilation rates have also all been standardized to net rates.
- The fit residuals are included with the output.

If you are writing a wrapper that will be used often, it may be worth taking the
time to add nicer features like these ones.

With this wrapper, we can fit all the curves in a dataset and examine the
results using the following code, which is simpler than the previous version:

```{r make_and_examine_fit}
# Fit each curve
c3_aci_results <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_c3_aci_plantecophys           # The function to apply to each chunk of `licor_data`
))

# Plot each fit
xyplot(
  Ameas + Ac + Aj + Ap + Amodel ~ Ci_original | curve_identifier,
  data = c3_aci_results$fits$main_data,
  type = 'l',
  auto.key = list(space = 'right'),
  grid = TRUE,
  xlab = 'Intercellular CO2 concentration (ppm)',
  ylab = 'Assimilation rate (micromol / m^2 / s)',
  par.settings = list(
    superpose.line = list(col = multi_curve_colors()),
    superpose.symbol = list(col = multi_curve_colors())
  )
)

# View the parameter values
columns_for_viewing <-
  c('instrument', 'species', 'plot', 'Vcmax', 'Jmax', 'Rd', 'TPU')

c3_aci_parameters <-
  c3_aci_results$parameters[ , columns_for_viewing, TRUE]

print(c3_aci_parameters)
```

It's also easy to plot the residuals now:

```{r plot_residuals}
# Plot the residuals
xyplot(
  A_residuals ~ Ci_original | curve_identifier,
  data = c3_aci_results$fits$main_data,
  type = 'b',
  pch = 16,
  grid = TRUE,
  xlab = paste0('Intercellular CO2 concentration (',
    c3_aci_results$fits$units$Ci_original, ')'),
  ylab = paste0('Assimilation rate residual (measured - fitted)\n(',
    c3_aci_results$fits$units$A_residuals, ')')
)
```

# Commands From This Document

The following code chunk includes all the central commands used throughout this
document. They are compiled here to make them easy to copy/paste into a text
file to initialize your own script.

```{r, eval = FALSE}
<<setup>>

<<loading_and_validating_data>>

<<better_wrapper>>

<<make_and_examine_fit>>

<<plot_residuals>>
```
