---
title: "Creating Your Own Processing Tools"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Creating Your Own Processing Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

The `PhotoGEA` package contains many functions for processing gas exchange data,
such as `apply_gm` and `fit_c3_aci`. (A complete list is available in the
[Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html)
vignette.) However, you may wish to perform some kind of processing that is not
already available from `PhotoGEA`. In this case, it is possible to create your
own processing tools that will be compatible with the other functions in
`PhotoGEA` that help with loading data, validating data, processing sets of
multiple reponse curves, and analyzing the results.

In this vignette, we will provide an example showing some of the best practices
for creating your own processing tools. If you create a tool that may be useful
to others and you would like to share it, contact the `PhotoGEA` package
maintainer about adding your function to the package.

# Loading Packages

As always, the first step is to load the packages we will be using. In addition
to `PhotoGEA`, we will also use the `lattice` package.

```{r setup}
# Load required packages
library(PhotoGEA)
library(lattice)
```

If the `lattice` package is not installed on your R setup, you can install it by
typing `install.packages('lattice')`.

# Loading and Validating Data

For this example, we will load the same set of C~3~ A-Ci curves that are used in
the [Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) vignette, and we
will perform the same steps for organizing and cleaning the data. See that
vignette for more details about these steps. For brevity, the commands are not
included here, but they can be found at the end of this vignette in
[Commands From This Document].


```{r loading_and_validating_data, echo = FALSE}
# Define a vector of paths to the files we wish to load
file_paths <- c(
  system.file('extdata', 'c3_aci_1.xlsx', package = 'PhotoGEA', mustWork = TRUE),
  system.file('extdata', 'c3_aci_2.xlsx', package = 'PhotoGEA', mustWork = TRUE)
)

# Load each file, storing the result in a list
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_gasex_file(fpath, 'time')
})

# Get the names of all columns that are present in all of the Licor files
columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Use `rbind` to combine all the data
licor_data <- do.call(rbind, licor_exdf_list)

# Create a new identifier column formatted like `instrument - species - plot`
licor_data[ , 'curve_identifier'] <-
  paste(licor_data[ , 'instrument'], '-', licor_data[ , 'species'], '-', licor_data[ , 'plot'])

# Make sure the data meets basic requirements
check_licor_data(licor_data, 'curve_identifier', 16, 'CO2_r_sp')

# Remove points with duplicated `CO2_r_sp` values and order by `Ci`
licor_data <- organize_response_curve_data(
    licor_data,
    'curve_identifier',
    c(9, 10),
    'Ci'
)

# Only keep points where stability was achieved
licor_data <- licor_data[licor_data[, 'Stable'] == 2, , TRUE]

# Remove any curves that have fewer than three remaining points
npts <- by(licor_data, licor_data[, 'curve_identifier'], nrow)
ids_to_keep <- names(npts[npts > 2])
licor_data <- licor_data[licor_data[, 'curve_identifier'] %in% ids_to_keep, , TRUE]

# Remove points where `instrument` is `ripe1` and `CO2_r_sp` is 1800
licor_data <- remove_points(
  licor_data,
  list(instrument = 'ripe1', CO2_r_sp = 1800)
)
```

# Choosing a Model To Use

For this example, we will develop a function that fits a rectangular hyperbola
to an A-Ci curve. A rectangular hyperbola is an equation of the form
`f(x) = y_max * x / (x + x_half)`. We can understand a great deal about this
type of equation by examining its form:

- When `x` is very large, `x + x_half` can be approximated as `x`, so in this
  case, the function reduces to `y_max * x / x = y_max`. In other words, the
  function reaches a constant value of `y_max` when `x` is large.
- When `x` is `x_half`, the function's value is `y_max * x_half / (x_half +
  x_half) = y_max / 2`. In other words, `x_half` is the value of `x` where the
  function reaches half of its maximum value.
- When `x` is small, `x + x_half` can be approximated by `x_half`. In this case,
  the function reduces to `y_max * x / x_half`. In other words, the function is
  a straight line with slope `y_max / x_half` when `x` is small.
- When `x` is exactly zero, the function is also zero.

With this in mind, we can see that a rectangular hyperbola begins with a linear
portion and flattens out to a constant value. This is generally similar to the
shape of an A-Ci curve, so it might be reasonable to use this equation for
fitting. This type of model would be characterized as an "empirical" model (in
contrast to a mechanistic or process-based model) because there is no underlying
explanation for why this equation should be a good fit. Thus, it could be
considered to be a simpler alternative to the Farquhar-von-Caemmerer-Berry model
used in the `fit_c3_aci` function.

When applying this to an A-Ci curve, we will want to replace the independent
variable `x` with `Ci` and the calculated value with the net assimilation `A`.
One caveat is that `A` is generally negative when `Ci` is zero or very low, but
the rectangular hyperbola never returns negative values. To get a better fit, it
will be helpful to include respiration as a constant value subtracted from the
hyperbola: `A = A_max * Ci / (Ci + Ci_half) - Rd`.

# General Suggestions for PhotoGEA Fitting Functions

When creating a `PhotoGEA` fitting function, it is a good idea to follow these
rules, which will ensure that the function is compatible with `by + consolidate`
and has similar inputs and outputs to other fitting functions:

- The first input argument should be an `exdf` object that represents one "unit"
  of data; in this case, a single A-Ci curve. This argument is often called
  `replicate_exdf` as a reminder of what it represents and what type of object
  it should be.
- The first argument should be checked to make sure it is an `exdf`; this can be
  done using `is.exdf`.
- The name of each important column from the `exdf` object should be passed as
  an input argument with a default value.
- The function should check to make sure each important column is in the `exdf`
  object and that it has the expected units; this check can be accomplished with
  the `check_required_variables` function from `PhotoGEA`.
- The function should return a list of named `exdf` objects as its outputs; most
  fitting functions return `exdf` objects called `fits` and `parameters`.
- The `fits` return object should be a copy of the original data with additional
  columns for the fitted values and the fit residuals.
- The `parameters` return object should have one row, and three different types
  of columns:
  - Some of its columns should include identifying information about the curve
    such as `species` or `event`; this information can be obtained with the
    `identifier_columns` function from `PhotoGEA`.
  - Some of its columns should include statistics that describe the quality of
    the fit, such as the root mean squared error; this information can be
    obtained with the `residual_stats` function from `PhotoGEA`.
  - The other columns should include the best-fit values of the model's
    parameters and any other information that may be important to the user.
- Any `exdf` objects returned by the function should be fully documented with
  units for each relevant column.
- The category for each `exdf` column created by the function should be set to
  its name to provide a record of how the column was calculated.
- Any problems detected while checking the inputs should cause errors.
- If a fit fails, the function should return `NA` results rather than causing an
  error. Otherwise, this will cause problems while fitting many curves at once,
  since the process will be disrupted by any errors that are thrown.
- In general, the structure of the output (e.g. the number of `exdf` objects in
  the list, the names of the `exdf` objects in the list, and the columns in each
  `exdf` object) should always be the same, no matter what the function's inputs
  are and no matter what the fitting results are. Any outputs that are not
  relevant for a particular fit should just be set to `NA`.
- Provide default values for as many input arguments as possible. (No default
  should be provided for `replicate_exdf`, of course.)

In the next section, we will create a fitting function that meets these
criteria.

# Writing A Fitting Function

Here we write a function that fits a rectangular hyperbola to an A-Ci curve and
follows the suggestions outlined above:

```{r define_fitting_function}
# Define a custom fitting function
fit_hyperbola <- function(
  replicate_exdf,
  a_column_name = 'A',
  ci_column_name = 'Ci',
  initial_guess = list(A_max = 40, Ci_half = 100, Rd = 1)
)
{
  ### Check inputs

  if (!is.exdf(replicate_exdf)) {
      stop("fit_hyperbola requires an exdf object")
  }

  # Make sure the required variables are defined and have the correct units
  required_variables <- list()
  required_variables[[a_column_name]] <- "micromol m^(-2) s^(-1)"
  required_variables[[ci_column_name]] <- "micromol mol^(-1)"

  check_required_variables(replicate_exdf, required_variables)

  # Extract the values of several important columns
  A <- replicate_exdf[, a_column_name]
  Ci <- replicate_exdf[, ci_column_name]

  ### Perform processing operations

  # Wrap `stats::nls` in a `tryCatch` block so we can indicate fit failures by
  # setting `aci_fit` to `NULL`.
  aci_fit <- tryCatch(
    {
      stats::nls(A ~ A_max * Ci / (Ci + Ci_half) - Rd, start = initial_guess)
    },
    error = function(cond) {
      print('Having trouble fitting an A-Ci curve:')
      print(identifier_columns(replicate_exdf))
      print('Giving up on the fit :(')
      print(cond)
      return(NULL)
    },
    warning = function(cond) {
      print('Having trouble fitting an A-Ci curve:')
      print(identifier_columns(replicate_exdf))
      print('Giving up on the fit :(')
      print(cond)
      return(NULL)
    }
  )

  ### Collect and document outputs

  # Extract the fit results and add the fits and residuals to the exdf object
    if (is.null(aci_fit)) {
        A_max <- NA
        A_max_err <- NA
        Ci_half <- NA
        Ci_half_err <- NA
        Rd <- NA
        Rd_err <- NA
        replicate_exdf[, paste0(a_column_name, '_fit')] <- NA
        replicate_exdf[, paste0(a_column_name, '_residuals')] <- NA
    } else {
        fit_summary <- summary(aci_fit)
        A_max <- fit_summary$coefficients[1,1]
        A_max_err <- fit_summary$coefficients[1,2]
        Ci_half <- fit_summary$coefficients[2,1]
        Ci_half_err <- fit_summary$coefficients[2,2]
        Rd <- fit_summary$coefficients[3,1]
        Rd_err <- fit_summary$coefficients[3,2]

        replicate_exdf[, paste0(a_column_name, '_fit')] <-
          A_max * Ci / (Ci + Ci_half) - Rd

        replicate_exdf[, paste0(a_column_name, '_residuals')] <-
          fit_summary$residuals
    }

    # Document the columns that were added to the replicate exdf
    replicate_exdf <- document_variables(
        replicate_exdf,
        c('fit_hyperbola', paste0(a_column_name, '_fit'),       'micromol m^(-2) s^(-1)'),
        c('fit_hyperbola', paste0(a_column_name, '_residuals'), 'micromol m^(-2) s^(-1)')
    )

    # Get the replicate identifier columns
    replicate_identifiers <- identifier_columns(replicate_exdf)

    # Attach the residual stats to the identifiers
    replicate_identifiers <- cbind(
        replicate_identifiers,
        residual_stats(
            replicate_exdf[, paste0(a_column_name, '_residuals')],
            replicate_exdf$units[[a_column_name]],
            3
        )
    )

    # Add the values of the fitted parameters
    replicate_identifiers[, 'A_max'] <- A_max
    replicate_identifiers[, 'A_max_err'] <- A_max_err
    replicate_identifiers[, 'Ci_half'] <- Ci_half
    replicate_identifiers[, 'Ci_half_err'] <- Ci_half_err
    replicate_identifiers[, 'Rd'] <- Rd
    replicate_identifiers[, 'Rd_err'] <- Rd_err

    # Document the columns that were added
    replicate_identifiers <- document_variables(
        replicate_identifiers,
        c('fit_hyperbola', 'A_max',       'micromol m^(-2) s^(-1)'),
        c('fit_hyperbola', 'A_max_err',   'micromol m^(-2) s^(-1)'),
        c('fit_hyperbola', 'Ci_half',     'micromol mol^(-1)'),
        c('fit_hyperbola', 'Ci_half_err', 'micromol mol^(-1)'),
        c('fit_hyperbola', 'Rd',          'micromol m^(-2) s^(-1)'),
        c('fit_hyperbola', 'Rd_err',      'micromol m^(-2) s^(-1)')
    )

    return(list(
      fits = replicate_exdf,
      parameters = replicate_identifiers
    ))
}
```

Here we have split the code into three main sections corresponding to the main
steps that should be taken in any processing function:

- _Check inputs_: Here we check the type of `replicate_exdf` and the units of
  any columns that will be accessed during the subsequent processing.
- _Perform processing operations_: Here we actually perform the fit and get the
  results in their default form.
- _Collect and document outputs_: Here we reorganize the fit results into two
  `exdf` objects corresponding to the fits and parameters, and make sure that
  all units are documented. This section ends by returning the results as a list
  of named `exdf` objects.

To make the fit, we have chosen to use `nls`, a function from base R that
performs nonlinear least-squares fitting. This function is fairly
straightforward to use, and it is quite popular. It requires an initial guess
for the values of the model's parameters, so an `initial_guess` input argument
was included in `fit_hyperbola` allowing the user to specify the starting guess.
One issue with `nls` is that it will throw an error if the fit is not
successful. Dealing with these possible errors necessitates some extra code.
First, we wrap the call to `nls` in `tryCatch`, and then later we have to decide
what to return when there is a fit failure; here we just return `NA` for all the
variables normally determined from the fitting procedure.

One improvement that could be made is to provide a way to generate a better
initial guess for the starting parameter values, but we have left it out for
brevity.

# Using the Fitting Function

Now we can use the new fitting just as we would use any other processing
function from `PhotoGEA`. Here we will use it to fit all the response curves in
the example data set, and then examine the results by plotting the fits,
plotting the residuals, and viewing the parameter values. These commands are
nearly identical to ones from the
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) vignette.

```{r make_and_examine_fit}
# Fit each curve
c3_aci_results <- consolidate(by(
  licor_data,                       # The `exdf` object containing the curves
  licor_data[, 'curve_identifier'], # A factor used to split `licor_data` into chunks
  fit_hyperbola                     # The function to apply to each chunk of `licor_data`
))

# Plot each fit
xyplot(
  A + A_fit ~ Ci | curve_identifier,
  data = c3_aci_results$fits$main_data,
  type = 'l',
  auto.key = list(space = 'right'),
  grid = TRUE,
  xlab = 'Intercellular CO2 concentration (ppm)',
  ylab = 'Assimilation rate (micromol / m^2 / s)',
  par.settings = list(
    superpose.line = list(col = multi_curve_colors()),
    superpose.symbol = list(col = multi_curve_colors())
  )
)

# Plot the residuals
xyplot(
  A_residuals ~ Ci | curve_identifier,
  data = c3_aci_results$fits$main_data,
  type = 'b',
  pch = 16,
  grid = TRUE,
  xlab = paste0('Intercellular CO2 concentration (',
    c3_aci_results$fits$units$Ci, ')'),
  ylab = paste0('Assimilation rate residual (measured - fitted)\n(',
    c3_aci_results$fits$units$A_residuals, ')')
)

# View the parameter values
columns_for_viewing <-
  c('instrument', 'species', 'plot', 'A_max', 'Ci_half', 'Rd', 'RMSE')

c3_aci_parameters <-
  c3_aci_results$parameters[ , columns_for_viewing, TRUE]

print(c3_aci_parameters)
```

# More Examples

Another example can be found in the
[Combining PhotoGEA With Other Packages](combining_with_other_packages.html)
vignette, which discusses how to write wrappers for functions from other
packages. Essentially, this is a specialized case of the ideas discussed here in
this vignette.

Additional examples can be found by accessing the source code for the built-in
processing functions provided with `PhotoGEA`. One way to see the code is to
simply type the name of a function in the R terminal; for example,
`fit_ball_berry`. Although this method is convenient, the downside is that any
comments in the original code will not be included. An alternate way is to view
the code on GitHub, where all the comments will be retained. For example, the
source code for `fit_ball_berry` can be found by accessing the
[PhotoGEA GitHub page](https://github.com/eloch216/PhotoGEA/) and navigating to
[R/fit_ball_berry.R](https://github.com/eloch216/PhotoGEA/blob/main/R/fit_ball_berry.R).

# Commands From This Document

The following code chunk includes all the central commands used throughout this
document. They are compiled here to make them easy to copy/paste into a text
file to initialize your own script.

```{r, eval = FALSE}
<<setup>>

<<loading_and_validating_data>>

<<define_fitting_function>>

<<make_and_examine_fit>>
```
