---
title: "Developing a Data Analysis Pipeline"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Developing a Data Analysis Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  fig.align = "center"
)
```

```{r setup, echo = FALSE}
# Load required packages
library(PhotoGEA)
library(lattice)
```

# Overview

The main purpose of the `PhotoGEA` package is to provide tools for creating a
"data analysis pipeline" for photosynthetic gas exchange data. Although the
base version of R coupled with popular packages like `lattice` and `ggplot2`
provides an excellent set of general tools for data analysis, it is not
specialized for gas exchange data, and using it for this purpose can sometimes
be tedious because it usually requires writing various customized functions.
`PhotoGEA` is designed to address these gaps by providing several specialized
functions related to photosynthetic gas exchange data. By using these functions,
you can spend more time extracting important information from your data and less
time writing code.

In this vignette, we will describe the most important functions in `PhotoGEA`
and how they relate to a general data workflow or "pipeline." The description
given here will be general, but more detailed examples can be found in the other
vignettes such as [Analyzing Ball-Berry Data](analyzing_ball_berry_data.html).
We will also discuss some general strategies for writing scripts that are
customized for your own data and preferences.

# Key Steps in Analyzing Data

It is convenient to break up the process of "data analysis" into four key steps:

- _Translation_: Converting data from its original state into a more convenient
   format that can be understood by a piece of analysis software.
- _Validation_: Ensuring the data meets basic requirements for quality and
  consistency.
- _Processing_: Performing operations to extract new information from the data
  such as calculating the values of new quantities or fitting curves.
- _Synthesis_: Drawing conclusions from the data, commonly done by applying
  statistical operations such as computing averages across groups or determining
  whether observed differences between groups are significant.

A "data analysis pipeline" refers to a relatively simple and repeatable way to
perform each of these steps on a set of data. In the following sections, we will
explain how the `PhotoGEA` package, in conjunction with base R and a few other
popular packages, can be used to accomplish each of these steps.

**Note**: Documentation for any function mentioned in this vignette can be
obtained through R's built-in help system, which can be accessed using the `?`
command. For example, to get more information about the `read_licor_file`
function, type `?read_licor_file` in the R terminal.

## Translation

Since we are working exclusively in R, translation refers to the process of
creating R objects from instrument log files. `PhotoGEA` currently includes two
core functions related to translation:

- `read_licor_file`: Creates an R data object from a Microsoft Excel file
  created by a Licor LI-6800 portable gas exchange measurement system.
- `read_tdl_file`: Creates an R data object from a data file created by a
  tunable diode laser measurement system.

Both of these functions produce "extended data frame" (`exdf`) objects in R. The
`exdf` class is very similar to a regular data frame, with the main difference
being that an `exdf` object keeps track of the units associated with each
column. This data structure provides several key benefits when analyzing gas
exchange data, and more information is available in the
[Working with Extended Data Frames](working_with_extended_data_frames.html)
vignette.

It is often the case that a full data set is spread across multiple files.
Because of this, translating data into R often involves combining multiple
`exdf` objects (one from each file) together into one object that holds all of
the data. To help with this, the `PhotoGEA` package provides several functions
that can be used to select files via dialog windows:

- `choose_input_files`
- `choose_input_licor_files` (only available on Microsoft Windows)
- `choose_input_tdl_files` (only available on Microsoft Windows)

It also provides the `identify_common_columns` function, which can be used along
with `rbind` to easily combine a set of `exdf` objects. This operation is
described in the _Combining Data From Several Files_ section of the
[Working with Extended Data Frames](working_with_extended_data_frames.html#combining-data-from-several-files)
vignette.

The process of translation "sets the stage" for the validation, processing, and
synthesis steps in the sense that it determines the structure of the data. In
other words, because these translation functions produce `exdf` objects, many
of the other functions used in the remaining data analysis steps are also
designed to work with `exdf` objects.

## Validation

There is a phrase that originated in computer science but applies to many human
endeavors, including data analysis: "Garbage in, garbage out." In this context
we can understand it to mean that any processing or synthesis we apply to our
data will produce meaningless results if the data itself does not meet certain
requirements for quality and consistency.

As an example, we will try fitting a straight line to a set of curved data:

```{r}
# Generate some data using a cubic function
X <- seq(from = 0, to = 10, length.out = 21)
Y <- (X - 3) * (X - 5) * (X - 7)

# Fit a linear model to the data
linear_fit <- lm(Y ~ X)

# Plot the fit results
xyplot(
  linear_fit$fitted.values + linear_fit$model$Y ~ linear_fit$model$X,
  type = 'b',
  pch = 16,
  auto = TRUE,
  grid = TRUE
)
```

In this case, the fitting procedure does not report any errors, and it even
returns a fairly high value of R^2^ (`r summary(linear_fit)$r.squared`), but it
is clear that a linear model is not applicable to this data. In other words,
this is a case of "garbage in," and we have produced a set of "garbage out."

The goal of data validation is to ensure that any subsequent processing is
appropriate; in other words, to avoid a "garbage in" scenario. Validation often
consists of three parts:

 1. Ensuring that the data set is properly organized into subsets (if
    applicable).
 2. Identifying any problematic points in the data set.
 3. "Cleaning" the data set by removing those points.

There are many ways to validate data, and methods can broadly be categorized as
automated vs. manual and objective vs. subjective. Here are a few examples:

- Plotting the raw data to locate abnormal points is an example of a manual and
  subjective form of validation, since it requires you to visually take a look.
- Removing points from the raw data where the Li-6800's stability criteria were
  not met is an example of automated and objective validation, since it can
  proceed without any judegement from a human.

Typically, best results are achieved using a mix of multiple forms of
validation. The `PhotoGEA` package includes several functions to help with data
validation:

- `check_licor_data`: If a data set represents multiple "response curves"
  measured by a Licor, this function can be used to make sure that each response
  curve can be properly located and that all curves follow the same sequence of
  "driving" variable values.
- `organize_response_curve_data`: Again, if a data set represents multiple
  "response curves" measured by a Licor, this function can be used to remove
  certain points from each curve and to reorder the data in a convenient way for
  plotting.
- `identify_tdl_cycles`: If a data set represents measurements from a tunable
  diode laser, this function can be used to identify complete cycles within the
  data as the gas handling system periodically measures gas from multiple lines.
- `remove_points`: This function simplifies the process of removing individual
  points or even entire response curves from a data set if they are found to be
  unreliable.
- `exclude_outliers`: This function can be used to remove points where
  measurement conditions are unusual; for example, if a leaf temperature
  value recorded by a Li-6800 is significantly different from the recorded
  temperatures for the other points in a response curve.

In addition to these functions, there are more basic operations for subsetting
and plotting that can be used to help with validation; see the
[Working with Extended Data Frames](working_with_extended_data_frames.html)
vignette for more details about subsetting and plotting data from `exdf`
objects.

## Processing

In the processing step, new information is extracted from the raw data in the
data set. The `PhotoGEA` package includes many functions that are specialized
for processing photosynthetic gas exchange data. Broadly, they fall into two
types:

- Functions that calculate the values of new variables from the values of ones
  already included in the data set.
- Functions that apply a fitting procedure to single response curves within the
  full data set.

In general, the "processing" step will require several of these functions, and
will depend strongly on the type of measurements that were made; for example,
A-Ci curves measured from C~3~ plants will require a different type of
processing than A-Ci curves measured from C~4~ plants.

All of these functions are designed to be applied to `exdf` objects. In this
case, `exdf` objects provide a large advantage over regular data frames because
they include information about units; in fact, all of these functions will check
the units of any required variables before proceeding to make sure the results
are correct.

### Functions for Calculating New Variable Values

The `PhotoGEA` package includes the following functions for calculating the
values of new variables from the values of variables that already exist in the
data set:

- `calculate_arrhenius`: Calculates temperature-dependent values of
  photosynthetic parameters using Arrhenius exponents.
- `calculate_ball_berry_index`: Calculates the Ball-Berry index from values of
  net assimilation, relative humidity at the leaf surface, and CO~2~
  concentration at the leaf surface.
- `calculate_c3_assimilation`: Calculates assimilation rates using the
  Farquhar-von-Caemmerer-Berry model for C3 photosynthesis.
- `calculate_cc`: Calculates values of the chloroplast CO~2~ concentration.
- `calculate_gas_properties`: Calculates gas properties that are not included in
  the default Licor log files, such as the H~2~O concentration at the leaf
  surface and the stomatal conductance to CO~2~.

(This list is not exhaustive; instead, it only includes functions for which
documentation has been written.)

These functions generally treat each row in an `exdf` object as an independent
observation, and so they can be applied to objects that include data from many
plants or response curves.

### Functions for Fitting Response Curves

The `PhotoGEA` package includes the following functions for fitting individual
response curves:

- `fit_ball_berry`: Calculates a linear fit of stomatal conductance to H~2~O vs.
  the Ball-Berry index.
- `fit_c3_aci`: Calculates a nonlinear fit of net CO2 assimilation vs.
  chloroplastic CO~2~ concentration.

(This list is not exhaustive; instead, it only includes functions for which
documentation has been written.)

These functions generally assume that the input `exdf` object represents a
single response curve, and they return both the optimized parameter values and
the fitted values of the measured varaibles; for example, `fit_ball_berry`
performs the fit on a single Ball-Berry curve and returns the Ball-Berry
intercept, the Ball-Berry slope, and the fitted values of stomatal conductance.

These functions cannot be applied directly to an `exdf` object that contains a
full set of multiple response curves, because they assume that their input is a
single response curve. However, they can be used in conjunction with the `by`
and `consolidate` functions to automatically split a large `exdf` into
individual curves, apply the fitting function to each curve, and combine the
results from each fit; see the _Processing Multiple Pieces of an Extended Data
Frame_ section of the
[Working with Extended Data Frames](working_with_extended_data_frames.html#processing-multiple-pieces-of-an-extended-data-frame)
vignette for more details about doing this.

### Functions for Calibrating TDL Data

The `PhotoGEA` package includes the following functions for calibrating TDL
data:

- `process_tdl_cycle_erml`: Designed for the TDL in ERML.
- `process_tdl_cycle_polynomial`: Applies a correction based on a polynomial fit
  to the readings from two or more reference tanks.

As with the fitting functions above, these functions generally assume that the
input `exdf` object represents data from a single TDL cycle, and must be used
along with `by` and `consolidate` when processing a full TDL data set.

## Synthesis

In the synthesis step, statistical operations are used to draw conclusions about
the data, and visualizations are generated to make them easier to understand.
The `PhotoGEA` package includes a few tools for basic operations:

- `exclude_outliers`: This function splits an `exdf` object into subsets
  according to the values of one or more identifier columns, determines any rows
  in each subset where the value of a certain column is an outlier, and removes
  those rows. For example, this function could be used to "clean" the output
  from a Ball-Berry fitting procedure by excluding curves from each species
  where the Ball-Berry slope is an outlier.
- `basic_stats`: This function splits an `exdf` object into subsets according to
  the values of one or more identifier columns, and then calculates the mean and
  standard deviation for each applicable column in the object. For example, this
  function could be used to calculate the average value of the Ball-Berry slope
  and intercept for each species present in a data set.
- `barchart_with_errorbars`: This function can be used to create a barchart
  where the height of each bar is the average value from a set of measurements
  and the error bars are determined from the standard error.
- `xyplot_avg_rc`: This function can be used to plot average response curves
  created by averaging individual curves measured from multiple plants.

More complex statistical operations can be performed with functions from other R
packages such as `onewaytests` or `DescTools`.

# Creating Data Analysis Scripts

An R script is essentially just a collection of R commands stored in a
plain-text document, usually with a `.R` extension. The commands in the script
can be copied and pasted into the R terminal to run the code, or the code can be
executed automatically using the `source` function. Writing scripts can save a
lot of time by making code more reusable, and it can serve as a record of the
exact steps you used while analyzing your data.

The `PhotoGEA` package includes several vignettes that go through detailed
examples of data analysis, and the end of each of these vignettes includes a
collection of all the code used in the vignette. This code is designed to be the
start of a script. You are encouraged to copy it and make any necessary
modifications that might be required to apply it to your own data. This is a
great way to get started with analyzing data in R.

In general, there are two types of scripts that you might consider making:

- General-purpose scripts that can apply to many different data sets.
- Narrow scripts that only apply to one particular data set.

## General-Purpose Scripts

A general-purpose script would be a script that is designed to apply to a
particular type of data (for example, C~3~ CO~2~ response curves) but not a
particular set of data.

- This kind of script would only include minimal data validation because the
  process of "cleaning" the data is often specific to a particular data set.
- This kind of script would avoid hard-coding any particular file names;
  instead, it would make use of the `choose_input_files`,
  `choose_input_licor_files`, and `choose_input_tdl_files` functions so that it
  can be flexibly applied to different groups of data files.

The main reason to use a script like this would be to take a quick first look at
a new set of data. If your script is well-written and you always include the
same "identifying" user constants in your Licor files (e.g. `species`, `event`,
`replicate`, etc), then it should be possible to do a basic investigation of
your data in just a few minutes.

As you begin to analyze different types of data, you may develop a collection of
general purpose scripts. It is recommended to store this in a dedicated
directory that does not include any data files.

## Narrow Scripts

A narrow script would be designed to apply to a particular set of data.

- This kind of script would include data cleaning steps that only apply to
  one set of data.
- This kind of script would use hard-coded file names to clarify which files it
  should be used with, and it would typically be stored in the same directory as
  its corresponding data files.

Often, a narrow script can be formed by copying a general-purpose script and
then modifying it to hard-code file names and include data cleaning steps.

The main reason to use a script like this would be to create a reproducible
record of the exact steps you took to analyze the data. This way, you will never
forget what you did. If you need to make a change to the analysis, it will be
simple to just modify part of this script. Trust me: this will save a lot of
time for your future self.

In addition to saving your time, this kind of script will let you easily share
your analysis with other researchers. In fact, a dedicated analysis script that
specifies the versions of each package it uses can later be included with your
data when you publish, making it simple to fulfill the requirements for
"open science." There is a great deal of research showing that papers with
easily reproducible analysis (facilitated by the availability of data and
scripts) is cited more often. For example, see the following:

- @piwowar_sharing_2007
- @bueno_de_mesquita_symposium_2003
- @dorch_data_2015
- @henneken_linking_2011
- @popkin_data_2019

# References
