---
title: "Frequently Asked Questions"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  fig.align = "center"
)
```

# How Do I Install An Old Version of PhotoGEA?

All PhotoGEA releases are tagged on GitHub, and the tag names are formatted as
`vX.Y.Z`, where `X.Y.Z` is the version number. Because of this,
`remotes::install_github` can be used to install old versions from within R,
supplying the appropriate tag name as the `ref` input argument. For example,
version `1.0.0` can be installed as follows:

```{r, eval = FALSE}
remotes::install_github('eloch216/PhotoGEA', ref = 'v1.0.0')
```

Note that this command requires the `remotes` package, which can be installed
from within R by typing `install.packages('remotes')`.

# How Do I Install The Development Version of PhotoGEA?

The development version of PhotoGEA contains the latest changes, although it may
be a "work in progress" and subject to suddent changes without any warning. It
can always be found on the `unreleased` branch on GitHub, and the package can be
installed from this branch as follows:

```{r, eval = FALSE}
remotes::install_github('eloch216/PhotoGEA', ref = 'unreleased')
```

Note that this command requires the `remotes` package, which can be installed
from within R by typing `install.packages('remotes')`.

# What If My Response Curves Have Different Numbers of Points?

The analysis guides use data sets where each response curve has the same number
of points, making it easy to apply `check_response_curve_data` and
`organize_response_curve_data`. However, data sets may sometimes have curves
measured using different sequences of setpoints.

To deal with this situation, one strategy is to split the full set into groups
that are expected to use the same sequences, then separately check and organize
each group, and finally recombine all the groups back together. The code
snippets below show two examples of how this could be accomplished.

## Splitting response curves according to the number of points in each curve

The following code would replace the calls to `check_response_curve_data` and
`organize_response_curve_data` in the user guides. This code was originally
written for a set of A-Ci curves that used different numbers of recovery points.
The user wished to keep the final recovery point to use for subsequent analysis.

The curves used the following sequences of `CO2_r` setpoint values:

- 16 points: 400, 300, 200, 120, 70, 30, 10, 400, 400, 400, 600, 800, 1200,
  1500, 1800, 400

- 18 points: 400, 300, 200, 120, 70, 30, 10, 400, 400, 400, 400, 400, 600, 800,
  1200, 1500, 1800, 400

- 19 points: 400, 300, 200, 120, 70, 30, 10, 400, 400, 400, 400, 400, 400, 600,
  800, 1200, 1500, 1800, 400

```{r, eval = FALSE}
# Add a new column called `curve_npts` that stores the number of points in each
# response curve
licor_data <- do.call(rbind, by(licor_data, licor_data[, 'curve_identifier'], function(x) {
  x[, 'curve_npts'] <- nrow(x)
  x
}))

# Choose points to remove, depending on how many points are in the curve
pts_to_remove <- list(
  '16' = c(1, 8:9,  16),
  '18' = c(1, 8:11, 18),
  '19' = c(1, 8:12, 19)
)

# Check and process each group of curves depending on how many points are in the
# curve
licor_exdf_list_processed <- by(licor_data, licor_data[, 'curve_npts'], function(x) {
  # Get the number of points in these curves
  npts <- x[1, 'curve_npts']

  # Make sure info is specified for this group of curves
  if (!as.character(npts) %in% names(pts_to_remove)) {
    stop('Points to remove were not specified for npts = `', npts, '`')
  }

  # Make sure the data meets basic requirements
  check_response_curve_data(x, 'curve_identifier', npts, 'CO2_r_sp')

  # Remove points with duplicated `CO2_r_sp` values and order by `Ci`
  organize_response_curve_data(
    x,
    'curve_identifier',
    pts_to_remove[[as.character(npts)]],
    'Ci'
  )
})

# Use `rbind` to recombine all the data
licor_data <- do.call(rbind, licor_exdf_list_processed)
```

## Splitting the response curves according to the date they were measured and the number of points in each curve

This code would replace the calls to `check_response_curve_data` and
`organize_response_curve_data` in the user guides. This code was originally
written for a set of A-Ci curves that used different sequences of `CO2_r`
setpoints on different days.

The curves used the following sequences of `CO2_r` setpoint values:

- 2023-03-21: 18 points: 400, 300, 200, 150, 100, 75, 50, 40, 30, 20, 10, 400,
  400, 600, 800, 1000, 1200, 1500

- 2023-03-23: 19 points: 400, 300, 200, 150, 100, 75, 50, 40, 30, 20, 10, 400,
  400, 500, 600, 800, 1000, 1200, 1500

- 2023-03-24: 19 points: 400, 300, 200, 150, 100, 75, 50, 40, 30, 20, 400, 400,
  450, 500, 600, 800, 1000, 1200, 1500

```{r, eval = FALSE}
# Add a new column called `curve_npts` that stores the number of points in each
# response curve
licor_data <- do.call(rbind, by(licor_data, licor_data[, 'curve_identifier'], function(x) {
  x[, 'curve_npts'] <- nrow(x)
  x
}))

# Add a new column called `date_ymd` that stores the date formatted as
# YYYY-MM-DD
licor_data[, 'date_ymd'] <- paste(
  substring(licor_data[, 'date'], 1, 4),
  substring(licor_data[, 'date'], 5, 6),
  substring(licor_data[, 'date'], 7, 8),
  sep = '-'
)

# Add a new column called `date_ymd_npts` that combines the date and the number
# of points
licor_data[, 'date_ymd_npts'] <-
  paste(licor_data[, 'date_ymd'], licor_data[, 'curve_npts'], sep = ' - ')

# Choose points to remove, depending on the date the curve was measured and the
# number of points it contains
pts_to_remove <- list(
  '2023-03-21 - 18' = c(12, 13),
  '2023-03-23 - 19' = c(12, 13),
  '2023-03-24 - 19' = c(11, 12)
)

# Check and process each group of curves depending on the date and the number of
# points
licor_exdf_list_processed <- by(licor_data, licor_data[, 'date_ymd_npts'], function(x) {
  # Get the date and number of points in these curves
  date_ymd_npts <- x[1, 'date_ymd_npts']
  npts <- x[1, 'curve_npts']

  # Make sure info is specified for this group of curves
  if (!date_ymd_npts %in% names(pts_to_remove)) {
    stop('Points to remove were not specified for date_ymd_npts = `', date_ymd_npts, '`')
  }

  # Make sure the data meets basic requirements
  check_response_curve_data(x, 'curve_identifier', npts, 'CO2_r_sp')

  # Remove points with duplicated `CO2_r_sp` values and order by `Ci`
  organize_response_curve_data(
    x,
    'curve_identifier',
    pts_to_remove[[date_ymd_npts]],
    'Ci'
  )
})

# Use `rbind` to recombine all the data
licor_data <- do.call(rbind, licor_exdf_list_processed)
```
