---
title: "Getting Started With PhotoGEA"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started With PhotoGEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

**PhotoGEA** (short for **photo**synthetic **g**as **e**xchange **a**nalysis) is
an R package that provides a suite of tools for loading, processing, and
analyzing photosynthetic gas exchange data. See
[Lochocki, Salesse-Smith, & McGrath (2025)](https://doi.org/10.1111/pce.15501)
for more information.

This vignette is designed to be viewed online; some of the links may not work
properly in a local version. To see the online version, visit
https://eloch216.github.io/PhotoGEA/articles/PhotoGEA.html.

# Installing PhotoGEA

`PhotoGEA` is available on CRAN, so the easiest way to install the latest
release is to type the following from within an R terminal:

```{r, eval = FALSE}
install.packages('PhotoGEA')
```

# An Example: C~3~ CO~2~ Response Curves

As an example, we will read data from two Licor Li-6800 log files that
contain several A-Ci curves measured from tobacco and soybean plants, fit a
model to each response curve, and then plot some of the results. This is a
basic example that just scratches the surface of what is possible with
`PhotoGEA`.

(**Note:** When loading your own files for analysis, it is not advisable to use
`PhotoGEA_example_file_path` as we have done in the code below. Instead, file
paths can be directly written, or files can be chosen using an interactive
window. See the
[Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html#input-files) vignette
for more information.)

## Fitting the Curves

The following code can be used to read the data and fit each curve:

```{r example, eval = FALSE}
# Load required packages
library(PhotoGEA)
library(lattice)

# Define a vector of paths to the files we wish to load; in this case, we are
# loading example files included with the PhotoGEA package
file_paths <- c(
  PhotoGEA_example_file_path('c3_aci_1.xlsx'),
  PhotoGEA_example_file_path('c3_aci_2.xlsx')
)

# Load the data from each file
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_gasex_file(fpath, 'time')
})

# Get the names of all columns that are present in all of the Licor files
columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Combine the data from all the files
licor_data <- do.call(rbind, licor_exdf_list)

# Define a new column that uniquely identifies each curve
licor_data[, 'curve_id'] <-
  paste(licor_data[, 'species'], '-', licor_data[, 'plot'] )

# Organize the data
licor_data <- organize_response_curve_data(
    licor_data,
    'curve_id',
    c(9, 10, 16),
    'CO2_r_sp'
)

# Calculate the total pressure
licor_data <- calculate_total_pressure(licor_data)

# Calculate temperature-dependent values of C3 photosynthetic parameters
licor_data <- calculate_temperature_response(licor_data, c3_temperature_param_bernacchi)

# The default optimizer uses randomness, so we will set a seed to ensure the
# results from this fit are always identical
set.seed(1234)

# Fit all curves in the data set
aci_results <- consolidate(by(
  licor_data,
  licor_data[, 'curve_id'],
  fit_c3_aci,
  Ca_atmospheric = 420
))
```

```{r, echo = FALSE}
timing <- system.time({
<<example>>
})
```

When this document was generated, evaluating this code required the following
amount of time:

```{r, echo = FALSE}
timing
```

The timing results may vary depending on the particular machine used to run the
code. Nevertheless, this is a small time investment for an advanced algorithm
that uses derivative-free optimizers for robust fitting and calculates
nonparametric confidence intervals to determine which estimated parameter values
are reliable.

This example contains 13 commands, so it certainly isn't short; however, a close
look reveals that much of the commands are general and would apply to _any_ set
of C~3~ response curves. In fact, only a few parts would need to be modified,
such as the list of files to read, the curve identifier, and the value of
mesophyll conductance. While using `PhotoGEA`, you are encouraged to copy this
example and any others to use as the base of your own scripts; work smarter, not
harder!

## Viewing the Results

Having fit the response curves, it is also possible to view the fits, the
parameter estimates, and their confidence intervals. `PhotoGEA` provides several
tools for doing this, which enable users to check the fit quality and ensure
that only reliable parameter estimates are used in subsequent analysis.

We can plot the measured values of net assimilation (black circles), the fitted
values of net assimilation (`An`), and each of the limiting assimilation rates
calculated during the fitting procedure: the Rubisco limited rate (`Ac`), the
RuBP regeneration limited rate (`Aj`), and the triose phosphate utilization
(TPU) limited rate (`Ap`). This is a basic quality check where we can make sure
that the fits make sense and look believable:

```{r}
plot_c3_aci_fit(aci_results, 'curve_id', 'Ci', ylim = c(-10, 80))
```

In this figure, some curves are missing one or more of the potential limiting
rates. When this occurs, it means that no points in the curve were found to be
limited by that process.

Another way to check the overall quality of the fits is to plot the residuals,
which should be randomly distributed:

```{r}
xyplot(
  A_residuals ~ Ci | curve_id,
  data = aci_results$fits$main_data,
  type = 'b',
  pch = 16,
  grid = TRUE,
  xlab = paste0('Intercellular CO2 concentration (', aci_results$fits$units$Ci, ')'),
  ylab = paste0('Assimilation rate residual (measured - fitted)\n(', aci_results$fits$units$A, ')'),
)
```

For individual parameters, we can take a look at the best-fit values and the
associated confidence intervals. Here is an example showing values of `Tp_at_25`, the
maximum rate of triose phosphate utilization.

```{r}
aci_results$parameters[, c('curve_id', 'Tp_at_25_lower', 'Tp_at_25', 'Tp_at_25_upper')]
```

Some of these estimates have an upper limit of `Inf` and a best estimate of
`NA`. A comparison with the fits shown above indicates that for these curves,
insufficiently many points were found to be TPU-limited, preventing a reliable
estimate of `Tp_at_25`.

It is also possible to plot the best-fit values of a parameter averaged across
subsets of the data in a bar chart, where the error bars represent the standard
error of the mean. Any values of `NA` will be excluded. Here we plot values of
`Vcmax` at 25 degrees C for each species.

```{r}
barchart_with_errorbars(
  aci_results$parameters[, 'Vcmax_at_25'],
  aci_results$parameters[, 'species'],
  xlab = 'Species',
  ylab = paste0('Vcmax at 25 degrees C (', aci_results$parameters$units$Vcmax_at_25, ')'),
  ylim = c(0, 200)
)
```

# Learning More

The `PhotoGEA` package includes extensive documentation, and more articles are
being added all the time:

- [Publication Gallery](publications.html): Includes a list of publications that
  describe or use `PhotoGEA`, along with links to their corresponding analysis
  scripts.
- Analysis guides:
  - [Analyzing C3 A-Ci Curves](web_only/analyzing_c3_aci_curves.html) (This
    example is more detailed than the analysis demonstrated here.)
  - [Analyzing C4 A-Ci Curves](web_only/analyzing_c4_aci_curves.html)
  - [Analyzing Ball-Berry Data](web_only/analyzing_ball_berry_data.html)
  - [Analyzing TDL Data](web_only/analyzing_tdl_data.html)
  - [Analyzing Mesophyll Conductance Data](web_only/analyzing_gm_data.html)
- General `PhotoGEA` advice:
  - [Creating Your Own Processing Tools](web_only/creating_your_own_processing_tools.html):
    Discusses how to create functions compatible with `PhotoGEA` that apply new
    types of processing.
  - [Combining PhotoGEA With Other Packages](web_only/combining_with_other_packages.html):
    Discusses how to create wrappers for functions from other packages to extend
    the processing capabilities of `PhotoGEA`.
  - [Developing a Data Analysis Pipeline](web_only/developing_a_data_analysis_pipeline.html):
    Discusses how `PhotoGEA` provides functionality for all parts of a data
    analysis pipeline, including loading and validating the data -- and how it
    can help save your time and improve the reproducibility of your data
    analysis!
  - [Working With Extended Data Frames](web_only/working_with_extended_data_frames.html):
    Discusses how to work with extended data frames, which are a critical part
    of `PhotoGEA`.
  - [Frequently Asked Questions](web_only/frequent_questions.html): Provides
    answers to frequently asked questions about `PhotoGEA`.
- General gas exchange advice:
  - [Guide to Licor LI-6800 User Constants](web_only/LI6800_user_constants.html):
    Discusses how to include metadata in log files when using a Licor LI-6800
    gas exchange measurement system, and provides advice for dealing with data
    sets that do not contain user constants.
