---
title: "Getting Started With PhotoGEA"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started With PhotoGEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

**PhotoGEA** (short for **photo**synthetic **g**as **e**xchange **a**nalysis) is
an R package that provides a suite of tools for loading, processing, and
analyzing photosynthetic gas exchange data.

# Installing PhotoGEA

The easiest way to install `PhotoGEA` is to type the following from within the R
terminal:

```{r, eval = FALSE}
remotes::install_github('eloch216/PhotoGEA')
```

Note that this method requires the `remotes` package, which can be installed
from within R by typing `install.packages('remotes')`.

# An Example: C~3~ CO~2~ Response Curves

As an example, we will read data from two Licor Li-6800 log files that
contain several A-Ci curves measured from tobacco and soybean plants, fit a
model to each response curve, and then plot some of the results. This is a
basic example that just scratches the surface of what is possible with
`PhotoGEA`.

## Fitting the Curves

The following code can be used to read the data and fit each curve:

```{r example, eval = FALSE}
# Load required packages
library(PhotoGEA)
library(lattice)

# Define a vector of paths to the files we wish to load
file_paths <- c(
  system.file('extdata', 'c3_aci_1.xlsx', package = 'PhotoGEA'),
  system.file('extdata', 'c3_aci_2.xlsx', package = 'PhotoGEA')
)

# Load the data from each file
licor_exdf_list <- lapply(file_paths, function(fpath) {
  read_licor_file(fpath, c(3, 5, 7, 9, 11, 13), 14, 15, 16, 17, 'time')
})

# Get the names of all columns that are present in all of the Licor files
columns_to_keep <- do.call(identify_common_columns, licor_exdf_list)

# Extract just these columns
licor_exdf_list <- lapply(licor_exdf_list, function(x) {
  x[ , columns_to_keep, TRUE]
})

# Combine the data from all the files
licor_data <- do.call(rbind, licor_exdf_list)

# Define a new column that uniquely identifies each curve
licor_data[, 'curve_id'] <-
  paste(licor_data[, 'species'], '-', licor_data[, 'plot'] )

# Organize the data
licor_data <- organize_response_curve_data(
    licor_data,
    'curve_id',
    c(9, 10, 16),
    'CO2_r_sp'
)

# Specify separate mesophyll conductance values for each species
licor_data <- set_variable(
  licor_data, 'gmc', 'mol m^(-2) s^(-1) bar^(-1)',
  id_column = 'species',
  value_table = list(soybean = 0.9, tobacco = 1.1)
)

# Calculate Cc
licor_data <- calculate_cc(licor_data)

# Calculate temperature-dependent values of C3 photosynthetic parameters
licor_data <- calculate_arrhenius(licor_data, c3_arrhenius_bernacchi)

# Fit all curves in the data set
aci_results <- consolidate(by(
  licor_data,
  licor_data[, 'curve_id'],
  fit_c3_aci
))
```

```{r, echo = FALSE}
timing <- system.time({
<<example>>
})
```

When this document was generated, evaluating this code required the following
amount of time:

```{r, echo = FALSE}
timing
```

The timing results may vary depending on the particular machine used to run the
code. Nevertheless: _Do you think you could fit six A-Ci curves in less than
`r ceiling(timing[['elapsed']])` seconds using any other method?_

This example contains 11 commands, so it certainly isn't short; however, a close
look reveals that much of the commands are general and would apply to _any_ set
of C~3~ response curves. In fact, only a few parts would need to be modified,
such as the list of files to read, the curve identifier, and the value of
mesophyll conductance. While using `PhotoGEA`, you are encouraged to copy this
example and any others to use as the base of your own scripts; work smarter, not
harder!

## Viewing the Results

Having fit the response curves, it is also possible to view the fits and the
extracted parameters. For example, we can plot the measured values of net
assimilation (`A`), the fitted values of net assimilation (`A_fit`), and each of
the limiting assimilation rates calculated during the fitting procedure: the
rubisco-limited rate (`Ac`) and the electron-transport-limited rate (`Aj`).

```{r}
lattice::xyplot(
  A + Ac + Aj + A_fit ~ Ci | curve_id,
  data = aci_results$fits$main_data,
  type = 'l',
  auto.key = list(space = 'right'),
  grid = TRUE,
  xlab = paste0('Intercellular CO2 concentration (', aci_results$fits$units$Ci, ')'),
  ylab = paste0('Assimilation rate (', aci_results$fits$units$A, ')'),
  par.settings = list(
    superpose.line = list(col = multi_curve_colors()),
    superpose.symbol = list(col = multi_curve_colors())
  )
)
```

Another way to check the quality of the fits is to plot the residuals, which
should be randomly distributed:

```{r}
lattice::xyplot(
  A_residuals ~ Ci | curve_id,
  data = aci_results$fits$main_data,
  type = 'b',
  pch = 16,
  grid = TRUE,
  xlab = paste0('Intercellular CO2 concentration (', aci_results$fits$units$Ci, ')'),
  ylab = paste0('Assimilation rate residual (measured - fitted)\n(', aci_results$fits$units$A, ')'),
)
```

It is also possible to plot one or more of the fitting parameters averaged
across species in a bar chart with error bars; here we plot values of `Vcmax` at
25 degrees C.

```{r}
barchart_with_errorbars(
  aci_results$parameters[, 'Vcmax_at_25'],
  aci_results$parameters[, 'species'],
  xlab = 'Species',
  ylab = paste0('Vcmax at 25 degrees C (', aci_results$parameters$units$Vcmax_at_25, ')'),
  ylim = c(0, 200)
)
```

# Learning More

The `PhotoGEA` package includes extensive documentation, and more articles are
being added all the time:

- [Developing a Data Analysis Pipeline](developing_a_data_analysis_pipeline.html):
  Discusses how `PhotoGEA` provides functionality for all parts of a data
  analysis pipeline, including loading and validating the data -- and how it can
  help save your time and improve the reproducibility of your data analysis!
- [Working With Extended Data Frames](working_with_extended_data_frames.html):
  Discusses how to work with extended data frames, which are a critical part of
  `PhotoGEA`.
- Analysis case studies:
  - [Analyzing C3 A-Ci Curves](analyzing_c3_aci_curves.html) (This example is
    more detailed than the analysis demonstrated here.)
  - [Analyzing Ball-Berry Data](analyzing_ball_berry_data.html)
  - [Analyzing TDL Data](analyzing_tdl_data.html)
- More advanced topics:
  - [Creating Your Own Processing Tools](creating_your_own_processing_tools.html):
    Discusses how to create functions compatible with `PhotoGEA` that apply new
    types of processing.
  - [Combining PhotoGEA With Other Packages](combining_with_other_packages.html):
    Discusses how to create wrappers for functions from other packages to extend
    the processing capabilities of `PhotoGEA`.
