---
title: "Getting Started With PhotoGEA"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started With PhotoGEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center"
)
```

# Overview

**PhotoGEA** (short for **photo**synthetic **g**as **e**xchange **a**nalysis) is
an R package that provides a suite of tools for loading, processing, and
analyzing photosynthetic gas exchange data. See
[Lochocki, Salesse-Smith, & McGrath (2025)](https://doi.org/10.1111/pce.15501)
for more information.

# Installing PhotoGEA

`PhotoGEA` is available on CRAN, so the easiest way to install the latest
release is to type the following from within an R terminal:

```{r, eval = FALSE}
install.packages('PhotoGEA')
```

# Using PhotoGEA

The `PhotoGEA` package includes extensive documentation, and more articles are
being added all the time:

- Analysis guides:
  - [Analyzing C3 A-Ci Curves](https://eloch216.github.io/PhotoGEA/articles/web_only/analyzing_c3_aci_curves.html)
  - [Analyzing C4 A-Ci Curves](https://eloch216.github.io/PhotoGEA/articles/web_only/analyzing_c4_aci_curves.html)
  - [Analyzing Ball-Berry Data](https://eloch216.github.io/PhotoGEA/articles/web_only/analyzing_ball_berry_data.html)
  - [Analyzing TDL Data](https://eloch216.github.io/PhotoGEA/articles/web_only/analyzing_tdl_data.html)
  - [Analyzing Mesophyll Conductance Data](https://eloch216.github.io/PhotoGEA/articles/web_only/analyzing_gm_data.html)
- [Frequently Asked Questions](https://eloch216.github.io/PhotoGEA/articles/web_only/frequent_questions.html):
  Provides answers to frequently asked questions about `PhotoGEA`.
- [Publication Gallery](https://eloch216.github.io/PhotoGEA/articles/web_only/publications.html):
  Includes a list of publications that describe or use `PhotoGEA`, along with
  links to their corresponding analysis scripts.
- General `PhotoGEA` advice:
  - [Creating Your Own Processing Tools](https://eloch216.github.io/PhotoGEA/articles/web_only/creating_your_own_processing_tools.html):
    Discusses how to create functions compatible with `PhotoGEA` that apply new
    types of processing.
  - [Combining PhotoGEA With Other Packages](https://eloch216.github.io/PhotoGEA/articles/web_only/combining_with_other_packages.html):
    Discusses how to create wrappers for functions from other packages to extend
    the processing capabilities of `PhotoGEA`.
  - [Developing a Data Analysis Pipeline](https://eloch216.github.io/PhotoGEA/articles/web_only/developing_a_data_analysis_pipeline.html):
    Discusses how `PhotoGEA` provides functionality for all parts of a data
    analysis pipeline, including loading and validating the data -- and how it
    can help save your time and improve the reproducibility of your data
    analysis!
  - [Working With Extended Data Frames](https://eloch216.github.io/PhotoGEA/articles/web_only/working_with_extended_data_frames.html):
    Discusses how to work with extended data frames, which are a critical part
    of `PhotoGEA`.
- General gas exchange advice:
  - [Guide to Licor LI-6800 User Constants](https://eloch216.github.io/PhotoGEA/articles/web_only/LI6800_user_constants.html):
    Discusses how to include metadata in log files when using a Licor LI-6800
    gas exchange measurement system, and provides advice for dealing with data
    sets that do not contain user constants.
